---
title: "Pilot Study 2 LC-MS Data"
output:
  html_document: default
  pdf_document: default
date: "2026-08-10"
---


## 1. load packages

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(tibble)
library(pheatmap)
```


#2. Import positive-mode MS-DIAL data

```{r}
file_path <- "C:/Users/norba/Downloads/postive mode height.xlsx"
```

#2. read positive 
_dim shows how big the data is
_names show the names of the columns
```{r}
msdial <- read_excel(
  path = file_path,
  sheet = "postive mode height",
  skip = 4
)

dim(msdial)
names(msdial)
```


#3. Identify sample-intensity columns

_These are the columns containing the actual MS-DIAL feature peak heights.

```{r}
sample_columns <- c(
  "blank_1", "blank_2", "blank_3", "blank_4",
  "exudate_1", "exudate_2", "exudate_3",
  "FC_1", "FC_2", "FC_3",
  "fungi_old_1", "fungi_old_2", "fungi_old_3",
  "leaf_1", "leaf_2", "leaf_3",
  "new_in_P08", "new_in_P09", "new_in_P10",
  "new_out_P08", "new_out_P09", "new_out_P10",
  "root_1", "root_2", "root_3",
  "Soil_control_1", "Soil_control_2", "Soil_control_3"
)
```



#Safety check: making sure every expected sample column exists
#can use setdiff() to know what column is missing from one set to the other

#validation step
```{r}
stopifnot(
  all(sample_columns %in% names(msdial))
)
```


#how many elements are contained in this object?
```{r}
length(sample_columns)

```

#can inspect data with other functions including head(), str(), summary()

#features include: alignment ID, average RT(min), average m/z, metabolite name, adduct type, ms/ms assigned

#4. Make sure intensities are numeric

```{r}
msdial <- msdial %>%
  mutate(
    across(
      all_of(sample_columns),
      as.numeric
    )
  )
```


#Check:

```{r}
sapply(
  msdial[, sample_columns],
  class
)
```



#5. Create sample metadata
#note: \d= digit, += one or more, $=end of string

This tells R what each sample actually represents biologically.

```{r}
sample_metadata <- tibble(
  sample = sample_columns
) %>%
  mutate(

    group = case_when(
      str_detect(sample, "^blank") ~
        "Blank",

      str_detect(sample, "^exudate") ~
        "Plant-soil exudate",

      str_detect(sample, "^Soil_control") ~
        "Soil control",

      str_detect(sample, "^FC") ~
        "Fungal control",

      str_detect(sample, "^fungi_old") ~
        "Fungi old",

      str_detect(sample, "^leaf") ~
        "Leaf",

      str_detect(sample, "^root") ~
        "Root",

      str_detect(sample, "^new_in") ~
        "Fungal agar in",

      str_detect(sample, "^new_out") ~
        "Fungal agar out",

      TRUE ~ "Unknown"
    ),

    replicate = case_when(
      str_detect(sample, "P08") ~ "1",
      str_detect(sample, "P09") ~ "2",
      str_detect(sample, "P10") ~ "3",

      TRUE ~
        str_extract(sample, "\\d+$")
    ),

    broad_matrix = case_when(

      group %in% c(
        "Plant-soil exudate",
        "Soil control"
      ) ~ "Soil extract",

      group %in% c(
        "Fungal control",
        "Fungi old",
        "Fungal agar in",
        "Fungal agar out"
      ) ~ "Fungal agar",

      group %in% c(
        "Leaf",
        "Root"
      ) ~ "Plant tissue",

      group == "Blank" ~
        "Blank",

      TRUE ~ "Unknown"
    )
  )

```


#Check it: view table

```{r}
sample_metadata
```

#And: how many replicates for each group

```{r}
sample_metadata %>%
  count(group)
```


#6. Initial sample-level QC
_no filtering yet
#Calculate: number of features with nonzero peak height; sum of detected MS-DIAL feature peak heights. sum of MS-DIAL feature peak heights, not the raw LC-MS total ion chromatogram.

```{r}
sample_qc <- tibble(

  sample = sample_columns,

  summed_feature_intensity = colSums(
    as.matrix(
      msdial[, sample_columns]
    ),
    na.rm = TRUE
  ),

  detected_features = colSums(
    as.matrix(
      msdial[, sample_columns]
    ) > 0,
    na.rm = TRUE
  )

) %>%
  left_join(
    sample_metadata,
    by = "sample"
  )

sample_qc
```




#7. Find failed/empty injections
```{r}
sample_qc %>%
  filter(
    detected_features == 0 |
      summed_feature_intensity == 0
  )
```



#8. Plot summed feature signal
```{r}
signal_plot <- ggplot(
  sample_qc,
  aes(
    x = reorder(
      sample,
      summed_feature_intensity
    ),
    y = summed_feature_intensity,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title =
      "Sum of detected positive-mode LC-MS feature intensities",
    x = NULL,
    y =
      "Sum of MS-DIAL feature peak heights",
    fill = "Sample group"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )

signal_plot
```


#9. Plot number of detected features based on peak heights above 0

```{r}
feature_count_plot <- ggplot(
  sample_qc,
  aes(
    x = reorder(
      sample,
      detected_features
    ),
    y = detected_features,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title =
      "Detected positive-mode LC-MS features per sample",
    x = NULL,
    y =
      "Features with peak height > 0",
    fill =
      "Sample group"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )

feature_count_plot
```


#10. Calculate replicate QC and CV, cv BEING COEFFICIENT OF VARIATION. CV measures how variable a group is relative to its mean. Lower= replicates are consistent relative to their average. High= more variability across replicates. 
```{r}
group_qc_summary <- sample_qc %>%
  filter(
    sample != "blank_4"
  ) %>%
  group_by(group) %>%
  summarise(

    n = n(),

    mean_features =
      mean(detected_features),

    sd_features =
      sd(detected_features),

    feature_CV_percent =
      100 *
      sd(detected_features) /
      mean(detected_features),

    mean_summed_signal =
      mean(summed_feature_intensity),

    sd_summed_signal =
      sd(summed_feature_intensity),

    signal_CV_percent =
      100 *
      sd(summed_feature_intensity) /
      mean(summed_feature_intensity),

    .groups = "drop"
  )

group_qc_summary
```



#11. Build the quantitative feature table
_Now we shift from:
_What does each sample look like overall?
_to:
_What happened to each individual LC-MS feature?

#create a smaller analysis focused table by choosing coloumns: core feature information first, 
```{r}
feature_table <- msdial %>%
  select(

    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `Fill %`,
    `MS/MS assigned`,

    any_of(c(
      "Reference RT",
      "Reference m/z",
      "Formula",
      "Ontology",
      "INCHIKEY",
      "SMILES",
      "Annotation tag (VS1.0)",
      "RT matched",
      "m/z matched",
      "MS/MS matched",
      "Total score",
      "RT similarity",
      "Dot product",
      "Reverse dot product",
      "Fragment presence %",
      "S/N average"
    )),

    all_of(sample_columns)
  )
```


#12. Give every feature a readable unique name
_Example:
_F1540_mz187.1078_rt0.74
_means:
_MS-DIAL Alignment ID = 1540
_m/z ≈ 187.1078
_RT ≈ 0.74 min

```{r}
feature_table <- feature_table %>%
  mutate(

    feature_name = paste0(

      "F",
      `Alignment ID`,

      "_mz",
      round(
        `Average Mz`,
        4
      ),

      "_rt",
      round(
        `Average Rt(min)`,
        2
      )
    )
  )
```




#13. Define usable blanks and biological samples
```{r}
valid_blank_columns <- c(
  "blank_1",
  "blank_2",
  "blank_3"
)

biological_columns <- setdiff(
  sample_columns,
  c(
    valid_blank_columns,
    "blank_4"
  )
)
```


#Check:

```{r}
length(biological_columns)
```



#14. Calculate blank/background information for every feature. Background screening step. Essentially, how strong is this feature in biological samples compared with the blanks? Searching for at least 1 biological condition far above the background. Limited because of skewing. A sample could have a feature signal of 0, 0, 10,000.

#blank filter: is the signal sufficiently above analytical background somewhere?

#different from a prevalence filter: is the signal reproduciply detected in a biological sample?

```{r}
feature_qc <- feature_table %>%
  mutate(

    blank_mean = rowMeans(
      across(
        all_of(valid_blank_columns)
      ),
      na.rm = TRUE
    ),

    maximum_biological_signal = apply(
      as.matrix(
        pick(
          all_of(biological_columns)
        )
      ),
      1, #operate across rows (margin) 1=row, 2-coloumn
      max, #return the largest value (function)
      na.rm = TRUE
    ),

    sample_to_blank_ratio =
      (
        maximum_biological_signal + 1 #prevents division by zeros
      ) /
      (
        blank_mean + 1
      )
  )
```


#15. Examine different blank thresholds
#Before filtering, save a copy:

```{r}
feature_qc_unfiltered <- feature_qc
```




#Then compare 3×, 5× and 10×: Is the ratio between blank and feature signals 3x? 5x? 10x? Does my retained feature set change dramatically if I make the background rule stricter?

```{r}
blank_filter_summary <- tibble(

  threshold = c(
    "3x",
    "5x",
    "10x"
  ),

  retained_features = c(

    sum(
      feature_qc_unfiltered$
        sample_to_blank_ratio >= 3,
      na.rm = TRUE
    ),

    sum(
      feature_qc_unfiltered$
        sample_to_blank_ratio >= 5,
      na.rm = TRUE
    ),

    sum(
      feature_qc_unfiltered$
        sample_to_blank_ratio >= 10,
      na.rm = TRUE
    )
  )
)

blank_filter_summary

```


#For our main exploratory analysis: use the threshold of 5x. Taje unfiltered feature table, keep only features whose maximum biological signal is at least 5x that of their blank mean signal.

```{r}
feature_qc <- feature_qc_unfiltered %>%
  filter(
    sample_to_blank_ratio >= 5
  )
```



#16. Group-aware prevalence filter

```{r}
group_columns <- list(

  exudate = c(
    "exudate_1",
    "exudate_2",
    "exudate_3"
  ),

  fungal_control = c(
    "FC_1",
    "FC_2",
    "FC_3"
  ),

  fungi_old = c(
    "fungi_old_1",
    "fungi_old_2",
    "fungi_old_3"
  ),

  leaf = c(
    "leaf_1",
    "leaf_2",
    "leaf_3"
  ),

  fungal_agar_in = c(
    "new_in_P08",
    "new_in_P09",
    "new_in_P10"
  ),

  fungal_agar_out = c(
    "new_out_P08",
    "new_out_P09",
    "new_out_P10"
  ),

  root = c(
    "root_1",
    "root_2",
    "root_3"
  ),

  soil_control = c(
    "Soil_control_1",
    "Soil_control_2",
    "Soil_control_3"
  )
)

```



#Calculate detection for features above 0 for each group. For each feature and each sample group, was that feature detected in at least 2 samples?

```{r}
presence_by_group <- sapply(

  group_columns,

  function(cols) {

    rowSums(
      as.matrix(
        feature_qc[, cols]
      ) > 0,
      na.rm = TRUE
    ) >= 2
  }
)
```



#Keep features detected in at least 2 of 3 replicates in at least one biological group:

```{r}
feature_qc$present_in_any_group <-
  apply(
    presence_by_group,
    1,
    any
  )

feature_qc <- feature_qc %>%
  filter(
    present_in_any_group
  )
```



#calculate overall detection count just for reporting:

```{r}
feature_qc <- feature_qc %>%
  mutate(

    number_detected = rowSums(

      across(
        all_of(biological_columns),
        ~ .x > 0
      ),

      na.rm = TRUE
    )
  )
```



#Check filtering: original feature table and filtered feature table

```{r}
nrow(feature_table)
nrow(feature_qc)
```



#17. Build the PCA intensity matrix


```{r}
intensity_matrix <- feature_qc %>%
  select(
    feature_name,
    all_of(biological_columns)
  ) %>%
  column_to_rownames(
    "feature_name"
  ) %>%
  as.matrix()

```


#Transpose:rows = samples, columns = LC-MS features.

```{r}
pca_matrix_raw <- t(
  intensity_matrix
)

dim(pca_matrix_raw)
```


#18. Replace zeros

#For PCA only, replace zeros/missing values with half of that feature's smallest positive measurement.
```{r}
replace_zero_half_min <- function(x) {

  positive_values <-
    x[
      x > 0 &
        is.finite(x)
    ]

  if (
    length(positive_values) == 0
  ) {

    return(
      rep(
        NA_real_,
        length(x)
      )
    )
  }

  replacement <-
    min(positive_values) / 2

  x[
    x == 0 |
      is.na(x)
  ] <- replacement

  x
}

```


#Apply:

```{r}
pca_matrix_imputed <- apply(
  pca_matrix_raw,
  2,
  replace_zero_half_min
)
```



#Remove anything that still cannot be analyzed:

```{r}
pca_matrix_imputed <-
  pca_matrix_imputed[
    ,
    colSums(
      is.na(
        pca_matrix_imputed
      )
    ) == 0,
    drop = FALSE
  ]
```


#19. Median normalize: one normalization method for this pipeline. Calculate one median intensity for each sample, then take the median of all sample medians for target scaling, then sweep or apply the operation acorss a matrix. Rescale every sample so its overall median intensity is comparable with the overall dataset median. for samples.

```{r}
sample_medians <- apply(
  pca_matrix_imputed,
  1,
  median,
  na.rm = TRUE
)

global_median <-
  median(
    sample_medians
  )

pca_matrix_normalized <- sweep(
  pca_matrix_imputed,
  1,
  sample_medians,
  "/"
) * global_median
```


#20. Log2 transformation to compress the scale. for features.

```{r}
pca_matrix_log <- log2(
  pca_matrix_normalized + 1
)
```


#21. Remove zero-variance features

```{r}
feature_variance <- apply(
  pca_matrix_log,
  2,
  var,
  na.rm = TRUE
)

keep_variable_features <-
  is.finite(
    feature_variance
  ) &
  feature_variance > 0

pca_matrix_log <-
  pca_matrix_log[
    ,
    keep_variable_features,
    drop = FALSE
  ]
```


#22. Run broad PCA

```{r}
pca_result <- prcomp(
  pca_matrix_log,
  center = TRUE,
  scale. = TRUE
)

summary(pca_result)
```


#23. Variance explained
```{r}
variance_explained <-
  (
    pca_result$sdev^2 /
      sum(
        pca_result$sdev^2
      )
  ) * 100

variance_explained[1:10]
```



#Create table:

```{r}
variance_table <- tibble(

  PC = paste0(
    "PC",
    seq_along(
      variance_explained
    )
  ),

  variance_percent =
    variance_explained,

  cumulative_percent =
    cumsum(
      variance_explained
    )
)

head(
  variance_table,
  10
)
```


#24. PCA sample scores

```{r}
pca_scores <-
  as.data.frame(
    pca_result$x
  ) %>%
  rownames_to_column(
    "sample"
  ) %>%
  left_join(
    sample_metadata,
    by = "sample"
  )
```


  
#25. Broad PC1 versus PC2 plot

```{r}
broad_pca_plot <- ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC2,
    fill = group,
    shape = broad_matrix
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(

    title =
      "PCA of positive-mode LC-MS features",

    subtitle =
      "5× blank-filtered, median-normalized and log2-transformed",

    x = paste0(
      "PC1 (",
      round(
        variance_explained[1],
        1
      ),
      "%)"
    ),

    y = paste0(
      "PC2 (",
      round(
        variance_explained[2],
        1
      ),
      "%)"
    ),

    fill =
      "Sample group",

    shape =
      "Sample matrix"
  ) +
  theme_classic()

broad_pca_plot
```


#26. PC1 vs PC3
```{r}
pc1_pc3_plot <- ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC3,
    fill = group,
    shape = broad_matrix
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(

    title =
      "PCA: PC1 versus PC3",

    x = paste0(
      "PC1 (",
      round(
        variance_explained[1],
        1
      ),
      "%)"
    ),

    y = paste0(
      "PC3 (",
      round(
        variance_explained[3],
        1
      ),
      "%)"
    )
  ) +
  theme_classic()

pc1_pc3_plot
```


#27. Scree plot. post creating new dimensions of the LC-MS features, the scree plot tells us how much of the total variation in my LC-MS dataset it explained by each principal component? Essentially asks, how much variation does each PC explain?

```{r}
scree_plot <- variance_table %>%
  slice_head(
    n = 15
  ) %>%
  ggplot(
    aes(
      x = factor(
        PC,
        levels = PC
      ),
      y = variance_percent,
      group = 1
    )
  ) +
  geom_col() +
  geom_point() +
  geom_line() +
  labs(
    title =
      "Positive-mode PCA scree plot",
    x =
      "Principal component",
    y =
      "Variance explained (%)"
  ) +
  theme_classic()

scree_plot
```


#28. Create reusable focused-PCA function: instead of writing 50 lines three times, you write them once and change only the inputs

#Instead of repeating the same code three times. This code chunk is programming a function that can be used in another code.
```{r}
run_focused_pca <- function(
    groups_to_use,
    title_text
) {

  samples_to_use <-
    sample_metadata %>%
    filter(
      group %in% groups_to_use
    ) %>%
    pull(sample)

  matrix_subset <-
    pca_matrix_log[
      samples_to_use,
      ,
      drop = FALSE
    ]

  variances <- apply(
    matrix_subset,
    2,
    var,
    na.rm = TRUE
  )

  matrix_subset <-
    matrix_subset[
      ,
      is.finite(variances) &
        variances > 0,
      drop = FALSE
    ]

  result <- prcomp(
    matrix_subset,
    center = TRUE,
    scale. = TRUE
  )

  variance <-
    (
      result$sdev^2 /
        sum(
          result$sdev^2
        )
    ) * 100

  scores <-
    as.data.frame(
      result$x
    ) %>%
    rownames_to_column(
      "sample"
    ) %>%
    left_join(
      sample_metadata,
      by = "sample"
    )

  plot <- ggplot(
    scores,
    aes(
      x = PC1,
      y = PC2,
      fill = group
    )
  ) +
    geom_point(
      shape = 21,
      size = 5,
      color = "black"
    ) +
    geom_text_repel(
      aes(label = sample)
    ) +
    labs(
      title = title_text,

      x = paste0(
        "PC1 (",
        round(
          variance[1],
          1
        ),
        "%)"
      ),

      y = paste0(
        "PC2 (",
        round(
          variance[2],
          1
        ),
        "%)"
      )
    ) +
    theme_classic()

  list(
    pca = result,
    scores = scores,
    variance = variance,
    plot = plot
  )
}
```


#29. Exudate versus soil PCA
```{r}
exudate_soil_pca <-
  run_focused_pca(
    groups_to_use = c(
      "Plant-soil exudate",
      "Soil control"
    ),
    title_text =
      "PCA: plant-soil exudate versus soil control"
  )

exudate_soil_pca$plot
```


#30. Fungal agar PCA


```{r}
fungal_pca <-
  run_focused_pca(
    groups_to_use = c(
      "Fungal control",
      "Fungal agar in",
      "Fungal agar out"
    ),
    title_text =
      "PCA of fungal agar samples"
  )

fungal_pca$plot
```


#31. Leaf versus root PCA

```{r}
tissue_pca <-
  run_focused_pca(
    groups_to_use = c(
      "Leaf",
      "Root"
    ),
    title_text =
      "PCA of leaf and root extracts"
  )

tissue_pca$plot
```


#32. Extract PCA loadings. PCA loadings essentially asks, which features are driving those directions?

```{r}
pca_loadings <-
  as.data.frame(
    pca_result$rotation
  ) %>%
  rownames_to_column(
    "feature_name"
  )
```



#Create annotation table:

```{r}
feature_annotations <- feature_qc %>%
  select(
    feature_name,
    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `Fill %`,
    `MS/MS assigned`,
    any_of(c(
      "Formula",
      "Ontology",
      "INCHIKEY",
      "SMILES",
      "Annotation tag (VS1.0)",
      "RT matched",
      "m/z matched",
      "MS/MS matched",
      "Total score",
      "RT similarity",
      "Dot product",
      "Reverse dot product",
      "Fragment presence %",
      "S/N average"
    )),
    blank_mean,
    sample_to_blank_ratio,
    number_detected
  )
```


#Join:


```{r}
pca_loadings <- pca_loadings %>%
  left_join(
    feature_annotations,
    by = "feature_name"
  )
```


#33. Identify strongest PC1, PC2 and PC3 features. Which 30 LC-MS features have the strongest relationship with this PCA axis? Scores are samples, loadings are features. 

#Reusable function: top 30, focused interpretation.visualization

```{r}
get_top_loadings <- function(
    loading_table,
    pc_column,
    number_to_keep = 30
) {

  loading_table %>%
    mutate(
      selected_loading =
        .data[[pc_column]],

      absolute_loading =
        abs(
          .data[[pc_column]]
        )
    ) %>%
    arrange(
      desc(
        absolute_loading
      )
    ) %>%
    slice_head(
      n = number_to_keep
    )
}
```


#Run:

```{r}

top_pc1_loadings <-
  get_top_loadings(
    pca_loadings,
    "PC1"
  )

top_pc2_loadings <-
  get_top_loadings(
    pca_loadings,
    "PC2"
  )

top_pc3_loadings <-
  get_top_loadings(
    pca_loadings,
    "PC3"
  )

```

#Inspect:

```{r}
top_pc1_loadings %>%
  select(
    feature_name,
    PC1,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```



#And:

```{r}
top_pc2_loadings %>%
  select(
    feature_name,
    PC2,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```

#Reusable function: top 100, broad interpretation/visualization

```{r}
get_top_loadings100 <- function(
    loading_table,
    pc_column,
    number_to_keep = 100
) {

  loading_table %>%
    mutate(
      selected_loading =
        .data[[pc_column]],

      absolute_loading =
        abs(
          .data[[pc_column]]
        )
    ) %>%
    arrange(
      desc(
        absolute_loading
      )
    ) %>%
    slice_head(
      n = number_to_keep
    )
}
```


#Run:

```{r}

top_pc1_loadings100 <-
  get_top_loadings100(
    pca_loadings,
    "PC1"
  )

top_pc2_loadings100 <-
  get_top_loadings100(
    pca_loadings,
    "PC2"
  )

top_pc3_loadings100 <-
  get_top_loadings100(
    pca_loadings,
    "PC3"
  )

```

#Inspect:

```{r}
top_pc1_loadings100 %>%
  select(
    feature_name,
    PC1,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```

#And:
```{r}
top_pc2_loadings100 %>%
  select(
    feature_name,
    PC2,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```

#34 quality control

#34A. Check suspicious very-early RT loadings. Many high-loading features occur around RT 0.4–0.8 min. Could indicate potential loss of anlyte retention, co-elution with interfering matrix components, or system faults like phase mismatch, flow instability, and column degradation. 

#Quantify early-RT features among the top 100 PC1/PC2 drivers
#rank by strongest PC1 or PC2 loading, take the top 100, how many have RT <1min?

```{r}

early_rt_summary <- pca_loadings %>%
  mutate(
    important_PC12 =
      pmax(
        abs(PC1),
        abs(PC2)
      )
  ) %>%
  arrange(
    desc(
      important_PC12
    )
  ) %>%
  slice_head(
    n = 100
  ) %>%
  count(
    early_RT =
      `Average Rt(min)` < 1
  )

early_rt_summary
```
# 34B. Save and inspect the top 100 PC1/PC2 drivers 
#which of those major PCA drivers how RT <1MIN?

```{r}
early_rt_top100 <- pca_loadings %>%
  mutate(
    important_PC12 =
      pmax(
        abs(PC1),
        abs(PC2)
      )
  ) %>%
  arrange(
    desc(
      important_PC12
    )
  ) %>%
  slice_head(
    n = 100
  ) %>%
  mutate(
    early_RT =
      `Average Rt(min)` < 1
  )

# Confirm the early-RT counts

early_rt_top100 %>%
  count(
    early_RT
  )


# Inspect only the early-RT features among the top 100 PCA drivers

early_rt_top100 %>%
  filter(
    early_RT
  ) %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    PC1,
    PC2,
    important_PC12,
    sample_to_blank_ratio
  )
```

# 34C. Find the 30 strongest PC1-loading features among
#      ALL features with RT < 1 min
#start with all RT <1MIN features, which 30 most strongly drive PC1? Also examine their blank ratios. 

```{r}
early_rt_pc1_top30 <- pca_loadings %>%
  filter(
    `Average Rt(min)` < 1
  ) %>%
  arrange(
    desc(
      abs(PC1)
    )
  ) %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    PC1,
    PC2,
    sample_to_blank_ratio
  ) %>%
  slice_head(
    n = 30
  )

early_rt_pc1_top30
```



#35. Create combined PCA candidate table. It is a master ranking of PCA-important features across PC1, PC2, PC3. Essential question: which LC-MS features contribute most strongly to any of the first three principal components? For example, a feature with no PC1 loading but a huge PC3 loading could still be important. *Does not weight a loading by how much variance its PC explains. 

```{r}
important_pca_features <-
  pca_loadings %>%
  mutate(

    maximum_absolute_loading =
      pmax(
        abs(PC1),
        abs(PC2),
        abs(PC3),
        na.rm = TRUE
      )

  ) %>%
  arrange(
    desc(
      maximum_absolute_loading
    )
  )

```


#Inspect top 50: the strongest PCA contributors according to this particular ranking criterion. Across PC1-PC3 generally, which features are the major PCA contributors?

```{r}
important_pca_features %>%
  slice_head(
    n = 50
  )
```


#36. PCA feature heatmap. Essential question, for the top 30 PCA-driving LC-MS features, what does their intensity pattern actually look like across all of my samples? 

#issues with annotation heatmap, will revisit. 

```{r}
top_feature_names <-
  important_pca_features %>%
  slice_head(
    n = 30
  ) %>%
  pull(
    feature_name
  )

heatmap_matrix <-
  t(
    pca_matrix_log[
      ,
      top_feature_names,
      drop = FALSE
    ]
  )

```


#Plot: Top 30 features are the most strong contributors to the major multivariate PCA patterns. Essential question: how do the strongest PCA-driving features vary acorss my samples?
```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  clustering_distance_rows =
    "euclidean",
  clustering_distance_cols =
    "euclidean",
  clustering_method =
    "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 8,
  main =
    "Top features driving positive-mode PCA"
)

```
#36 additional

#PCA Feature-reduction/sensitivity analysis 
#alternative to a relatively small subset of strongly contributing features recapitulates much of the multivariate structure to: the chemical differences are distributed across many features rather than being driven by a small handful of ions. We want to know if biological/sample structure is retained or changes. 


## 36. PCA feature-reduction / sensitivity analysis
#
# Goal:
# Start with the PCA-ranked features and progressively reduce the
# number of features used in PCA.
#
# This asks:
# 1. Does the overall sample structure remain similar?
# 2. How much variance do PC1 and PC2 explain at each feature count?
# 3. At what point does the PCA structure noticeably change?
# 4. Which features remain strong contributors after PCA is recalculated?

# NOTE:
# This is exploratory sensitivity analysis.
# The features are initially ranked using the full PCA, so preservation
# of structure is not independent validation.


```{r}

# 36A. Define feature-set sizes
# ------------------------------------------------------------

feature_set_sizes <- c(
  ncol(pca_matrix_log),  # all available PCA features
  2000,
  1000,
  500,
  200,
  100,
  50,
  30,
  20,
  10
)

# Remove sizes larger than the number of available features
# and remove duplicates

feature_set_sizes <-
  feature_set_sizes[
    feature_set_sizes <=
      ncol(pca_matrix_log)
  ] %>%
  unique()

feature_set_sizes
```
#36B. Function to rerun PCA using a selected number of highly ranked PCA features. Essentially, we're creating a reusable function that reruns PCA for each reduced feature set.

```{r}

run_reduced_feature_pca <- function(
    number_features,
    ranked_feature_table,
    matrix_to_use,
    metadata
) {

  # If asking for all features, keep everything available
  if (
    number_features >=
      ncol(matrix_to_use)
  ) {

    selected_features <-
      colnames(matrix_to_use)

  } else {

    selected_features <-
      ranked_feature_table %>%
      filter(
        feature_name %in%
          colnames(matrix_to_use)
      ) %>%
      slice_head(
        n = number_features
      ) %>%
      pull(
        feature_name
      )
  }


  # Subset matrix to selected features

  reduced_matrix <-
    matrix_to_use[
      ,
      selected_features,
      drop = FALSE
    ]


  # Recheck variance because some selected features could
  # become non-variable in a subset

  reduced_variance <-
    apply(
      reduced_matrix,
      2,
      var,
      na.rm = TRUE
    )

  reduced_matrix <-
    reduced_matrix[
      ,
      is.finite(
        reduced_variance
      ) &
        reduced_variance > 0,
      drop = FALSE
    ]


  # Run PCA again

  reduced_pca <-
    prcomp(
      reduced_matrix,
      center = TRUE,
      scale. = TRUE
    )


  # Calculate variance explained

  variance_percent <-
    (
      reduced_pca$sdev^2 /
        sum(
          reduced_pca$sdev^2
        )
    ) * 100


  # Extract sample scores

  reduced_scores <-
    as.data.frame(
      reduced_pca$x
    ) %>%
    rownames_to_column(
      "sample"
    ) %>%
    left_join(
      metadata,
      by = "sample"
    )


  # Extract loadings from the NEW PCA

  reduced_loadings <-
    as.data.frame(
      reduced_pca$rotation
    ) %>%
    rownames_to_column(
      "feature_name"
    ) %>%
    mutate(

      maximum_absolute_loading =
        pmax(
          abs(PC1),
          abs(PC2),
          abs(PC3),
          na.rm = TRUE
        )
    ) %>%
    arrange(
      desc(
        maximum_absolute_loading
      )
    )


  # Create PCA plot

  reduced_plot <-
    ggplot(
      reduced_scores,
      aes(
        x = PC1,
        y = PC2,
        fill = group
      )
    ) +

    geom_point(
      shape = 21,
      size = 4,
      color = "black"
    ) +

    geom_text_repel(
      aes(
        label = sample
      ),
      size = 3
    ) +

    labs(

      title =
        paste0(
          "PCA using top ",
          ncol(reduced_matrix),
          " features"
        ),

      x =
        paste0(
          "PC1 (",
          round(
            variance_percent[1],
            1
          ),
          "%)"
        ),

      y =
        paste0(
          "PC2 (",
          round(
            variance_percent[2],
            1
          ),
          "%)"
        )
    ) +

    theme_classic()


  # Return all useful objects

  list(

    requested_features =
      number_features,

    actual_features =
      ncol(
        reduced_matrix
      ),

    selected_features =
      colnames(
        reduced_matrix
      ),

    pca =
      reduced_pca,

    scores =
      reduced_scores,

    loadings =
      reduced_loadings,

    variance =
      variance_percent,

    plot =
      reduced_plot
  )
}
```


# 36C. Run PCA across progressively smaller feature sets. Essentially, run that function for every feature-set size.

```{r}
reduced_pca_results <-
  lapply(
    feature_set_sizes,

    function(n_features) {

      run_reduced_feature_pca(

        number_features =
          n_features,

        ranked_feature_table =
          important_pca_features,

        matrix_to_use =
          pca_matrix_log,

        metadata =
          sample_metadata
      )
    }
  )

names(
  reduced_pca_results
) <-
  paste0(
    "Top_",
    feature_set_sizes
  )
```



#check
```{r}
names(
  reduced_pca_results
)
```
# 36D. Create variance summary table. summarize how the variance changes2.

```{r}
pca_reduction_summary <-
  bind_rows(

    lapply(
      names(reduced_pca_results),

      function(result_name) {

        x <- reduced_pca_results[[result_name]]

        tibble(

          feature_set =
            result_name,

          requested_features =
            x$requested_features,

          actual_features =
            x$actual_features,

          PC1_percent =
            x$variance[1],

          PC2_percent =
            x$variance[2],

          PC3_percent =
            x$variance[3],

          PC1_PC2_percent =
            x$variance[1] +
            x$variance[2],

          PC1_PC2_PC3_percent =
            x$variance[1] +
            x$variance[2] +
            x$variance[3]
        )
      }
    )
  )

pca_reduction_summary
```
## 36E. Plot PC1 and PC2 variance across feature-set sizes
```{r}
pca_reduction_variance_long <-
  pca_reduction_summary %>%

  select(
    actual_features,
    PC1_percent,
    PC2_percent,
    PC1_PC2_percent
  ) %>%

  pivot_longer(
    cols =
      c(
        PC1_percent,
        PC2_percent,
        PC1_PC2_percent
      ),
    names_to =
      "variance_measure",
    values_to =
      "percent"
  )
```

#plot. essentially answers the question: as i reduce thousands of features toward 10, does the variance become increasingly concentrated in PC1/PC2?
```{r}
ggplot(
  pca_reduction_variance_long,
  aes(
    x = actual_features,
    y = percent,
    group = variance_measure,
    linetype = variance_measure
  )
) +

  geom_point(
    size = 3
  ) +

  geom_line() +

  scale_x_log10() +

  labs(
    title =
      "PCA variance across progressively reduced feature sets",

    subtitle =
      "Features ranked using strongest loading across PC1-PC3",

    x =
      "Number of features used in PCA (log scale)",

    y =
      "Variance explained (%)",

    linetype =
      NULL
  ) +

  theme_classic()
```


# 36F. Display selected reduced-feature PCA plots

```{r}

reduced_pca_results$Top_2000$plot

reduced_pca_results$Top_1000$plot

reduced_pca_results$Top_500$plot

reduced_pca_results$Top_200$plot

reduced_pca_results$Top_100$plot

reduced_pca_results$Top_50$plot

reduced_pca_results$Top_30$plot

reduced_pca_results$Top_20$plot

reduced_pca_results$Top_10$plot
```


```{r}
reduced_pca_results[[1]]$plot
```
# 36G. Compare the strongest loadings after recalculating PCA. Inspect whether the newly calculated loadings remain similar as you reduce the feature set. 


```{r}

top_loadings_by_reduction <-
  bind_rows(

    lapply(
      names(reduced_pca_results),

      function(result_name) {

        x <- reduced_pca_results[[result_name]]

        x$loadings %>%

          slice_head(
            n = 20
          ) %>%

          mutate(
            feature_set =
              result_name
          )
      }
    )
  )

top_loadings_by_reduction
```



# 36H. Identify features repeatedly appearing among top loadings. Count how often each feature appears among the top 20 loadings after PCA is recalculated.

#a feature with times in the top20 at 8, is repeatedly appearing as a strong PCA contributor across many reduced PCAs. A feature with times in top 20=1 may be important only under one particular feature configuration.
```{r}
stable_pca_drivers <-
  top_loadings_by_reduction %>%

  count(
    feature_name,
    sort = TRUE,
    name =
      "times_in_top20"
  )

stable_pca_drivers
```



# 36I. Add feature annotation to stable PCA drivers. Join annotation information back into analysis. 
```{r}

stable_pca_drivers_annotated <-
  stable_pca_drivers %>%

  left_join(

    feature_qc %>%
      select(
        feature_name,
        `Average Rt(min)`,
        `Average Mz`,
        `Metabolite name`,
        `Adduct type`,
        `MS/MS assigned`,
        Formula,
        Ontology,
        sample_to_blank_ratio
      ),

    by =
      "feature_name"
  )

stable_pca_drivers_annotated %>%
  slice_head(
    n = 100
  )
```


# 36J. Plot most stable PCA-driving features
#create a simple stability plot
```{r}
stable_pca_driver_plot_data <-
  stable_pca_drivers_annotated %>%
  slice_head(
    n = 30
  )
```


#plot
```{r}
ggplot(
  stable_pca_driver_plot_data,
  aes(
    x =
      reorder(
        feature_name,
        times_in_top20
      ),
    y =
      times_in_top20
  )
) +

  geom_col() +

  coord_flip() +

  labs(
    title =
      "Features repeatedly driving PCA across feature reductions",

    subtitle =
      "Count of reduced PCA analyses in which each feature appeared among the top 20 loadings",

    x =
      NULL,

    y =
      "Number of PCA reductions"
  ) +

  theme_classic()
```

# 36K. Reusable heatmap function for reduced feature sets. Heat map. 
```{r}

plot_reduced_heatmap <- function(
    number_to_plot,
    title_text
) {

  feature_names <-
    important_pca_features %>%

    filter(
      feature_name %in%
        colnames(
          pca_matrix_log
        )
    ) %>%

    slice_head(
      n = number_to_plot
    ) %>%

    pull(
      feature_name
    )


  heatmap_data <-
    t(
      pca_matrix_log[
        ,
        feature_names,
        drop = FALSE
      ]
    )


  pheatmap(
    heatmap_data,

    scale =
      "row",

    clustering_distance_rows =
      "euclidean",

    clustering_distance_cols =
      "euclidean",

    clustering_method =
      "complete",

    show_rownames =
      number_to_plot <= 100,

    show_colnames =
      TRUE,

    fontsize_row =
      ifelse(
        number_to_plot <= 30,
        7,
        4
      ),

    main =
      title_text
  )
}
```


# 36L. Heatmaps at progressively smaller feature sets
```{r}

plot_reduced_heatmap(
  500,
  "Top 500 PCA-driving features"
)

plot_reduced_heatmap(
  100,
  "Top 100 PCA-driving features"
)

plot_reduced_heatmap(
  30,
  "Top 30 PCA-driving features"
)

plot_reduced_heatmap(
  10,
  "Top 10 PCA-driving features"
)
```


# 36M. Save feature-reduction results
```{r}
write.csv(
  pca_reduction_summary,
  "positive_mode_PCA_feature_reduction_summary.csv",
  row.names = FALSE
)

write.csv(
  top_loadings_by_reduction,
  "positive_mode_PCA_top_loadings_across_reductions.csv",
  row.names = FALSE
)

write.csv(
  stable_pca_drivers_annotated,
  "positive_mode_stable_PCA_drivers.csv",
  row.names = FALSE
)
```


# 36N. Function to calculate PCA sample-distance matrix
#      using enough PCs to explain >= 80% of variance

#essentially asking, how much of the original multivariate relationship among samples survives feature reduction? A Mantel permutation test gives us a p-value for the relationship between distance matrices. 
```{r}

get_pca_distance <- function(
    pca_object,
    variance_percent,
    variance_cutoff = 80
) {

  cumulative_variance <-
    cumsum(
      variance_percent
    )

  number_pcs <-
    which(
      cumulative_variance >=
        variance_cutoff
    )[1]

  scores_to_use <-
    pca_object$x[
      ,
      seq_len(
        number_pcs
      ),
      drop = FALSE
    ]

  dist(
    scores_to_use,
    method = "euclidean"
  )
}
```


#calculate the full-data reference
```{r}
full_result <-
  reduced_pca_results[[1]]

full_distance <-
  get_pca_distance(
    pca_object =
      full_result$pca,

    variance_percent =
      full_result$variance,

    variance_cutoff =
      80
  )
```



#correlate every reduced PCA with the full PCA. R=0.78 means that at 500 features, they are similar to the relationships in the full dataset
```{r}
pca_structure_correlations <-
  bind_rows(

    lapply(
      names(
        reduced_pca_results
      ),

      function(result_name) {

        x <-
          reduced_pca_results[[result_name]]

        reduced_distance <-
          get_pca_distance(
            pca_object =
              x$pca,

            variance_percent =
              x$variance,

            variance_cutoff =
              80
          )

        correlation <-
          cor(
            as.vector(
              full_distance
            ),

            as.vector(
              reduced_distance
            ),

            method =
              "spearman"
          )

        tibble(

          feature_set =
            result_name,

          actual_features =
            x$actual_features,

          distance_correlation =
            correlation
        )
      }
    )
  )

pca_structure_correlations
```


# 36P. Install/load vegan for Mantel permutation tests
```{r}
library(
  vegan
)
```

# 36Q. Mantel tests:
#      Does each reduced PCA preserve the full-data
#      sample-distance structure more than expected by chance?


```{r}
pca_structure_tests <-
  bind_rows(

    lapply(
      names(
        reduced_pca_results
      ),

      function(result_name) {

        x <-
          reduced_pca_results[[result_name]]

        reduced_distance <-
          get_pca_distance(
            pca_object =
              x$pca,

            variance_percent =
              x$variance,

            variance_cutoff =
              80
          )

        mantel_result <-
          vegan::mantel(
            full_distance,
            reduced_distance,

            method =
              "spearman",

            permutations =
              9999
          )

        tibble(

          feature_set =
            result_name,

          actual_features =
            x$actual_features,

          mantel_r =
            unname(
              mantel_result$statistic
            ),

          p_value =
            mantel_result$signif
        )
      }
    )
  )

pca_structure_tests
```

#does the top N PCA-driving feature set preserve the full metabolomic sample structure better than would be expected from randomly choosing N features?

# 36N. Function to calculate sample-distance structure
# from a PCA using enough PCs to explain >= 80% variance
```{r}

get_pca_distance <- function(
    pca_object,
    variance_percent,
    variance_cutoff = 80
) {

  cumulative_variance <-
    cumsum(
      variance_percent
    )

  number_pcs <-
    which(
      cumulative_variance >=
        variance_cutoff
    )[1]

  scores_to_use <-
    pca_object$x[
      ,
      seq_len(number_pcs),
      drop = FALSE
    ]

  dist(
    scores_to_use,
    method = "euclidean"
  )
}
```




# 36O. Create full-data reference distance matrix. Establish the full-feature PCA as the reference.

```{r}

full_result <-
  reduced_pca_results[[1]]

full_distance <-
  get_pca_distance(
    pca_object =
      full_result$pca,

    variance_percent =
      full_result$variance,

    variance_cutoff =
      80
  )

full_distance
```


# 36P. Measure preservation of full PCA sample structure
#      using Spearman correlation of pairwise distances

```{r}
pca_structure_correlations <-
  bind_rows(

    lapply(
      names(reduced_pca_results),

      function(result_name) {

        x <-
          reduced_pca_results[[result_name]]

        reduced_distance <-
          get_pca_distance(
            pca_object =
              x$pca,

            variance_percent =
              x$variance,

            variance_cutoff =
              80
          )

        structure_correlation <-
          cor(
            as.vector(full_distance),
            as.vector(reduced_distance),
            method = "spearman"
          )

        tibble(

          feature_set =
            result_name,

          actual_features =
            x$actual_features,

          structure_correlation =
            structure_correlation
        )
      }
    )
  )

pca_structure_correlations
```

#correlation near 1
→ reduced PCA preserves the full sample relationships very well

correlation around 0.5
→ moderate preservation

correlation near 0
→ reduced feature set no longer resembles the full structure

# 36Q. Plot preservation of PCA structure as features decrease
```{r}

pca_structure_plot <-
  ggplot(
    pca_structure_correlations,
    aes(
      x = actual_features,
      y = structure_correlation
    )
  ) +

  geom_point(
    size = 3
  ) +

  geom_line() +

  scale_x_log10() +

  labs(
    title =
      "Preservation of full PCA structure during feature reduction",

    subtitle =
      "Correlation of pairwise sample distances with the full-feature PCA",

    x =
      "Number of features used in PCA (log scale)",

    y =
      "Spearman correlation with full PCA"
  ) +

  theme_classic()

pca_structure_plot
```


# 36R. Mantel permutation tests. Mantel test asks: are the pairwise sample relationships in this reduced PCA associated with the pairwise sample relationships in the full PCA?
```{r}

library(vegan)

pca_structure_mantel <-
  bind_rows(

    lapply(
      names(reduced_pca_results),

      function(result_name) {

        x <-
          reduced_pca_results[[result_name]]

        reduced_distance <-
          get_pca_distance(
            pca_object =
              x$pca,

            variance_percent =
              x$variance,

            variance_cutoff =
              80
          )

        mantel_result <-
          vegan::mantel(
            full_distance,
            reduced_distance,
            method = "spearman",
            permutations = 9999
          )

        tibble(

          feature_set =
            result_name,

          actual_features =
            x$actual_features,

          mantel_r =
            unname(
              mantel_result$statistic
            ),

          mantel_p_value =
            mantel_result$signif
        )
      }
    )
  )

pca_structure_mantel
```

# 36S. Function to calculate PCA structure from a random
# feature subset
#essential question: if I randomly chose the same number of features, how well would they preserve the full PCA structure? First create a function that runs PCA on a random subset. 
```{r}

get_random_subset_correlation <- function(
    number_features,
    matrix_to_use,
    full_distance,
    variance_cutoff = 80
) {

  random_features <-
    sample(
      colnames(matrix_to_use),
      size = number_features,
      replace = FALSE
    )

  random_matrix <-
    matrix_to_use[
      ,
      random_features,
      drop = FALSE
    ]

  random_variance <-
    apply(
      random_matrix,
      2,
      var,
      na.rm = TRUE
    )

  random_matrix <-
    random_matrix[
      ,
      is.finite(random_variance) &
        random_variance > 0,
      drop = FALSE
    ]

  random_pca <-
    prcomp(
      random_matrix,
      center = TRUE,
      scale. = TRUE
    )

  random_variance_percent <-
    (
      random_pca$sdev^2 /
        sum(random_pca$sdev^2)
    ) * 100

  random_distance <-
    get_pca_distance(
      pca_object =
        random_pca,

      variance_percent =
        random_variance_percent,

      variance_cutoff =
        variance_cutoff
    )

  cor(
    as.vector(full_distance),
    as.vector(random_distance),
    method = "spearman"
  )
}
```


# 36T. Feature sizes for random-subset permutation testing. Define the feature-set sizes for the randomization analysis. 
```{r}

permutation_feature_sizes <-
  c(
    2000,
    1000,
    500,
    200,
    100,
    50,
    30,
    20,
    10
  )

permutation_feature_sizes <-
  permutation_feature_sizes[
    permutation_feature_sizes <
      ncol(pca_matrix_log)
  ]

permutation_feature_sizes
```
```{r}
number_permutations <- 1000
```



# 36U. Run random feature-set permutations
```{r}
set.seed(123)

random_structure_results <-
  bind_rows(

    lapply(
      permutation_feature_sizes,

      function(n_features) {

        random_correlations <-
          replicate(

            number_permutations,

            get_random_subset_correlation(
              number_features =
                n_features,

              matrix_to_use =
                pca_matrix_log,

              full_distance =
                full_distance,

              variance_cutoff =
                80
            )
          )

        tibble(

          actual_features =
            n_features,

          permutation =
            seq_along(
              random_correlations
            ),

          random_structure_correlation =
            random_correlations
        )
      }
    )
  )

random_structure_results
```



# 36V. Observed PCA-selected feature correlations. Retrieve the correlation obtained with the actual PCA-selected features. 

```{r}

observed_structure_results <-
  pca_structure_correlations %>%

  filter(
    actual_features %in%
      permutation_feature_sizes
  ) %>%

  select(
    actual_features,
    observed_correlation =
      structure_correlation
  )

observed_structure_results
```

# 36W. Empirical permutation p-values. For every feature-set size, how often did a random feature set preserve the full PCA structure at least as well as our PCA-selected feature set?
```{r}
pca_reduction_permutation_tests <-
  observed_structure_results %>%

  left_join(

    random_structure_results %>%

      group_by(
        actual_features
      ) %>%

      summarise(

        random_mean =
          mean(
            random_structure_correlation,
            na.rm = TRUE
          ),

        random_median =
          median(
            random_structure_correlation,
            na.rm = TRUE
          ),

        random_sd =
          sd(
            random_structure_correlation,
            na.rm = TRUE
          ),

        random_lower_95 =
          quantile(
            random_structure_correlation,
            0.025,
            na.rm = TRUE
          ),

        random_upper_95 =
          quantile(
            random_structure_correlation,
            0.975,
            na.rm = TRUE
          ),

        .groups =
          "drop"
      ),

    by =
      "actual_features"
  )
```

#calculate the empirical p-value separately
```{r}
pca_reduction_permutation_tests <-
  pca_reduction_permutation_tests %>%

  rowwise() %>%

  mutate(

    empirical_p_value = {

      random_values <-
        random_structure_results %>%

        filter(
          actual_features ==
            .data$actual_features
        ) %>%

        pull(
          random_structure_correlation
        )

      (
        sum(
          random_values >=
            observed_correlation,
          na.rm = TRUE
        ) + 1
      ) /
        (
          length(
            random_values
          ) + 1
        )
    }
  ) %>%

  ungroup()

pca_reduction_permutation_tests
```



# 36X. Plot observed structure preservation against
# random feature-set expectation
```{r}
pca_reduction_random_plot <-
  ggplot(
    pca_reduction_permutation_tests,
    aes(
      x = actual_features
    )
  ) +

  geom_ribbon(
    aes(
      ymin =
        random_lower_95,

      ymax =
        random_upper_95
    ),

    alpha =
      0.25
  ) +

  geom_line(
    aes(
      y =
        random_mean
    ),

    linetype =
      "dashed"
  ) +

  geom_point(
    aes(
      y =
        observed_correlation
    ),

    size =
      3
  ) +

  geom_line(
    aes(
      y =
        observed_correlation
    )
  ) +

  scale_x_log10() +

  labs(

    title =
      "PCA structure retained during feature reduction",

    subtitle =
      "Observed PCA-selected features compared with random feature subsets",

    x =
      "Number of features (log scale)",

    y =
      "Correlation with full PCA sample structure"
  ) +

  theme_classic()

pca_reduction_random_plot
```

#figure interpretation
#solid observed line
how well PCA-selected features preserve the full structure

dashed line = average preservation from random feature sets

ribbon = 95% range from random feature subsets

# 36Y. Empirical p-values across feature reductions. Tests whether PCA-selected features preserve structure. Essential question: do PCA-selected features preserve gloabl structure better than random subjects?
```{r}

pca_reduction_pvalue_plot <-
  ggplot(
    pca_reduction_permutation_tests,
    aes(
      x =
        actual_features,

      y =
        empirical_p_value
    )
  ) +

  geom_point(
    size = 3
  ) +

  geom_line() +

  geom_hline(
    yintercept =
      0.05,

    linetype =
      "dashed"
  ) +

  scale_x_log10() +

  labs(

    title =
      "Permutation evidence across PCA feature reductions",

    subtitle =
      "Tests whether PCA-selected features preserve structure better than random subsets",

    x =
      "Number of features (log scale)",

    y =
      "Empirical p-value"
  ) +

  theme_classic()

pca_reduction_pvalue_plot
```


# 36Z. Save quantitative PCA-reduction results
```{r}

write.csv(
  pca_structure_correlations,
  "positive_mode_PCA_structure_correlations.csv",
  row.names = FALSE
)

write.csv(
  pca_structure_mantel,
  "positive_mode_PCA_structure_mantel_tests.csv",
  row.names = FALSE
)

write.csv(
  random_structure_results,
  "positive_mode_PCA_random_feature_permutations.csv",
  row.names = FALSE
)

write.csv(
  pca_reduction_permutation_tests,
  "positive_mode_PCA_feature_reduction_permutation_tests.csv",
  row.names = FALSE
)
```

# most of 36 is about "As I progressively restrict the metabolome to the strongest PCA-associated features, do I still recover the same relationships among my biological samples?"



#37. Presence/absence analysis across biological groups. Use feature_qc, not the imputed PCA matrix. Presence/absence must be based on actual observed peak heights.

#First define all sample groups:

# ------------------------------------------------------------
# PRESENCE / ABSENCE ANALYSIS
# ------------------------------------------------------------

# Define sample columns for each biological group


```{r}
presence_groups <- list(

  Exudate = c(
    "exudate_1",
    "exudate_2",
    "exudate_3"
  ),

  Soil = c(
    "Soil_control_1",
    "Soil_control_2",
    "Soil_control_3"
  ),

  Leaf = c(
    "leaf_1",
    "leaf_2",
    "leaf_3"
  ),

  Root = c(
    "root_1",
    "root_2",
    "root_3"
  ),

  Fungal_in = c(
    "new_in_P08",
    "new_in_P09",
    "new_in_P10"
  ),

  Fungal_out = c(
    "new_out_P08",
    "new_out_P09",
    "new_out_P10"
  ),

  Fungal_control = c(
    "FC_1",
    "FC_2",
    "FC_3"
  ),

  Fungi_old = c(
    "fungi_old_1",
    "fungi_old_2",
    "fungi_old_3"
  )
)
```


38. Function for any pairwise presence/absence comparison

We'll define a feature as present when it occurs in at least 2 of 3 biological replicates.

```{r}
run_presence_comparison <- function(
    data,
    group1_cols,
    group2_cols,
    group1_name,
    group2_name,
    min_replicates = 2
) {

  result <- data %>%
    mutate(

      group1_detected_n = rowSums(
        across(
          all_of(group1_cols),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

      group2_detected_n = rowSums(
        across(
          all_of(group2_cols),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

      group1_present =
        group1_detected_n >= min_replicates,

      group2_present =
        group2_detected_n >= min_replicates,

      presence_category = case_when(

        group1_present &
          group2_present ~
          "Shared",

        group1_present &
          !group2_present ~
          paste0(
            group1_name,
            " only"
          ),

        !group1_present &
          group2_present ~
          paste0(
            group2_name,
            " only"
          ),

        TRUE ~
          "Neither consistently detected"
      )
    )

  summary <- result %>%
    count(
      presence_category
    ) %>%
    mutate(

      percent =
        100 * n / sum(n),

      comparison =
        paste0(
          group1_name,
          " vs ",
          group2_name
        )
    ) %>%
    select(
      comparison,
      presence_category,
      n,
      percent
    )

  return(
    list(
      feature_table = result,
      summary = summary
    )
  )
}
```


39. Exudate vs soil

This asks:

Which features appear with the plant-containing soil extraction compared with soil alone?

```{r}
presence_exudate_soil <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Exudate,

    group2_cols =
      presence_groups$Soil,

    group1_name =
      "Exudate",

    group2_name =
      "Soil"
  )

presence_exudate_soil$summary
```


40. Leaf vs root

This asks:

Which features are shared between aboveground and belowground plant tissue, and which are tissue-associated?

```{r}
presence_leaf_root <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Leaf,

    group2_cols =
      presence_groups$Root,

    group1_name =
      "Leaf",

    group2_name =
      "Root"
  )

presence_leaf_root$summary
```


41. Exudate vs root

Yes, I definitely want this comparison.

It asks:

Which features detected in roots are also detectable in the plant-soil extract?

```{r}
presence_exudate_root <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Exudate,

    group2_cols =
      presence_groups$Root,

    group1_name =
      "Exudate",

    group2_name =
      "Root"
  )

presence_exudate_root$summary
```



A feature classified as:

Shared

means it occurred in ≥2/3 exudates and ≥2/3 roots.

That is potentially interesting.

But:

shared ≠ secreted.

It only establishes chemical correspondence at the aligned-feature level.

42. Exudate vs leaf

This comparison is also worthwhile:

```{r}
presence_exudate_leaf <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Exudate,

    group2_cols =
      presence_groups$Leaf,

    group1_name =
      "Exudate",

    group2_name =
      "Leaf"
  )

presence_exudate_leaf$summary
```



This can help distinguish whether exudate-associated signals resemble general plant chemistry or are more strongly associated with roots.

43. Fungal agar in vs out

```{r}
presence_fungal_in_out <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Fungal_in,

    group2_cols =
      presence_groups$Fungal_out,

    group1_name =
      "Fungal in",

    group2_name =
      "Fungal out"
  )

presence_fungal_in_out$summary
```


44. Fungal agar in vs agar control

```{r}
presence_fungal_in_control <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Fungal_in,

    group2_cols =
      presence_groups$Fungal_control,

    group1_name =
      "Fungal in",

    group2_name =
      "Fungal control"
  )

presence_fungal_in_control$summary
```


45. Fungal agar out vs agar control

I would add this too.

It tells us whether the region outside visible fungal growth has nevertheless been chemically altered relative to the control.

```{r}
presence_fungal_out_control <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Fungal_out,

    group2_cols =
      presence_groups$Fungal_control,

    group1_name =
      "Fungal out",

    group2_name =
      "Fungal control"
  )

presence_fungal_out_control$summary
```



That could actually be biologically interesting if diffusible metabolites extend beyond the visible colony.

46. Combine all pairwise presence results

```{r}
presence_summary_all <- bind_rows(

  presence_exudate_soil$summary,

  presence_leaf_root$summary,

  presence_exudate_root$summary,

  presence_exudate_leaf$summary,

  presence_fungal_in_out$summary,

  presence_fungal_in_control$summary,

  presence_fungal_out_control$summary
)

presence_summary_all

```


Sort it:

```{r}
presence_summary_all <- presence_summary_all %>%
  arrange(
    comparison,
    desc(n)
  )

presence_summary_all
```



Save:

```{r}

write.csv(
  presence_summary_all,
  "positive_mode_presence_absence_summary.csv",
  row.names = FALSE
)
```

47. Now do the more interesting three-way plant analysis

Pairwise comparisons are useful, but what you're proposing is actually better expressed as:

Which features occur across leaf → root → exudate?

Let's calculate presence in all three.

```{r}
plant_compartment_presence <- feature_qc %>%
  mutate(

    leaf_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Leaf
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    root_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Root
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    exudate_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Exudate
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    soil_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Soil
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    leaf_present =
      leaf_detected_n >= 2,

    root_present =
      root_detected_n >= 2,

    exudate_present =
      exudate_detected_n >= 2,

    soil_present =
      soil_detected_n >= 2
  )
```


48. Classify plant-compartment patterns
```{r}
plant_compartment_presence <-
  plant_compartment_presence %>%
  mutate(

    plant_pattern =
      case_when(

        leaf_present &
          root_present &
          exudate_present ~
          "Leaf + Root + Exudate",

        !leaf_present &
          root_present &
          exudate_present ~
          "Root + Exudate",

        leaf_present &
          !root_present &
          exudate_present ~
          "Leaf + Exudate",

        leaf_present &
          root_present &
          !exudate_present ~
          "Leaf + Root",

        leaf_present &
          !root_present &
          !exudate_present ~
          "Leaf only",

        !leaf_present &
          root_present &
          !exudate_present ~
          "Root only",

        !leaf_present &
          !root_present &
          exudate_present ~
          "Exudate only",

        TRUE ~
          "None consistently detected"
      )
  )
```



Summarize:

```{r}
plant_pattern_summary <-
  plant_compartment_presence %>%
  count(
    plant_pattern,
    sort = TRUE
  ) %>%
  mutate(
    percent =
      100 * n / sum(n)
  )

plant_pattern_summary
```



This is probably one of the most useful exploratory tables for your plant samples.

49. Identify possible plant-associated exudate features

Now let's get closer to your biological idea.

We want features that:

are present in exudate;
are present in root and/or leaf;
are not consistently detected in soil control.

That helps eliminate features that are simply normal soil-associated signals.

Root + exudate, absent from soil

```{r}
root_exudate_not_soil <-
  plant_compartment_presence %>%
  filter(
    root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  root_exudate_not_soil
)
```



Inspect:

```{r}
root_exudate_not_soil %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio,
    leaf_present,
    root_present,
    exudate_present,
    soil_present
  ) %>%
  head(50)
```



This is a very interesting candidate list.

50. Leaf + root + exudate, absent from soil

These may represent broadly plant-associated compounds detectable from the soil/exudate extraction.

```{r}
leaf_root_exudate_not_soil <-
  plant_compartment_presence %>%
  filter(
    leaf_present,
    root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  leaf_root_exudate_not_soil
)

```


View:

```{r}
leaf_root_exudate_not_soil %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio
  ) %>%
  head(50)
```


51. Root + exudate but NOT leaf and NOT soil

This may be even more interesting for root-associated chemistry.

```{r}
root_exudate_specific_candidates <-
  plant_compartment_presence %>%
  filter(
    !leaf_present,
    root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  root_exudate_specific_candidates
)
```



Inspect:

```{r}
root_exudate_specific_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio
  ) %>%
  head(50)
```



This category would get my attention.

Conceptually:

Leaf       absent
Root       present
Exudate    present
Soil       absent

That pattern is consistent with a root-associated compound appearing in the rhizosphere/exudate extraction.

Still not proof of secretion—but much stronger biological prioritization than just “exudate peak high.”

52. Exudate-only candidates absent from soil

You should also keep these.

A compound doesn't have to accumulate strongly inside root tissue to be released outside the plant.

```{r}
exudate_only_not_soil <-
  plant_compartment_presence %>%
  filter(
    !leaf_present,
    !root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  exudate_only_not_soil
)
```



Those could potentially include:

low-abundance secreted compounds;
compounds rapidly exported from roots;
root-derived transformation products;
microbially transformed plant compounds;
compounds produced specifically in the rhizosphere.

So do not discard them simply because they aren't detected in root tissue.

53. Save the plant-compartment candidate tables

```{r}
write.csv(
  plant_compartment_presence,
  "positive_mode_plant_compartment_presence.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  plant_pattern_summary,
  "positive_mode_plant_pattern_summary.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  root_exudate_not_soil,
  "positive_mode_root_exudate_not_soil.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  leaf_root_exudate_not_soil,
  "positive_mode_leaf_root_exudate_not_soil.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  root_exudate_specific_candidates,
  "positive_mode_root_exudate_specific_candidates.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  exudate_only_not_soil,
  "positive_mode_exudate_only_not_soil.csv",
  row.names = FALSE
)
```

54. Define all quantitative comparisons
# ------------------------------------------------------------
# COMPARISON-SPECIFIC QUANTITATIVE ANALYSIS
# ------------------------------------------------------------

```{r}
quantitative_comparisons <- list(

  # -------------------------
  # PLANT / SOIL COMPARISONS
  # -------------------------

  Exudate_vs_Soil = list(
    group1_cols = presence_groups$Exudate,
    group2_cols = presence_groups$Soil,
    group1_name = "Exudate",
    group2_name = "Soil"
  ),

  Exudate_vs_Root = list(
    group1_cols = presence_groups$Exudate,
    group2_cols = presence_groups$Root,
    group1_name = "Exudate",
    group2_name = "Root"
  ),

  Exudate_vs_Leaf = list(
    group1_cols = presence_groups$Exudate,
    group2_cols = presence_groups$Leaf,
    group1_name = "Exudate",
    group2_name = "Leaf"
  ),

  Leaf_vs_Root = list(
    group1_cols = presence_groups$Leaf,
    group2_cols = presence_groups$Root,
    group1_name = "Leaf",
    group2_name = "Root"
  ),

  # -------------------------
  # FUNGAL COMPARISONS
  # -------------------------

  Fungal_in_vs_out = list(
    group1_cols = presence_groups$Fungal_in,
    group2_cols = presence_groups$Fungal_out,
    group1_name = "Fungal in",
    group2_name = "Fungal out"
  ),

  Fungal_in_vs_control = list(
    group1_cols = presence_groups$Fungal_in,
    group2_cols = presence_groups$Fungal_control,
    group1_name = "Fungal in",
    group2_name = "Fungal control"
  ),

  Fungal_out_vs_control = list(
    group1_cols = presence_groups$Fungal_out,
    group2_cols = presence_groups$Fungal_control,
    group1_name = "Fungal out",
    group2_name = "Fungal control"
  ),

  # old fungal sample vs agar control
  Fungi_old_vs_control = list(
    group1_cols = presence_groups$Fungi_old,
    group2_cols = presence_groups$Fungal_control,
    group1_name = "Fungi old",
    group2_name = "Fungal control"
  ),

  # directly tests whether old and newer colony-underneath agar differ
  Fungi_old_vs_Fungal_in = list(
    group1_cols = presence_groups$Fungi_old,
    group2_cols = presence_groups$Fungal_in,
    group1_name = "Fungi old",
    group2_name = "Fungal in"
  ),

  # optional spatial comparison with older fungus
  Fungi_old_vs_Fungal_out = list(
    group1_cols = presence_groups$Fungi_old,
    group2_cols = presence_groups$Fungal_out,
    group1_name = "Fungi old",
    group2_name = "Fungal out"
  )
)
```



The most biologically direct fungal age comparison is:

Fungi old vs Fungal in

because both represent fungus-associated agar. If those are very similar, age/maturity may matter less than the mere presence of fungal growth.

55. Function to prepare each comparison

This separates:

detection/presence patterns;
features suitable for quantitative abundance testing.

```{r}
prepare_quantitative_comparison <- function(
    data,
    group1_cols,
    group2_cols,
    group1_name,
    group2_name,
    min_replicates = 2
) {

  detection_table <- data %>%
    mutate(

      group1_n = rowSums(
        across(
          all_of(group1_cols),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

      group2_n = rowSums(
        across(
          all_of(group2_cols),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

      detection_category = case_when(

        group1_n >= min_replicates &
          group2_n >= min_replicates ~
          "Quantifiable in both",

        group1_n >= min_replicates &
          group2_n < min_replicates ~
          paste0(
            group1_name,
            "-associated detection"
          ),

        group1_n < min_replicates &
          group2_n >= min_replicates ~
          paste0(
            group2_name,
            "-associated detection"
          ),

        TRUE ~
          "Not consistently detected in either"
      )
    )

  detection_summary <- detection_table %>%
    count(
      detection_category
    ) %>%
    mutate(
      percent =
        100 * n / sum(n)
    )

  abundance_features <- detection_table %>%
    filter(
      detection_category ==
        "Quantifiable in both"
    )

  sample_cols <- c(
    group1_cols,
    group2_cols
  )

  raw_matrix <- abundance_features %>%
    select(
      feature_name,
      all_of(sample_cols)
    ) %>%
    column_to_rownames(
      "feature_name"
    ) %>%
    as.matrix() %>%
    t()

  list(

    detection_table =
      detection_table,

    detection_summary =
      detection_summary,

    abundance_features =
      abundance_features,

    raw_matrix =
      raw_matrix,

    group1_name =
      group1_name,

    group2_name =
      group2_name,

    group1_cols =
      group1_cols,

    group2_cols =
      group2_cols
  )
}
```


56. Run all comparisons

```{r}
prepared_comparisons <- lapply(

  quantitative_comparisons,

  function(x) {

    prepare_quantitative_comparison(

      data =
        feature_qc,

      group1_cols =
        x$group1_cols,

      group2_cols =
        x$group2_cols,

      group1_name =
        x$group1_name,

      group2_name =
        x$group2_name
    )
  }
)
```



Check exudate vs soil:

```{r}
prepared_comparisons$
  Exudate_vs_Soil$
  detection_summary
```



Check fungi old vs control:

```{r}
prepared_comparisons$
  Fungi_old_vs_control$
  detection_summary
```

```{r}
# Create combined detection summary across all comparisons

detection_summary_all <- bind_rows(
  lapply(
    names(prepared_comparisons),
    function(comparison_name) {

      prepared_comparisons[[comparison_name]]$detection_summary %>%
        mutate(
          comparison = comparison_name
        )
    }
  )
) %>%
  select(
    comparison,
    detection_category,
    n,
    percent
  )

detection_summary_all
```

Save
```{r}
write.csv(
  detection_summary_all,
  "positive_mode_all_detection_summaries.csv",
  row.names = FALSE
)
```



Check old vs new/in:

```{r}
prepared_comparisons$
  Fungi_old_vs_Fungal_in$
  detection_summary
```

57. Combine detection summaries for all comparisons
```{r}
detection_summary_all <- bind_rows(
  
  lapply(
    names(prepared_comparisons),
    
    function(comparison_name) {
      
      prepared_comparisons[[comparison_name]]$detection_summary %>%
        mutate(
          comparison = comparison_name
        )
    }
  )
  
) %>%
  select(
    comparison,
    detection_category,
    n,
    percent
  )

detection_summary_all
```





Save:

```{r}
prepared_comparisons$Fungi_old_vs_Fungal_in$detection_summary
```

58. Count quantifiable features in every comparison

```{r}
quantifiable_feature_counts <- tibble(
  
  comparison = names(prepared_comparisons),
  
  quantifiable_features = sapply(
    prepared_comparisons,
    function(x) {
      nrow(x$abundance_features)
    }
  )
)

quantifiable_feature_counts
```



This table will be very useful. It tells you how much chemical overlap is available for actual abundance testing.

For example:

Exudate vs Soil               1386
Fungal in vs control          ?
Fungi old vs control          ?
Fungi old vs Fungal in        ?


# 59. Summarize remaining zero values
```{r}
zero_summary_all <- bind_rows(
  
  lapply(
    names(prepared_comparisons),
    
    function(comparison_name) {
      
      x <- prepared_comparisons[[comparison_name]]$raw_matrix
      
      tibble(
        
        comparison = comparison_name,
        
        total_values = length(x),
        
        zero_values = sum(
          x == 0,
          na.rm = TRUE
        ),
        
        zero_percent = 100 *
          sum(
            x == 0,
            na.rm = TRUE
          ) /
          length(x)
      )
    }
  )
)

zero_summary_all
```
Save:

```{r}
write.csv(
  zero_summary_all,
  "positive_mode_quantitative_zero_summary.csv",
  row.names = FALSE
)
```




60. Normalize each comparison separately

```{r}
normalize_comparison_matrix <- function(
    raw_matrix
) {

  sample_medians <- apply(
    raw_matrix,
    1,
    function(x) {

      positive_values <-
        x[
          x > 0 &
            is.finite(x)
        ]

      median(
        positive_values,
        na.rm = TRUE
      )
    }
  )

  global_median <-
    median(
      sample_medians,
      na.rm = TRUE
    )

  normalized_matrix <-
    sweep(
      raw_matrix,
      1,
      sample_medians,
      "/"
    ) *
    global_median

  log_matrix <-
    log2(
      normalized_matrix + 1
    )

  list(

    normalized =
      normalized_matrix,

    log =
      log_matrix,

    sample_medians =
      sample_medians,

    global_median =
      global_median
  )
}

```


Run:

```{r}
normalized_comparisons <- lapply(

  prepared_comparisons,

  function(x) {

    normalize_comparison_matrix(
      x$raw_matrix
    )
  }
)
```


#61. Comparison-specific PCA function. PCA Scored tell us which samples/groups separate along those PCs? Or, where the samples fall along those axes. 

```{r}
run_quantitative_pca <- function(
    log_matrix,
    metadata,
    title_text
) {

  feature_var <- apply(
    log_matrix,
    2,
    var,
    na.rm = TRUE
  )

  matrix_use <-
    log_matrix[
      ,
      is.finite(feature_var) &
        feature_var > 0,
      drop = FALSE
    ]

  pca <-
    prcomp(
      matrix_use,
      center = TRUE,
      scale. = TRUE
    )

  variance <-
    (
      pca$sdev^2 /
        sum(
          pca$sdev^2
        )
    ) * 100

  scores <-
    as.data.frame(
      pca$x
    ) %>%
    rownames_to_column(
      "sample"
    ) %>%
    left_join(
      metadata,
      by = "sample"
    )

  plot <-
    ggplot(
      scores,
      aes(
        x = PC1,
        y = PC2,
        fill = group
      )
    ) +
    geom_point(
      shape = 21,
      size = 5,
      color = "black"
    ) +
    geom_text_repel(
      aes(label = sample)
    ) +
    labs(

      title =
        title_text,

      x = paste0(
        "PC1 (",
        round(
          variance[1],
          1
        ),
        "%)"
      ),

      y = paste0(
        "PC2 (",
        round(
          variance[2],
          1
        ),
        "%)"
      )
    ) +
    theme_classic()

  list(
    pca = pca,
    variance = variance,
    scores = scores,
    plot = plot
  )
}
```


62. Run PCA for every quantitative comparison

```{r}
quantitative_pca_results <- lapply(

  names(
    normalized_comparisons
  ),

  function(comparison_name) {

    run_quantitative_pca(

      log_matrix =
        normalized_comparisons
          [[comparison_name]]
        $log,

      metadata =
        sample_metadata,

      title_text =
        paste0(
          "PCA: ",
          gsub(
            "_",
            " ",
            comparison_name
          ),
          "\nQuantifiable features only"
        )
    )
  }
)

names(
  quantitative_pca_results
) <-
  names(
    normalized_comparisons
  )
```


Inspect fungal-age PCAs

Old vs agar control:

```{r}
quantitative_pca_results$
  Fungi_old_vs_control$plot
```



Old vs new/in:

```{r}
quantitative_pca_results$
  Fungi_old_vs_Fungal_in$plot
```



Old vs new/out:

```{r}
quantitative_pca_results$
  Fungi_old_vs_Fungal_out$plot
```



These three together are particularly informative.

Conceptually:

Old vs control
→ Does older fungal growth alter the agar metabolome?

New/in vs control
→ Does current fungal growth alter the agar metabolome?

Old vs new/in
→ Does colony age/maturity alter the fungal-associated metabolome?

Old vs new/out
→ How does older fungus-associated agar compare with agar outside
  the newer colony?
  
63. Load limma

Install once in Console if necessary:

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("limma")

```


Then add near your other library() calls:

```{r}
library(limma)
```


64. Reusable limma function

```{r}
run_limma_comparison <- function(
    log_matrix,
    group1_cols,
    group2_cols,
    group1_name,
    group2_name,
    annotation_table
) {

  # limma requires rows = features,
  # columns = samples

  limma_matrix <-
    t(
      log_matrix
    )

  sample_order <-
    colnames(
      limma_matrix
    )

  comparison_group <-
    ifelse(
      sample_order %in%
        group1_cols,
      group1_name,
      group2_name
    )

  comparison_group <-
    factor(
      comparison_group,
      levels = c(
        group2_name,
        group1_name
      )
    )

  design <-
    model.matrix(
      ~ comparison_group
    )

  fit <-
    lmFit(
      limma_matrix,
      design
    )

  fit <-
    eBayes(
      fit
    )

  results <-
    topTable(
      fit,
      coef = 2,
      number = Inf,
      adjust.method = "BH",
      sort.by = "P"
    ) %>%
    rownames_to_column(
      "feature_name"
    ) %>%
    left_join(
      annotation_table,
      by = "feature_name"
    ) %>%
    mutate(

      comparison =
        paste0(
          group1_name,
          " vs ",
          group2_name
        ),

      higher_in =
        case_when(

          logFC > 0 ~
            group1_name,

          logFC < 0 ~
            group2_name,

          TRUE ~
            "Equal"
        ),

      result_category =
        case_when(

          adj.P.Val < 0.05 &
            logFC >= 1 ~
            paste0(
              group1_name,
              " enriched - FDR supported"
            ),

          adj.P.Val < 0.05 &
            logFC <= -1 ~
            paste0(
              group2_name,
              " enriched - FDR supported"
            ),

          P.Value < 0.05 &
            logFC >= 1 ~
            paste0(
              group1_name,
              " enriched - exploratory"
            ),

          P.Value < 0.05 &
            logFC <= -1 ~
            paste0(
              group2_name,
              " enriched - exploratory"
            ),

          TRUE ~
            "Not selected"
        )
    )

  results
}
```


65. Run limma for all comparisons

```{r}
limma_results <- lapply(
  names(normalized_comparisons),
  
  function(comparison_name) {
    
    comparison_info <- quantitative_comparisons[[comparison_name]]
    
    run_limma_comparison(
      
      log_matrix =
        normalized_comparisons[[comparison_name]]$log,
      
      group1_cols =
        comparison_info$group1_cols,
      
      group2_cols =
        comparison_info$group2_cols,
      
      group1_name =
        comparison_info$group1_name,
      
      group2_name =
        comparison_info$group2_name,
      
      annotation_table =
        feature_annotations
    )
  }
)

names(limma_results) <- names(normalized_comparisons)
```


66. Inspect the fungal-old statistical comparisons

Old vs control:
```{r}
head(
  limma_results$
    Fungi_old_vs_control,
  20
)
```



Old vs new/in:
```{r}

head(
  limma_results$
    Fungi_old_vs_Fungal_in,
  20
)
```


Old vs new/out:

```{r}
head(
  limma_results$
    Fungi_old_vs_Fungal_out,
  20
)
```


67. Summarize all limma comparisons

```{r}
limma_summary_all <- bind_rows(

  lapply(

    names(
      limma_results
    ),

    function(comparison_name) {

      x <-limma_results[[comparison_name]]
        

      tibble(

        comparison =
          comparison_name,

        tested_features =
          nrow(x),

        raw_p_under_05 =
          sum(
            x$P.Value < 0.05,
            na.rm = TRUE
          ),

        FDR_under_05 =
          sum(
            x$adj.P.Val < 0.05,
            na.rm = TRUE
          ),

        FDR_and_twofold =
          sum(
            x$adj.P.Val < 0.05 &
              abs(
                x$logFC
              ) >= 1,
            na.rm = TRUE
          ),

        group1_enriched_FDR =
          sum(
            x$adj.P.Val < 0.05 &
              x$logFC >= 1,
            na.rm = TRUE
          ),

        group2_enriched_FDR =
          sum(
            x$adj.P.Val < 0.05 &
              x$logFC <= -1,
            na.rm = TRUE
          )
      )
    }
  )
)

limma_summary_all
```


68. Create a fungal-only statistical summary

```{r}
fungal_limma_summary <-
  limma_summary_all %>%
  filter(
    comparison %in% c(
      "Fungal_in_vs_out",
      "Fungal_in_vs_control",
      "Fungal_out_vs_control",
      "Fungi_old_vs_control",
      "Fungi_old_vs_Fungal_in",
      "Fungi_old_vs_Fungal_out"
    )
  )

fungal_limma_summary

```


This will let you compare, side by side, how many features differ in each fungal contrast.

69. Find features shared between old and new fungal effects

This is especially useful for your age question.

First get features significantly different from control in the old fungal group:

```{r}
old_vs_control_sig <-
  limma_results$
    Fungi_old_vs_control %>%
  filter(
    adj.P.Val < 0.05,
    abs(logFC) >= 1
  ) %>%
  pull(
    feature_name
  )
```



Get features significantly different from control in new/in:

```{r}

new_in_vs_control_sig <-
  limma_results$
    Fungal_in_vs_control %>%
  filter(
    adj.P.Val < 0.05,
    abs(logFC) >= 1
  ) %>%
  pull(
    feature_name
  )
```


Shared fungal-effect features:

```{r}
shared_old_new_fungal_features <-
  intersect(
    old_vs_control_sig,
    new_in_vs_control_sig
  )

length(
  shared_old_new_fungal_features
)
```


Inspect them:

```{r}
shared_old_new_fungal_results <-
  feature_annotations %>%
  filter(
    feature_name %in%
      shared_old_new_fungal_features
  )

shared_old_new_fungal_results
```


These are strong candidates for:

features associated with fungal growth regardless of colony age.

70. Features different between old and new fungal colonies

```{r}
age_sensitive_fungal_features <-
  limma_results$
    Fungi_old_vs_Fungal_in %>%
  filter(
    adj.P.Val < 0.05,
    abs(logFC) >= 1
  )

age_sensitive_fungal_features
```



If this table is small while both old-vs-control and new-in-vs-control are large, that would support:

fungal presence strongly affects the agar chemistry, while colony age has a smaller effect.

If this table is large, age/maturity matters more.

71. Quantify overlap between old and new fungal effects

```{r}
fungal_overlap_summary <- tibble(

  old_vs_control_significant =
    length(
      old_vs_control_sig
    ),

  new_in_vs_control_significant =
    length(
      new_in_vs_control_sig
    ),

  shared_significant_features =
    length(
      shared_old_new_fungal_features
    ),

  old_only =
    length(
      setdiff(
        old_vs_control_sig,
        new_in_vs_control_sig
      )
    ),

  new_only =
    length(
      setdiff(
        new_in_vs_control_sig,
        old_vs_control_sig
      )
    )
)

fungal_overlap_summary
```


72. Save the new fungal outputs

```{r}
write.csv(
  fungal_limma_summary,
  "positive_mode_fungal_limma_summary.csv",
  row.names = FALSE
)

write.csv(
  shared_old_new_fungal_results,
  "positive_mode_shared_old_new_fungal_features.csv",
  row.names = FALSE
)

write.csv(
  age_sensitive_fungal_features,
  "positive_mode_age_sensitive_fungal_features.csv",
  row.names = FALSE
)

write.csv(
  fungal_overlap_summary,
  "positive_mode_fungal_overlap_summary.csv",
  row.names = FALSE
)
```



And save all limma tables:

```{r}
for (
  comparison_name in
  names(limma_results)
) {

  write.csv(

    limma_results
      [[comparison_name]]
    ,

    paste0(
      "positive_mode_limma_",
      comparison_name,
      ".csv"
    ),

    row.names = FALSE
  )
}

```


The first three outputs I’d examine after running this are:

```{r}
quantifiable_feature_counts
fungal_limma_summary
```



and:

```{r}
fungal_overlap_summary
```

73. Create common analytical flags for candidate features

We want to distinguish:

confidently retained RTs from very early RTs;
features with MS/MS;
features with an actual MS-DIAL metabolite annotation.

# ------------------------------------------------------------
# CANDIDATE PRIORITIZATION
# ------------------------------------------------------------

```{r}
candidate_annotations <- feature_annotations %>%
  mutate(

    RT_flag = case_when(
      `Average Rt(min)` < 0.5 ~
        "Very early RT",

      `Average Rt(min)` < 1.0 ~
        "Early RT",

      TRUE ~
        "RT >= 1 min"
    ),

    MSMS_flag = case_when(
      `MS/MS assigned` == TRUE ~
        "MS/MS available",

      TRUE ~
        "No MS/MS assignment"
    ),

    annotation_flag = case_when(

      !is.na(`Metabolite name`) &
        `Metabolite name` != "" &
        `Metabolite name` != "Unknown" ~
        "Annotated",

      TRUE ~
        "Unknown"
    )
  )
```



Check:

```{r}
candidate_annotations %>%
  count(
    RT_flag,
    MSMS_flag,
    annotation_flag
  )
```


74. Create plant-compartment annotation table

This gives every feature its leaf/root/exudate/soil detection pattern.

```{r}
plant_compartment_flags <-
  plant_compartment_presence %>%
  select(
    feature_name,
    leaf_detected_n,
    root_detected_n,
    exudate_detected_n,
    soil_detected_n,
    leaf_present,
    root_present,
    exudate_present,
    soil_present,
    plant_pattern
  )
```


75. Rank abundance-based exudate vs soil candidates

These are different from the presence-only candidates.

These features were sufficiently detected in both exudate and soil to enter limma.

```{r}
exudate_soil_abundance_candidates <-
  limma_results$Exudate_vs_Soil %>%

  left_join(
    plant_compartment_flags,
    by = "feature_name"
  ) %>%

  left_join(
    candidate_annotations %>%
      select(
        feature_name,
        RT_flag,
        MSMS_flag,
        annotation_flag
      ),
    by = "feature_name"
  ) %>%

  mutate(

    evidence_tier = case_when(

      adj.P.Val < 0.05 &
        abs(logFC) >= 1 ~
        "Tier 1: FDR + >=2-fold",

      adj.P.Val < 0.05 ~
        "Tier 2: FDR supported",

      P.Value < 0.05 &
        abs(logFC) >= 1 ~
        "Tier 3: exploratory + >=2-fold",

      abs(logFC) >= 1 ~
        "Tier 4: large effect only",

      TRUE ~
        "Tier 5: low priority"
    )
  ) %>%

  arrange(
    factor(
      evidence_tier,
      levels = c(
        "Tier 1: FDR + >=2-fold",
        "Tier 2: FDR supported",
        "Tier 3: exploratory + >=2-fold",
        "Tier 4: large effect only",
        "Tier 5: low priority"
      )
    ),
    adj.P.Val,
    desc(abs(logFC))
  )

```


Inspect the strongest:

```{r}
exudate_soil_abundance_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    RT_flag,
    leaf_present,
    root_present,
    exudate_present,
    soil_present,
    logFC,
    P.Value,
    adj.P.Val,
    evidence_tier
  ) %>%
  head(50)
```


76. Count exudate vs soil candidate tiers
```{r}
exudate_soil_abundance_candidates %>%
  count(
    evidence_tier
  )
```



And specifically:

```{r}
sum(
  exudate_soil_abundance_candidates$
    adj.P.Val < 0.05,
  na.rm = TRUE
)
```


77. Create detection-based plant candidate table

Now handle the features that may be biologically very interesting because of differential detection.

These do not need to be quantifiable in soil.

```{r}
plant_detection_candidates <-
  plant_compartment_presence %>%

  filter(
    exudate_present,
    !soil_present
  ) %>%

  left_join(
    candidate_annotations %>%
      select(
        feature_name,
        RT_flag,
        MSMS_flag,
        annotation_flag
      ),
    by = "feature_name"
  ) %>%

  mutate(

    plant_candidate_class = case_when(

      !leaf_present &
        root_present &
        exudate_present &
        !soil_present ~
        "Root + Exudate; absent Leaf/Soil",

      leaf_present &
        root_present &
        exudate_present &
        !soil_present ~
        "Leaf + Root + Exudate; absent Soil",

      !leaf_present &
        !root_present &
        exudate_present &
        !soil_present ~
        "Exudate only; absent Soil",

      leaf_present &
        !root_present &
        exudate_present &
        !soil_present ~
        "Leaf + Exudate; absent Root/Soil",

      TRUE ~
        "Other plant-associated pattern"
    )
  )
```


78. Add raw group means to detection-based plant candidates

For these candidates, it is useful to see the actual MS-DIAL peak-height pattern.

```{r}
plant_detection_candidates <-
  plant_detection_candidates %>%
  mutate(

    leaf_mean_raw =
      rowMeans(
        across(
          all_of(
            presence_groups$Leaf
          )
        ),
        na.rm = TRUE
      ),

    root_mean_raw =
      rowMeans(
        across(
          all_of(
            presence_groups$Root
          )
        ),
        na.rm = TRUE
      ),

    exudate_mean_raw =
      rowMeans(
        across(
          all_of(
            presence_groups$Exudate
          )
        ),
        na.rm = TRUE
      ),

    soil_mean_raw =
      rowMeans(
        across(
          all_of(
            presence_groups$Soil
          )
        ),
        na.rm = TRUE
      )
  )

```


Rank:

```{r}
plant_detection_candidates <-
  plant_detection_candidates %>%
  arrange(
    factor(
      plant_candidate_class,
      levels = c(
        "Root + Exudate; absent Leaf/Soil",
        "Leaf + Root + Exudate; absent Soil",
        "Exudate only; absent Soil",
        "Leaf + Exudate; absent Root/Soil",
        "Other plant-associated pattern"
      )
    ),
    desc(
      sample_to_blank_ratio
    )
  )

```


Inspect:

```{r}
plant_detection_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    RT_flag,
    plant_candidate_class,
    leaf_mean_raw,
    root_mean_raw,
    exudate_mean_raw,
    soil_mean_raw,
    sample_to_blank_ratio
  ) %>%
  head(50)
```


79. Specifically rank root + exudate candidates

```{r}
ranked_root_exudate_candidates <-
  plant_detection_candidates %>%

  filter(
    plant_candidate_class ==
      "Root + Exudate; absent Leaf/Soil"
  ) %>%

  mutate(

    priority = case_when(

      `MS/MS assigned` == TRUE &
        `Average Rt(min)` >= 1 ~
        "Priority 1",

      `Average Rt(min)` >= 1 ~
        "Priority 2",

      `MS/MS assigned` == TRUE ~
        "Priority 3",

      TRUE ~
        "Priority 4"
    )
  ) %>%

  arrange(
    priority,
    desc(
      sample_to_blank_ratio
    )
  )

```


View:

```{r}
ranked_root_exudate_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `MS/MS assigned`,
    RT_flag,
    root_mean_raw,
    exudate_mean_raw,
    soil_mean_raw,
    sample_to_blank_ratio,
    priority
  )
```



This is where your earlier five root+exudate candidates should appear.

80. Rank exudate-only candidates

```{r}
ranked_exudate_only_candidates <-
  plant_detection_candidates %>%

  filter(
    plant_candidate_class ==
      "Exudate only; absent Soil"
  ) %>%

  mutate(

    priority = case_when(

      `MS/MS assigned` == TRUE &
        `Average Rt(min)` >= 1 ~
        "Priority 1",

      `Average Rt(min)` >= 1 ~
        "Priority 2",

      `MS/MS assigned` == TRUE ~
        "Priority 3",

      TRUE ~
        "Priority 4"
    )
  ) %>%

  arrange(
    priority,
    desc(
      sample_to_blank_ratio
    )
  )
```


81. Create fungal differential-detection candidate tables

Now do the same concept for fungi.

Helper:

```{r}
extract_detection_candidates <- function(
    comparison_name
) {

  prepared_comparisons[[comparison_name]] $detection_table %>%

    filter(
      detection_category !=
        "Quantifiable in both",

      detection_category !=
        "Not consistently detected in either"
    ) %>%

    left_join(
      candidate_annotations %>%
        select(
          feature_name,
          RT_flag,
          MSMS_flag,
          annotation_flag
        ),
      by = "feature_name"
    )
}

```


Important: keep [[comparison_name]] together. Use this exact corrected function instead:

```{r}
extract_detection_candidates <- function(comparison_name) {

  x <- prepared_comparisons[[comparison_name]]$detection_table

  x %>%
    filter(
      detection_category != "Quantifiable in both",
      detection_category != "Not consistently detected in either"
    ) %>%
    left_join(
      candidate_annotations %>%
        select(
          feature_name,
          RT_flag,
          MSMS_flag,
          annotation_flag
        ),
      by = "feature_name"
    )
}

```


Run:

```{r}
fungal_in_control_detection <-
  extract_detection_candidates(
    "Fungal_in_vs_control"
  )

fungal_out_control_detection <-
  extract_detection_candidates(
    "Fungal_out_vs_control"
  )

fungi_old_control_detection <-
  extract_detection_candidates(
    "Fungi_old_vs_control"
  )

fungi_old_new_detection <-
  extract_detection_candidates(
    "Fungi_old_vs_Fungal_in"
  )
```


82. Inspect old-fungus differential detection

```{r}
fungi_old_control_detection %>%
  count(
    detection_category
  )

```


View candidates:
```{r}
fungi_old_control_detection %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `MS/MS assigned`,
    detection_category,
    RT_flag,
    sample_to_blank_ratio
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  ) %>%
  head(50)
```


83. Build a master fungal abundance table

This is one of the most important fungal analyses.

We will combine:

old fungi vs control;
new/in fungi vs control;
old vs new/in.

First extract only the statistical columns.

```{r}
old_control_stats <-
  limma_results$Fungi_old_vs_control %>%
  select(
    feature_name,
    old_control_logFC = logFC,
    old_control_p = P.Value,
    old_control_FDR = adj.P.Val
  )
new_control_stats <-
  limma_results$Fungal_in_vs_control %>%
  select(
    feature_name,
    new_control_logFC = logFC,
    new_control_p = P.Value,
    new_control_FDR = adj.P.Val
  )
old_new_stats <-
  limma_results$Fungi_old_vs_Fungal_in %>%
  select(
    feature_name,
    old_new_logFC = logFC,
    old_new_p = P.Value,
    old_new_FDR = adj.P.Val
  )
```



Combine:

```{r}
fungal_age_master <-
  full_join(
    old_control_stats,
    new_control_stats,
    by = "feature_name"
  ) %>%

  full_join(
    old_new_stats,
    by = "feature_name"
  ) %>%

  left_join(
    candidate_annotations,
    by = "feature_name"
  )
```


84. Classify fungal age effects

This directly addresses your biological hypothesis.

```{r}
fungal_age_master <-
  fungal_age_master %>%

  mutate(

    old_significant =
      !is.na(old_control_FDR) &
      old_control_FDR < 0.05,

    new_significant =
      !is.na(new_control_FDR) &
      new_control_FDR < 0.05,

    age_significant =
      !is.na(old_new_FDR) &
      old_new_FDR < 0.05,

    same_direction =
      !is.na(old_control_logFC) &
      !is.na(new_control_logFC) &
      sign(old_control_logFC) ==
        sign(new_control_logFC),

    fungal_age_category =
      case_when(

        old_significant &
          new_significant &
          same_direction &
          !age_significant ~
          "Age-stable fungal effect",

        age_significant ~
          "Age-sensitive fungal effect",

        old_significant &
          !new_significant ~
          "Old-associated effect",

        !old_significant &
          new_significant ~
          "New-associated effect",

        old_significant &
          new_significant &
          !same_direction ~
          "Opposite old/new responses",

        TRUE ~
          "No strong fungal-age classification"
      )
  )
```


85. Summarize fungal age categories

```{r}
fungal_age_summary <-
  fungal_age_master %>%
  count(
    fungal_age_category,
    sort = TRUE
  )

fungal_age_summary
```



This is a major result.

If you see lots of:

Age-stable fungal effect

and relatively few:

Age-sensitive fungal effect

that supports:

The fungus changes agar chemistry, but colony age has a smaller effect.

86. Examine age-stable fungal features

```{r}
age_stable_fungal_features <-
  fungal_age_master %>%

  filter(
    fungal_age_category ==
      "Age-stable fungal effect"
  ) %>%

  arrange(
    pmax(
      old_control_FDR,
      new_control_FDR,
      na.rm = TRUE
    )
  )
```



View:

```{r}
age_stable_fungal_features %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `MS/MS assigned`,
    RT_flag,
    old_control_logFC,
    old_control_FDR,
    new_control_logFC,
    new_control_FDR,
    old_new_logFC,
    old_new_FDR
  ) %>%
  head(50)
```


87. Examine age-sensitive fungal features

```{r}
age_sensitive_fungal_ranked <-
  fungal_age_master %>%

  filter(
    fungal_age_category ==
      "Age-sensitive fungal effect"
  ) %>%

  arrange(
    old_new_FDR,
    desc(
      abs(
        old_new_logFC
      )
    )
  )

```


View:

```{r}
age_sensitive_fungal_ranked %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `MS/MS assigned`,
    RT_flag,
    old_new_logFC,
    old_new_p,
    old_new_FDR,
    old_control_logFC,
    new_control_logFC
  ) %>%
  head(50)
```


88. Create top plant abundance candidate heatmap

Take the top 30 exudate-vs-soil abundance candidates.

```{r}
top_exudate_soil_features <-
  exudate_soil_abundance_candidates %>%

  filter(
    adj.P.Val < 0.05
  ) %>%

  arrange(
    adj.P.Val,
    desc(
      abs(logFC)
    )
  ) %>%

  slice_head(
    n = 30
  ) %>%

  pull(
    feature_name
  )
```



Only keep features actually present in the quantitative matrix:

```{r}
top_exudate_soil_features <-
  intersect(
    top_exudate_soil_features,
    colnames(
      normalized_comparisons$
        Exudate_vs_Soil$log
    )
  )
```



Heatmap:

```{r}
if (
  length(
    top_exudate_soil_features
  ) >= 2
) {

  exudate_soil_heatmap <-
    t(
      normalized_comparisons$
        Exudate_vs_Soil$log[
          ,
          top_exudate_soil_features,
          drop = FALSE
        ]
    )

  pheatmap(
    exudate_soil_heatmap,
    scale = "row",
    clustering_distance_rows =
      "euclidean",
    clustering_distance_cols =
      "euclidean",
    clustering_method =
      "complete",
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = 7,
    main =
      "Top differential exudate vs soil features"
  )
}
```


```{r}
top_age_features <-
  age_sensitive_fungal_ranked %>%

  filter(
    old_new_FDR < 0.05
  ) %>%

  slice_head(
    n = 30
  ) %>%

  pull(
    feature_name
  )
```


Match:


```{r}
top_age_features <-
  intersect(
    top_age_features,
    colnames(
      normalized_comparisons$
        Fungi_old_vs_Fungal_in$log
    )
  )
```

Plot:


```{r}
if (
  length(
    top_age_features
  ) >= 2
) {

  fungal_age_heatmap <-
    t(
      normalized_comparisons$
        Fungi_old_vs_Fungal_in$log[
          ,
          top_age_features,
          drop = FALSE
        ]
    )

  pheatmap(
    fungal_age_heatmap,
    scale = "row",
    clustering_distance_rows =
      "euclidean",
    clustering_distance_cols =
      "euclidean",
    clustering_method =
      "complete",
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize_row = 7,
    main =
      "Features differing between old and newer fungal growth"
  )
}
```



90. Create reusable individual-feature plot function

This lets you inspect actual replicates for any limma comparison.


```{r}
plot_comparison_feature <- function(
    comparison_name,
    feature_to_plot
) {

  x <-
    normalized_comparisons[[comparison_name]]$log

  plot_data <-
    as.data.frame(x) %>%
    rownames_to_column(
      "sample"
    ) %>%

    select(
      sample,
      all_of(
        feature_to_plot
      )
    ) %>%

    rename(
      log2_intensity =
        all_of(
          feature_to_plot
        )
    ) %>%

    left_join(
      sample_metadata,
      by = "sample"
    )

  ggplot(
    plot_data,
    aes(
      x = group,
      y = log2_intensity,
      fill = group
    )
  ) +

    geom_boxplot(
      width = 0.5,
      outlier.shape = NA,
      alpha = 0.5
    ) +

    geom_jitter(
      width = 0.08,
      size = 3
    ) +

    labs(
      title =
        feature_to_plot,
      x = NULL,
      y =
        "Log2 normalized intensity"
    ) +

    theme_classic() +

    theme(
      axis.text.x =
        element_text(
          angle = 45,
          hjust = 1
        ),
      legend.position =
        "none"
    )
}
```


Again, fix the indexing formatting to keep brackets together. Use this exact function:



```{r}
plot_comparison_feature <- function(
    comparison_name,
    feature_to_plot
) {

  x <- normalized_comparisons[[comparison_name]]$log

  plot_data <-
    as.data.frame(x) %>%
    rownames_to_column("sample") %>%
    select(
      sample,
      all_of(feature_to_plot)
    )

  names(plot_data)[2] <-
    "log2_intensity"

  plot_data <-
    plot_data %>%
    left_join(
      sample_metadata,
      by = "sample"
    )

  ggplot(
    plot_data,
    aes(
      x = group,
      y = log2_intensity,
      fill = group
    )
  ) +
    geom_boxplot(
      width = 0.5,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    geom_jitter(
      width = 0.08,
      size = 3
    ) +
    labs(
      title = feature_to_plot,
      x = NULL,
      y = "Log2 normalized intensity"
    ) +
    theme_classic() +
    theme(
      axis.text.x =
        element_text(
          angle = 45,
          hjust = 1
        ),
      legend.position = "none"
    )
}
```



91. Plot strongest exudate vs soil feature



```{r}
top_exudate_abundance_feature <-
  exudate_soil_abundance_candidates %>%

  filter(
    adj.P.Val < 0.05
  ) %>%

  arrange(
    adj.P.Val,
    desc(
      abs(logFC)
    )
  ) %>%

  slice(1) %>%

  pull(
    feature_name
  )
```


Plot:


```{r}
plot_comparison_feature(
  "Exudate_vs_Soil",
  top_exudate_abundance_feature
)
```


92. Plot strongest fungal-age feature



```{r}
top_fungal_age_feature <-
  age_sensitive_fungal_ranked %>%

  slice(1) %>%

  pull(
    feature_name
  )
plot_comparison_feature(
  "Fungi_old_vs_Fungal_in",
  top_fungal_age_feature
)
```

93. Create annotation-priority subset

Now identify candidates for actual chemical-identification work.

```{r}
annotated_exudate_candidates <-
  exudate_soil_abundance_candidates %>%

  filter(
    annotation_flag ==
      "Annotated" |
      `MS/MS assigned` ==
        TRUE
  ) %>%

  arrange(
    adj.P.Val,
    desc(
      abs(logFC)
    )
  )
```



View:


```{r}
annotated_exudate_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    Formula,
    Ontology,
    `MS/MS assigned`,
    `Total score`,
    RT_flag,
    logFC,
    P.Value,
    adj.P.Val
  ) %>%
  head(50)
```


94. Annotation-priority fungal candidates

```{r}
annotated_fungal_candidates <-
  fungal_age_master %>%

  filter(
    annotation_flag ==
      "Annotated" |
      `MS/MS assigned` ==
        TRUE
  ) %>%

  filter(
    fungal_age_category !=
      "No strong fungal-age classification"
  ) %>%

  arrange(
    old_new_FDR
  )
```


95. Build a master candidate table



This gives you a compact set of high-interest positive-mode LC-MS candidates.

First abundance-based plant candidates:


```{r}
master_plant_abundance_candidates <-
  exudate_soil_abundance_candidates %>%

  filter(
    evidence_tier %in% c(
      "Tier 1: FDR + >=2-fold",
      "Tier 2: FDR supported",
      "Tier 3: exploratory + >=2-fold"
    )
  ) %>%

  mutate(
    candidate_source =
      "Exudate vs Soil abundance"
  )
```


Detection candidates:



```{r}
master_plant_detection_candidates <-
  plant_detection_candidates %>%

  filter(
    plant_candidate_class %in% c(
      "Root + Exudate; absent Leaf/Soil",
      "Leaf + Root + Exudate; absent Soil",
      "Exudate only; absent Soil"
    )
  ) %>%

  mutate(
    candidate_source =
      "Plant compartment detection"
  )
```



Because these tables have different statistical columns, keep them as separate master outputs rather than forcing them into one malformed table.


96. Create a standards-review table

This is the table you will eventually use to ask:

Is this candidate sufficiently interesting and sufficiently identified to justify an authentic chemical standard?

For abundance candidates:


```{r}
standards_review_abundance <-
  exudate_soil_abundance_candidates %>%

  filter(
    evidence_tier %in% c(
      "Tier 1: FDR + >=2-fold",
      "Tier 2: FDR supported",
      "Tier 3: exploratory + >=2-fold"
    )
  ) %>%

  transmute(

    feature_name,

    RT =
      `Average Rt(min)`,

    mz =
      `Average Mz`,

    metabolite_name =
      `Metabolite name`,

    adduct =
      `Adduct type`,

    formula =
      Formula,

    ontology =
      Ontology,

    MSMS =
      `MS/MS assigned`,

    RT_flag,

    blank_ratio =
      sample_to_blank_ratio,

    log2_fold_change =
      logFC,

    raw_p =
      P.Value,

    FDR =
      adj.P.Val,

    evidence_tier,

    possible_standard =
      case_when(

        annotation_flag ==
          "Annotated" &
          `MS/MS assigned` ==
          TRUE &
          RT_flag ==
          "RT >= 1 min" ~
          "High priority to investigate standard",

        annotation_flag ==
          "Annotated" ~
          "Investigate annotation first",

        `MS/MS assigned` ==
          TRUE ~
          "Unknown feature - investigate MS/MS",

        TRUE ~
          "Identification needed first"
      )
  )
```


View:


```{r}
standards_review_abundance
```


97. Save all candidate tables



```{r}
write.csv(
  exudate_soil_abundance_candidates,
  "positive_mode_ranked_exudate_soil_abundance_candidates.csv",
  row.names = FALSE
)

write.csv(
  plant_detection_candidates,
  "positive_mode_plant_detection_candidates.csv",
  row.names = FALSE
)

write.csv(
  ranked_root_exudate_candidates,
  "positive_mode_ranked_root_exudate_candidates.csv",
  row.names = FALSE
)

write.csv(
  ranked_exudate_only_candidates,
  "positive_mode_ranked_exudate_only_candidates.csv",
  row.names = FALSE
)

write.csv(
  fungal_age_master,
  "positive_mode_fungal_age_master.csv",
  row.names = FALSE
)

write.csv(
  fungal_age_summary,
  "positive_mode_fungal_age_summary.csv",
  row.names = FALSE
)

write.csv(
  age_stable_fungal_features,
  "positive_mode_age_stable_fungal_features.csv",
  row.names = FALSE
)

write.csv(
  age_sensitive_fungal_ranked,
  "positive_mode_age_sensitive_fungal_features.csv",
  row.names = FALSE
)

write.csv(
  annotated_exudate_candidates,
  "positive_mode_annotated_exudate_candidates.csv",
  row.names = FALSE
)

write.csv(
  annotated_fungal_candidates,
  "positive_mode_annotated_fungal_candidates.csv",
  row.names = FALSE
)

write.csv(
  standards_review_abundance,
  "positive_mode_candidates_for_standard_review.csv",
  row.names = FALSE
)
```


98. Save fungal differential-detection tables


```{r}
write.csv(
  fungal_in_control_detection,
  "positive_mode_fungal_in_control_detection_candidates.csv",
  row.names = FALSE
)

write.csv(
  fungal_out_control_detection,
  "positive_mode_fungal_out_control_detection_candidates.csv",
  row.names = FALSE
)

write.csv(
  fungi_old_control_detection,
  "positive_mode_fungi_old_control_detection_candidates.csv",
  row.names = FALSE
)

write.csv(
  fungi_old_new_detection,
  "positive_mode_fungi_old_vs_new_detection_candidates.csv",
  row.names = FALSE
)
```


99. Save important R objects

This is useful because you now have a lot of analysis objects.


```{r}
saveRDS(
  limma_results,
  "positive_mode_limma_results.rds"
)

saveRDS(
  prepared_comparisons,
  "positive_mode_prepared_comparisons.rds"
)

saveRDS(
  normalized_comparisons,
  "positive_mode_normalized_comparisons.rds"
)

saveRDS(
  fungal_age_master,
  "positive_mode_fungal_age_master.rds"
)
```


Later you can reload one with:

```{r}
limma_results <-
  readRDS(
    "positive_mode_limma_results.rds"
  )
```


100. Record the R session for reproducibility
```{r}
sessionInfo()

```

Save:
```{r}
writeLines(
  capture.output(
    sessionInfo()
  ),
  "positive_mode_R_session_info.txt"
)
```



After you run these sections, the most useful outputs to show me are:

```{r}
fungal_age_summary
exudate_soil_abundance_candidates %>%
  count(evidence_tier)
ranked_root_exudate_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `MS/MS assigned`,
    sample_to_blank_ratio,
    priority
  )

```


and:

```{r}
standards_review_abundance %>%
  head(30)
```



Plot the 5 candidate features across plant compartments
# ------------------------------------------------------------
# Plot 5 root-exudate candidate features across compartments
# ------------------------------------------------------------


```{r}
five_candidate_features <- c(
  "F1207_mz164.1072_rt1.26",
  "F5091_mz739.7451_rt9.7",
  "F2544_mz260.1973_rt0.9",
  "F4601_mz494.8109_rt0.42",
  "F1085_mz155.0428_rt0.44"
)

plant_sample_cols <- c(
  presence_groups$Leaf,
  presence_groups$Root,
  presence_groups$Exudate,
  presence_groups$Soil
)
```



Now reshape the raw intensity data:

```{r}
five_feature_plot_data <- feature_qc %>%
  
  filter(
    feature_name %in% five_candidate_features
  ) %>%
  
  select(
    feature_name,
    all_of(plant_sample_cols)
  ) %>%
  
  pivot_longer(
    cols = all_of(plant_sample_cols),
    names_to = "sample",
    values_to = "raw_intensity"
  ) %>%
  
  left_join(
    sample_metadata %>%
      select(
        sample,
        group
      ),
    by = "sample"
  )

```


Check it:

five_feature_plot_data
Give the compartments a biological order


```{r}
five_feature_plot_data <- five_feature_plot_data %>%
  mutate(
    
    group = factor(
      group,
      levels = c(
        "Leaf",
        "Root",
        "Plant-soil exudate",
        "Soil control"
      )
    ),
    
    feature_name = factor(
      feature_name,
      levels = five_candidate_features
    )
  )
```


Plot raw LC-MS intensities
```{r}
ggplot(
  five_feature_plot_data,
  aes(
    x = group,
    y = raw_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title = "Root-exudate candidate features across plant compartments",
    subtitle = "Positive-mode LC-MS; points represent biological replicates",
    x = NULL,
    y = "Raw LC-MS feature intensity"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    
    strip.text = element_text(
      face = "bold"
    )
  )
```


I also want you to make a log2 version

This will probably be much easier to interpret, because LC-MS peak heights can differ by orders of magnitude.

Use log2(intensity + 1) so zeros remain zeros:

```{r}
five_feature_plot_data <- five_feature_plot_data %>%
  mutate(
    log2_intensity =
      log2(raw_intensity + 1)
  )

```


Then:

```{r}
ggplot(
  five_feature_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title = "Root-exudate candidate features across plant compartments",
    subtitle = "Positive-mode LC-MS; log2-transformed raw feature intensities",
    x = NULL,
    y = "log2(Intensity + 1)"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    
    strip.text = element_text(
      face = "bold"
    )
  )
```


And make one summary table

This will let us see the numbers behind the plots:

```{r}
five_feature_summary <- five_feature_plot_data %>%
  group_by(
    feature_name,
    group
  ) %>%
  summarise(
    
    n_detected =
      sum(
        raw_intensity > 0,
        na.rm = TRUE
      ),
    
    mean_intensity =
      mean(
        raw_intensity,
        na.rm = TRUE
      ),
    
    median_intensity =
      median(
        raw_intensity,
        na.rm = TRUE
      ),
    
    sd_intensity =
      sd(
        raw_intensity,
        na.rm = TRUE
      ),
    
    .groups = "drop"
  )

five_feature_summary
```


Plot the 5 candidate features across plant compartments
# ------------------------------------------------------------
# Plot 5 root-exudate candidate features across compartments
# ------------------------------------------------------------

```{r}
five_candidate_features <- c(
  "F1207_mz164.1072_rt1.26",
  "F5091_mz739.7451_rt9.7",
  "F2544_mz260.1973_rt0.9",
  "F4601_mz494.8109_rt0.42",
  "F1085_mz155.0428_rt0.44"
)

plant_sample_cols <- c(
  presence_groups$Leaf,
  presence_groups$Root,
  presence_groups$Exudate,
  presence_groups$Soil
)
```



Now reshape the raw intensity data:
```{r}
five_feature_plot_data <- feature_qc %>%
  
  filter(
    feature_name %in% five_candidate_features
  ) %>%
  
  select(
    feature_name,
    all_of(plant_sample_cols)
  ) %>%
  
  pivot_longer(
    cols = all_of(plant_sample_cols),
    names_to = "sample",
    values_to = "raw_intensity"
  ) %>%
  
  left_join(
    sample_metadata %>%
      select(
        sample,
        group
      ),
    by = "sample"
  )

```


Check it:
five_feature_plot_data
Give the compartments a biological order


```{r}
five_feature_plot_data <- five_feature_plot_data %>%
  mutate(
    
    group = factor(
      group,
      levels = c(
        "Leaf",
        "Root",
        "Plant-soil exudate",
        "Soil control"
      )
    ),
    
    feature_name = factor(
      feature_name,
      levels = five_candidate_features
    )
  )
```


Plot raw LC-MS intensities


```{r}
ggplot(
  five_feature_plot_data,
  aes(
    x = group,
    y = raw_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title = "Root-exudate candidate features across plant compartments",
    subtitle = "Positive-mode LC-MS; points represent biological replicates",
    x = NULL,
    y = "Raw LC-MS feature intensity"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    
    strip.text = element_text(
      face = "bold"
    )
  )
```


I also want you to make a log2 version

This will probably be much easier to interpret, because LC-MS peak heights can differ by orders of magnitude.

Use log2(intensity + 1) so zeros remain zeros:

```{r}
five_feature_plot_data <- five_feature_plot_data %>%
  mutate(
    log2_intensity =
      log2(raw_intensity + 1)
  )
```



Then:

```{r}
ggplot(
  five_feature_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title = "Root-exudate candidate features across plant compartments",
    subtitle = "Positive-mode LC-MS; log2-transformed raw feature intensities",
    x = NULL,
    y = "log2(Intensity + 1)"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    
    strip.text = element_text(
      face = "bold"
    )
  )
```


And make one summary table

This will let us see the numbers behind the plots:

```{r}
five_feature_summary <- five_feature_plot_data %>%
  group_by(
    feature_name,
    group
  ) %>%
  summarise(
    
    n_detected =
      sum(
        raw_intensity > 0,
        na.rm = TRUE
      ),
    
    mean_intensity =
      mean(
        raw_intensity,
        na.rm = TRUE
      ),
    
    median_intensity =
      median(
        raw_intensity,
        na.rm = TRUE
      ),
    
    sd_intensity =
      sd(
        raw_intensity,
        na.rm = TRUE
      ),
    
    .groups = "drop"
  )

five_feature_summary
```

101. create fungal sample columns
```{r}
# ------------------------------------------------------------
# FUNGAL COMPARTMENT / AGE FEATURE PLOTS
# ------------------------------------------------------------

fungal_sample_cols <- c(
  presence_groups$Fungal_control,
  presence_groups$Fungal_out,
  presence_groups$Fungal_in,
  presence_groups$Fungi_old
)

fungal_sample_cols
```
102. Make fungal data long-format
```{r}
fungal_long <- feature_qc %>%
  
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    all_of(fungal_sample_cols)
  ) %>%
  
  pivot_longer(
    cols = all_of(fungal_sample_cols),
    names_to = "sample",
    values_to = "raw_intensity"
  ) %>%
  
  left_join(
    sample_metadata %>%
      select(
        sample,
        group
      ),
    by = "sample"
  )
```

check
```{r}
fungal_long %>%
  count(group)
```
103 Put the fungal conditions in biological order
```{r}
fungal_long <- fungal_long %>%
  mutate(
    
    group = factor(
      group,
      levels = c(
        "Fungal control",
        "Fungal agar out",
        "Fungal agar in",
        "Fungi old"
      )
    ),
    
    log2_intensity =
      log2(raw_intensity + 1)
  )
```

104. First identify the strongest old-vs-new features
```{r}
top_fungal_age_candidates <-
  age_sensitive_fungal_ranked %>%
  
  filter(
    old_new_FDR < 0.05
  ) %>%
  
  arrange(
    old_new_FDR,
    desc(abs(old_new_logFC))
  ) %>%
  
  slice_head(
    n = 10
  ) %>%
  
  pull(feature_name)

top_fungal_age_candidates
```

105. Plot the top 10 across all foud fungal conditions
```{r}
top_fungal_plot_data <-
  fungal_long %>%
  
  filter(
    feature_name %in%
      top_fungal_age_candidates
  )
```


plot the raw intensities Which metabolites/features change as the fungus ages?
```{r}
ggplot(
  top_fungal_plot_data,
  aes(
    x = group,
    y = raw_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title = "Top fungal age-sensitive LC-MS features",
    subtitle =
      "Positive-mode LC-MS; points represent biological replicates",
    x = NULL,
    y = "Raw LC-MS feature intensity"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      ),
    
    strip.text =
      element_text(
        face = "bold"
      )
  )
```

106 make the log2 version Which metabolites/features change as the fungus ages?
```{r}
ggplot(
  top_fungal_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title =
      "Top fungal age-sensitive LC-MS features",
    
    subtitle =
      "Control → outside colony → under colony → older fungal growth",
    
    x = NULL,
    y = "log2(Intensity + 1)"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      ),
    
    strip.text =
      element_text(
        face = "bold"
      )
  )
```

107 make a summary table
```{r}
fungal_candidate_summary <-
  top_fungal_plot_data %>%
  
  group_by(
    feature_name,
    group
  ) %>%
  
  summarise(
    
    detected_n =
      sum(
        raw_intensity > 0,
        na.rm = TRUE
      ),
    
    mean_intensity =
      mean(
        raw_intensity,
        na.rm = TRUE
      ),
    
    median_intensity =
      median(
        raw_intensity,
        na.rm = TRUE
      ),
    
    sd_intensity =
      sd(
        raw_intensity,
        na.rm = TRUE
      ),
    
    .groups = "drop"
  )

fungal_candidate_summary
```

108 and we should do the 14 age-stable features separately
```{r}
age_stable_feature_names <-
  age_stable_fungal_features %>%
  pull(feature_name)

length(age_stable_feature_names)
```
then
```{r}
age_stable_plot_data <-
  fungal_long %>%
  filter(
    feature_name %in%
      age_stable_feature_names
  )
```

plot Which metabolites/features appear to be consistently associated with fungal growth regardless of age?
```{r}
ggplot(
  age_stable_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +
  
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +
  
  geom_jitter(
    width = 0.08,
    size = 1.0
  ) +
  
  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +
  
  labs(
    title =
      "Age-stable fungal-associated LC-MS features",
    
    subtitle =
      "Features associated with fungal growth regardless of colony age",
    
    x = NULL,
    y = "log2(Intensity + 1)"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "none",
    
    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      ),
    
    strip.text =
      element_text(
        face = "bold"
      )
  )
```

109 define all biological compartments
```{r}
# ------------------------------------------------------------
# CROSS-COMPARTMENT FEATURE OVERLAP
# Plant tissues + exudate + fungal samples
# ------------------------------------------------------------

cross_compartment_groups <- list(

  Leaf =
    presence_groups$Leaf,

  Root =
    presence_groups$Root,

  Exudate =
    presence_groups$Exudate,

  Soil =
    presence_groups$Soil,

  Fungal_control =
    presence_groups$Fungal_control,

  Fungal_out =
    presence_groups$Fungal_out,

  Fungal_in =
    presence_groups$Fungal_in,

  Fungi_old =
    presence_groups$Fungi_old
)
```

110 count detection in every compartment
```{r}
cross_compartment_presence <- feature_qc %>%

  mutate(

    leaf_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Leaf),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    root_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Root),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    exudate_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Exudate),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    soil_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Soil),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    fungal_control_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Fungal_control),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    fungal_out_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Fungal_out),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    fungal_in_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Fungal_in),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    fungi_old_n =
      rowSums(
        across(
          all_of(cross_compartment_groups$Fungi_old),
          ~ .x > 0
        ),
        na.rm = TRUE
      )
  )
```


111 convert those counts into presence/absence
```{r}
cross_compartment_presence <-
  cross_compartment_presence %>%

  mutate(

    leaf_present =
      leaf_n >= 2,

    root_present =
      root_n >= 2,

    exudate_present =
      exudate_n >= 2,

    soil_present =
      soil_n >= 2,

    fungal_control_present =
      fungal_control_n >= 2,

    fungal_out_present =
      fungal_out_n >= 2,

    fungal_in_present =
      fungal_in_n >= 2,

    fungi_old_present =
      fungi_old_n >= 2
  )
```


112 how many features overlap plant and fungal samples. how many features are consistenly detected somewhere in the plant system and somewhere in fungus-associated samples?

```{r}
cross_compartment_presence <-
  cross_compartment_presence %>%

  mutate(

    plant_present =
      leaf_present |
      root_present |
      exudate_present,

    fungus_present =
      fungal_in_present |
      fungal_out_present |
      fungi_old_present,

    plant_and_fungus =
      plant_present &
      fungus_present
  )
```


count them
```{r}
cross_compartment_presence %>%
  summarise(

    total_features =
      n(),

    plant_features =
      sum(
        plant_present
      ),

    fungal_features =
      sum(
        fungus_present
      ),

    plant_and_fungal_features =
      sum(
        plant_and_fungus
      )
  )
```


113 look specifically for root + exudate + fungus
```{r}
root_exudate_fungal_features <-
  cross_compartment_presence %>%

  filter(

    root_present,

    exudate_present,

    fungal_in_present |
      fungi_old_present
  )
```

count
```{r}
nrow(
  root_exudate_fungal_features
)
```


114 exclude things also consistently present in controls
```{r}
root_exudate_fungal_not_controls <-
  cross_compartment_presence %>%

  filter(

    root_present,

    exudate_present,

    fungal_in_present |
      fungi_old_present,

    !soil_present,

    !fungal_control_present
  )
```


count
```{r}
nrow(
  root_exudate_fungal_not_controls
)
```
115 look for exudate +fungus but not root

```{r}
exudate_fungal_not_root <-
  cross_compartment_presence %>%

  filter(

    exudate_present,

    fungal_in_present |
      fungi_old_present,

    !root_present,

    !soil_present,

    !fungal_control_present
  )
```

count
```{r}
nrow(
  exudate_fungal_not_root
)
```
116 look for root + fungus but not exudate
```{r}
root_fungal_not_exudate <-
  cross_compartment_presence %>%

  filter(

    root_present,

    fungal_in_present |
      fungi_old_present,

    !exudate_present
  )
```

count
```{r}
nrow(
  root_fungal_not_exudate
)
```

117 create a biologicallt useful classification
```{r}
cross_compartment_presence <-
  cross_compartment_presence %>%

  mutate(

    cross_pattern = case_when(

      root_present &
        exudate_present &
        fungal_in_present &
        !soil_present &
        !fungal_control_present ~
        "Root + Exudate + Fungal in",

      root_present &
        exudate_present &
        fungi_old_present &
        !soil_present &
        !fungal_control_present ~
        "Root + Exudate + Fungi old",

      root_present &
        exudate_present &
        fungal_in_present &
        fungi_old_present &
        !soil_present &
        !fungal_control_present ~
        "Root + Exudate + New and Old fungus",

      exudate_present &
        fungal_in_present &
        !root_present &
        !soil_present &
        !fungal_control_present ~
        "Exudate + Fungal in",

      root_present &
        fungal_in_present &
        !exudate_present ~
        "Root + Fungal in",

      leaf_present &
        root_present &
        exudate_present &
        fungus_present ~
        "Leaf + Root + Exudate + Fungus",

      TRUE ~
        "Other pattern"
    )
  )
```



alternative: chooses the first condition that matches. the most specific category should come first.
```{r}
cross_compartment_presence <-
  cross_compartment_presence %>%

  mutate(

    cross_pattern = case_when(

      root_present &
        exudate_present &
        fungal_in_present &
        fungi_old_present &
        !soil_present &
        !fungal_control_present ~
        "Root + Exudate + New and Old fungus",

      root_present &
        exudate_present &
        fungal_in_present &
        !soil_present &
        !fungal_control_present ~
        "Root + Exudate + Fungal in",

      root_present &
        exudate_present &
        fungi_old_present &
        !soil_present &
        !fungal_control_present ~
        "Root + Exudate + Fungi old",

      exudate_present &
        fungal_in_present &
        !root_present &
        !soil_present &
        !fungal_control_present ~
        "Exudate + Fungal in",

      root_present &
        fungal_in_present &
        !exudate_present ~
        "Root + Fungal in",

      leaf_present &
        root_present &
        exudate_present &
        fungus_present ~
        "Leaf + Root + Exudate + Fungus",

      TRUE ~
        "Other pattern"
    )
  )
```


118 count those biological patterns

```{r}
cross_pattern_summary <-
  cross_compartment_presence %>%

  count(
    cross_pattern,
    sort = TRUE
  ) %>%

  mutate(
    percent =
      100 * n / sum(n)
  )

cross_pattern_summary
```

119 check whether our five root-exudate candidates occur in fungal samples

```{r}
five_cross_compartment <-
  cross_compartment_presence %>%

  filter(
    feature_name %in%
      five_candidate_features
  ) %>%

  select(

    feature_name,

    `Average Rt(min)`,

    `Average Mz`,

    `MS/MS assigned`,

    leaf_n,
    root_n,
    exudate_n,
    soil_n,

    fungal_control_n,
    fungal_out_n,
    fungal_in_n,
    fungi_old_n,

    leaf_present,
    root_present,
    exudate_present,
    soil_present,

    fungal_control_present,
    fungal_out_present,
    fungal_in_present,
    fungi_old_present
  )

five_cross_compartment
```
120 save the cross-compartment results
```{r}
write.csv(
  cross_compartment_presence,
  "positive_mode_cross_compartment_presence.csv",
  row.names = FALSE
)

write.csv(
  cross_pattern_summary,
  "positive_mode_cross_compartment_pattern_summary.csv",
  row.names = FALSE
)

write.csv(
  root_exudate_fungal_not_controls,
  "positive_mode_root_exudate_fungal_candidates.csv",
  row.names = FALSE
)

write.csv(
  five_cross_compartment,
  "positive_mode_five_candidates_cross_compartment.csv",
  row.names = FALSE
)
```

inspect the 4 features Notice that only one of your original five candidates appears to meet this stricter fungal-control criterion: F5091. The other three candidates among the four must therefore be features elsewhere in the dataset that weren't part of your original five-feature root/exudate shortlist.

I would inspect those four next

```{r}
root_exudate_fungal_not_controls %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio,

    leaf_n,
    root_n,
    exudate_n,
    soil_n,

    fungal_control_n,
    fungal_out_n,
    fungal_in_n,
    fungi_old_n
  )
```
make boxplotsa

```{r}
cross_candidate_names <-
  root_exudate_fungal_not_controls %>%
  pull(feature_name)
```

creat long daya
```{r}
cross_candidate_samples <- c(
  presence_groups$Leaf,
  presence_groups$Root,
  presence_groups$Exudate,
  presence_groups$Soil,
  presence_groups$Fungal_control,
  presence_groups$Fungal_out,
  presence_groups$Fungal_in,
  presence_groups$Fungi_old
)

cross_candidate_plot_data <- feature_qc %>%

  filter(
    feature_name %in%
      cross_candidate_names
  ) %>%

  select(
    feature_name,
    all_of(
      cross_candidate_samples
    )
  ) %>%

  pivot_longer(
    cols =
      all_of(
        cross_candidate_samples
      ),
    names_to =
      "sample",
    values_to =
      "raw_intensity"
  ) %>%

  left_join(
    sample_metadata %>%
      select(
        sample,
        group
      ),
    by = "sample"
  ) %>%

  mutate(
    group = factor(
      group,
      levels = c(
        "Leaf",
        "Root",
        "Plant-soil exudate",
        "Soil control",
        "Fungal control",
        "Fungal agar out",
        "Fungal agar in",
        "Fungi old"
      )
    ),

    log2_intensity =
      log2(
        raw_intensity + 1
      )
  )
```

plot
```{r}
ggplot(
  cross_candidate_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +

  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +

  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +

  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +

  labs(
    title =
      "Cross-system root–exudate–fungal candidate features",

    subtitle =
      "Positive-mode LC-MS; individual points are biological replicates",

    x = NULL,

    y =
      "log2(Intensity + 1)"
  ) +

  theme_classic() +

  theme(
    legend.position =
      "none",

    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      ),

    strip.text =
      element_text(
        face = "bold"
      )
  )
```

#simplified statistics

121. Master feature-count table by biological compartment

This tells you how many of the 4,144 filtered features are reproducibly detected in each individual sample type.

# ------------------------------------------------------------
# PI PRESENTATION: RESULTS SUMMARY
# ------------------------------------------------------------

```{r}
compartment_feature_counts <- tibble(

  compartment = c(
    "Leaf",
    "Root",
    "Plant-soil exudate",
    "Soil control",
    "Fungal control",
    "Fungal agar out",
    "Fungal agar in",
    "Fungi old"
  ),

  detected_features = c(

    sum(
      cross_compartment_presence$leaf_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$root_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$exudate_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$soil_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$fungal_control_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$fungal_out_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$fungal_in_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$fungi_old_present,
      na.rm = TRUE
    )
  )
)

compartment_feature_counts
```




122. Plot number of reproducibly detected features in every compartment
```{r}
compartment_feature_plot <- ggplot(
  compartment_feature_counts,
  aes(
    x = reorder(
      compartment,
      detected_features
    ),
    y = detected_features,
    fill = compartment
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(
      label = detected_features
    ),
    hjust = -0.1,
    size = 4
  ) +

  labs(
    title =
      "Reproducibly detected positive-mode LC-MS features",
    subtitle =
      "Presence defined as detection in at least 2 of 3 replicates",
    x = NULL,
    y = "Number of filtered LC-MS features"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )

compartment_feature_plot
```



Save it:
```{r}
ggsave(
  "PI_figure_feature_counts_by_compartment.png",
  compartment_feature_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```


123. Create the broad plant–fungus overlap summary

These are the numbers you just obtained.

```{r}
plant_fungal_overlap_summary <- tibble(

  category = c(
    "All filtered features",
    "Plant-associated",
    "Fungus-associated",
    "Plant + fungus overlap",
    "Strict root-exudate-fungus candidates"
  ),

  n = c(

    nrow(
      cross_compartment_presence
    ),

    sum(
      cross_compartment_presence$plant_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$fungus_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$plant_and_fungus,
      na.rm = TRUE
    ),

    nrow(
      root_exudate_fungal_not_controls
    )
  )
)

plant_fungal_overlap_summary

```



124. Plot the biological filtering funnel

```{r}
candidate_funnel_plot <- ggplot(
  plant_fungal_overlap_summary,
  aes(
    x = factor(
      category,
      levels = rev(category)
    ),
    y = n,
    fill = category
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(label = n),
    hjust = -0.1,
    size = 4
  ) +

  labs(
    title =
      "Filtering positive-mode LC-MS features toward cross-system candidates",
    subtitle =
      "The four candidates represent a stringent biological pattern, not the total plant-fungal overlap",
    x = NULL,
    y = "Number of LC-MS features"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )

candidate_funnel_plot
```



125. Calculate plant-only, fungus-only, and shared feature numbers

```{r}
plant_fungal_partition <- cross_compartment_presence %>%
  summarise(

    plant_only =
      sum(
        plant_present &
          !fungus_present,
        na.rm = TRUE
      ),

    fungus_only =
      sum(
        fungus_present &
          !plant_present,
        na.rm = TRUE
      ),

    plant_and_fungus =
      sum(
        plant_present &
          fungus_present,
        na.rm = TRUE
      ),

    neither =
      sum(
        !plant_present &
          !fungus_present,
        na.rm = TRUE
      )
  )

plant_fungal_partition
```



126. Plot the plant–fungus partition

First reshape:

```{r}
plant_fungal_partition_long <-
  plant_fungal_partition %>%
  pivot_longer(
    everything(),
    names_to = "category",
    values_to = "features"
  ) %>%
  mutate(

    category = recode(
      category,

      plant_only =
        "Plant only",

      fungus_only =
        "Fungus only",

      plant_and_fungus =
        "Plant + fungus",

      neither =
        "Neither consistently detected"
    )
  )

```


Plot:

```{r}
plant_fungal_partition_plot <- ggplot(
  plant_fungal_partition_long,
  aes(
    x = category,
    y = features,
    fill = category
  )
) +

  geom_col() +

  geom_text(
    aes(label = features),
    vjust = -0.3,
    size = 4
  ) +

  labs(
    title =
      "Distribution of filtered features across plant and fungal systems",
    x = NULL,
    y = "Number of features"
  ) +

  theme_classic() +

  theme(
    legend.position = "none",
    axis.text.x =
      element_text(
        angle = 30,
        hjust = 1
      )
  )

plant_fungal_partition_plot
```


127. Pairwise overlap table across all eight biological compartments


How many features are shared between root and exudate?
Root and fungal-in?
Exudate and fungal-in?
Fungal-in and old fungus?

Create a logical matrix:

```{r}
presence_matrix <- cross_compartment_presence %>%
  select(
    Leaf = leaf_present,
    Root = root_present,
    Exudate = exudate_present,
    Soil = soil_present,
    Fungal_control = fungal_control_present,
    Fungal_out = fungal_out_present,
    Fungal_in = fungal_in_present,
    Fungi_old = fungi_old_present
  )

```


Then:

```{r}
presence_matrix_numeric <-
  as.matrix(
    presence_matrix
  ) * 1

```


Calculate pairwise overlap:

```{r}
pairwise_overlap_matrix <-
  t(presence_matrix_numeric) %*%
  presence_matrix_numeric

pairwise_overlap_matrix
```



This is a very important table.

The diagonal tells you:

Leaf vs Leaf = total leaf features
Root vs Root = total root features
...

And an off-diagonal cell such as:

Root × Exudate

tells you exactly how many filtered features are reproducibly present in both.

128. Make the overlap matrix into a heatmap

```{r}
pheatmap(
  pairwise_overlap_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.0f",
  main =
    "Shared LC-MS features among plant, soil and fungal compartments"
)

```



Which sample types have the most chemistry in common?

129. Calculate percent overlap

Raw counts are useful, but percent overlap can also help.

For each pair, use the smaller feature set as the denominator:

```{r}
pairwise_percent_overlap <-
  matrix(
    NA_real_,
    nrow = ncol(presence_matrix_numeric),
    ncol = ncol(presence_matrix_numeric)
  )

rownames(pairwise_percent_overlap) <-
  colnames(presence_matrix_numeric)

colnames(pairwise_percent_overlap) <-
  colnames(presence_matrix_numeric)
```



Calculate:

```{r}
for (
  i in seq_len(
    ncol(presence_matrix_numeric)
  )
) {

  for (
    j in seq_len(
      ncol(presence_matrix_numeric)
    )
  ) {

    overlap <-
      sum(
        presence_matrix_numeric[, i] == 1 &
          presence_matrix_numeric[, j] == 1
      )

    smaller_total <-
      min(
        sum(
          presence_matrix_numeric[, i]
        ),
        sum(
          presence_matrix_numeric[, j]
        )
      )

    pairwise_percent_overlap[i, j] <-
      100 *
      overlap /
      smaller_total
  }
}

```


View:

```{r}
round(
  pairwise_percent_overlap,
  1
)
```



Heatmap:

```{r}
pheatmap(
  pairwise_percent_overlap,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.1f",
  main =
    "Percent feature overlap among biological compartments"
)
```


130. Plant-specific feature-pattern table

Now summarize just:

Leaf
Root
Exudate
Soil

```{r}
plant_pattern_detailed <- cross_compartment_presence %>%
  mutate(

    plant_detection_pattern =
      paste0(
        ifelse(
          leaf_present,
          "Leaf ",
          ""
        ),

        ifelse(
          root_present,
          "Root ",
          ""
        ),

        ifelse(
          exudate_present,
          "Exudate ",
          ""
        ),

        ifelse(
          soil_present,
          "Soil",
          ""
        )
      )
  ) %>%

  count(
    plant_detection_pattern,
    sort = TRUE
  )

plant_pattern_detailed
```



This tells you how many features are:

Leaf only
Root only
Leaf + root
Root + exudate
Leaf + root + exudate
Exudate + soil
all four
etc.

131. Fungal-specific feature-pattern table

Do the same for:

Fungal control
Fungal out
Fungal in
Fungi old

```{r}
fungal_pattern_detailed <- cross_compartment_presence %>%
  mutate(

    fungal_detection_pattern =
      paste0(

        ifelse(
          fungal_control_present,
          "Control ",
          ""
        ),

        ifelse(
          fungal_out_present,
          "Out ",
          ""
        ),

        ifelse(
          fungal_in_present,
          "In ",
          ""
        ),

        ifelse(
          fungi_old_present,
          "Old",
          ""
        )
      )
  ) %>%

  count(
    fungal_detection_pattern,
    sort = TRUE
  )

fungal_pattern_detailed
```



This is extremely useful for explaining fungal chemistry.

You may find categories like:


132. Plot the top fungal detection patterns

```{r}
fungal_pattern_plot_data <-
  fungal_pattern_detailed %>%
  filter(
    fungal_detection_pattern != ""
  ) %>%
  slice_head(
    n = 15
  )
```



Plot:

```{r}
fungal_pattern_plot <- ggplot(
  fungal_pattern_plot_data,
  aes(
    x = reorder(
      fungal_detection_pattern,
      n
    ),
    y = n
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(label = n),
    hjust = -0.1
  ) +

  labs(
    title =
      "Most common fungal LC-MS detection patterns",
    x = NULL,
    y = "Number of features"
  ) +

  theme_classic()

fungal_pattern_plot
```


133. Plot the top plant detection patterns

```{r}
plant_pattern_plot_data <-
  plant_pattern_detailed %>%
  filter(
    plant_detection_pattern != ""
  ) %>%
  slice_head(
    n = 15
  )
plant_pattern_plot <- ggplot(
  plant_pattern_plot_data,
  aes(
    x = reorder(
      plant_detection_pattern,
      n
    ),
    y = n
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(label = n),
    hjust = -0.1
  ) +

  labs(
    title =
      "Most common plant/soil LC-MS detection patterns",
    x = NULL,
    y = "Number of features"
  ) +

  theme_classic()

plant_pattern_plot
```


134. Strict cross-system candidate table

These are the four, and this is where you explain exactly what “four” means.

```{r}
strict_cross_system_candidates <-
  root_exudate_fungal_not_controls %>%

  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio,

    leaf_n,
    root_n,
    exudate_n,
    soil_n,

    fungal_control_n,
    fungal_out_n,
    fungal_in_n,
    fungi_old_n
  )

strict_cross_system_candidates
```



Add a descriptive column:

```{r}
strict_cross_system_candidates <-
  strict_cross_system_candidates %>%
  mutate(

    interpretation =
      "Root + exudate + fungal-associated; not consistently detected in soil or fungal control"
  )
```


135. Cross-system candidate boxplots

You've already generated this, but save a presentation-quality copy:

```{r}
cross_system_candidate_plot <- ggplot(
  cross_candidate_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +

  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +

  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +

  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +

  labs(
    title =
      "Cross-system root–exudate–fungal candidate features",

    subtitle =
      "Positive-mode LC-MS; individual points are biological replicates",

    x = NULL,

    y =
      "log2(Intensity + 1)"
  ) +

  theme_classic() +

  theme(
    legend.position = "none",

    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      ),

    strip.text =
      element_text(
        face = "bold"
      )
  )

cross_system_candidate_plot
```


Save:



```{r}
ggsave(
  "PI_figure_cross_system_candidates.png",
  cross_system_candidate_plot,
  width = 12,
  height = 8,
  dpi = 300
)
```


136. Summarize exudate-vs-soil detection evidence

You already obtained:

129 exudate-associated
124 soil-associated
1,386 quantifiable in both
2,505 not consistently detected in either.

Let's turn that into a PI figure.

```{r}
exudate_soil_detection_PI <-
  prepared_comparisons$
    Exudate_vs_Soil$
    detection_summary

exudate_soil_detection_PI
```



Plot:

```{r}
exudate_soil_detection_plot <- ggplot(
  exudate_soil_detection_PI,
  aes(
    x = reorder(
      detection_category,
      n
    ),
    y = n,
    fill = detection_category
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(
      label = paste0(
        n,
        " (",
        round(
          percent,
          1
        ),
        "%)"
      )
    ),
    hjust = -0.05
  ) +

  labs(
    title =
      "Exudate versus soil: feature detection categories",
    subtitle =
      "Presence requires detection in at least 2 of 3 replicates",
    x = NULL,
    y = "Number of features"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )

exudate_soil_detection_plot
```


137. Exudate-vs-soil quantitative evidence tiers

You already obtained:

Tier 1    1
Tier 3    159
Tier 4    215
Tier 5   1011

Generate the table:

```{r}
exudate_evidence_summary <-
  exudate_soil_abundance_candidates %>%
  count(
    evidence_tier
  )

exudate_evidence_summary
```



Plot:

```{r}
exudate_evidence_plot <- ggplot(
  exudate_evidence_summary,
  aes(
    x = reorder(
      evidence_tier,
      n
    ),
    y = n,
    fill = evidence_tier
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(label = n),
    hjust = -0.1
  ) +

  labs(
    title =
      "Strength of quantitative evidence: exudate vs soil",
    subtitle =
      "1,386 features were sufficiently detected in both groups for abundance testing",
    x = NULL,
    y = "Number of features"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )

exudate_evidence_plot
```


138. Fungal-age summary figure

Your current numbers are:

No strong classification       1697
New-vs-control only             649
Old-vs-new differential         262
Age-stable fungal effect         14
Old-vs-control only               6

Use your actual object:

```{r}
fungal_age_summary
```



Plot:

```{r}
fungal_age_summary_plot <- ggplot(
  fungal_age_summary,
  aes(
    x = reorder(
      fungal_age_category,
      n
    ),
    y = n,
    fill = fungal_age_category
  )
) +

  geom_col() +

  coord_flip() +

  geom_text(
    aes(label = n),
    hjust = -0.1
  ) +

  labs(
    title =
      "Fungal age and growth-associated LC-MS patterns",
    subtitle =
      "Positive-mode LC-MS",
    x = NULL,
    y = "Number of features"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )

fungal_age_summary_plot
```


139. Fungal-age heatmap

You already generated this. Save it explicitly when you rerun the heatmap:

```{r}
png(
  "PI_figure_fungal_age_heatmap.png",
  width = 2200,
  height = 1600,
  res = 200
)

pheatmap(
  fungal_age_heatmap,
  scale = "row",
  clustering_distance_rows =
    "euclidean",
  clustering_distance_cols =
    "euclidean",
  clustering_method =
    "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  main =
    "Features differing between old and newer fungal growth"
)

dev.off()
```


140. Fungal top-feature boxplots

Your top fungal-age plot is useful because the heatmap only shows relative differences.

Save the boxplot:

```{r}
fungal_age_boxplot <- ggplot(
  top_fungal_plot_data,
  aes(
    x = group,
    y = log2_intensity,
    fill = group
  )
) +

  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.45
  ) +

  geom_jitter(
    width = 0.08,
    size = 2.5
  ) +

  facet_wrap(
    ~ feature_name,
    scales = "free_y",
    ncol = 2
  ) +

  labs(
    title =
      "Top fungal age-sensitive LC-MS features",

    subtitle =
      "Control → outside colony → under colony → older fungal growth",

    x = NULL,

    y =
      "log2(Intensity + 1)"
  ) +

  theme_classic() +

  theme(
    legend.position = "none",

    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      )
  )

fungal_age_boxplot
```


141. PCA plots to save for Laura

You already have these objects, so don't rerun the entire PCA.

Save the broad PCA:

```{r}
ggsave(
  "PI_figure_global_PCA.png",
  broad_pca_plot,
  width = 9,
  height = 7,
  dpi = 300
)

```


Exudate vs soil:

```{r}
ggsave(
  "PI_figure_exudate_soil_PCA.png",
  quantitative_pca_results$
    Exudate_vs_Soil$plot,
  width = 8,
  height = 6,
  dpi = 300
)
```



Leaf vs root:

```{r}
ggsave(
  "PI_figure_leaf_root_PCA.png",
  quantitative_pca_results$
    Leaf_vs_Root$plot,
  width = 8,
  height = 6,
  dpi = 300
)
```



Fungal in vs control:

```{r}
ggsave(
  "PI_figure_fungal_in_control_PCA.png",
  quantitative_pca_results$
    Fungal_in_vs_control$plot,
  width = 8,
  height = 6,
  dpi = 300
)
```



Old vs new fungus:

```{r}
ggsave(
  "PI_figure_fungal_old_new_PCA.png",
  quantitative_pca_results$
    Fungi_old_vs_Fungal_in$plot,
  width = 8,
  height = 6,
  dpi = 300
)
```


142. Create one master PI summary table

```{r}
PI_summary_table <- tibble(

  result = c(

    "Raw MS-DIAL features",

    "Features after positive-mode filtering",

    "Plant-associated features",

    "Fungus-associated features",

    "Features shared between plant and fungus",

    "Strict root-exudate-fungus candidates",

    "Exudate-associated detection features",

    "Exudate-soil features quantifiable in both",

    "Exudate-soil Tier 1 abundance candidates",

    "Old-vs-new fungal differential features",

    "Age-stable fungal-associated features"
  ),

  n = c(

    nrow(feature_table),

    nrow(feature_qc),

    sum(
      cross_compartment_presence$
        plant_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$
        fungus_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$
        plant_and_fungus,
      na.rm = TRUE
    ),

    nrow(
      root_exudate_fungal_not_controls
    ),

    129,

    1386,

    sum(
      exudate_soil_abundance_candidates$
        evidence_tier ==
        "Tier 1: FDR + >=2-fold"
    ),

    sum(
      fungal_age_master$
        fungal_age_category ==
        "Age-sensitive fungal effect",
      na.rm = TRUE
    ),

    sum(
      fungal_age_master$
        fungal_age_category ==
        "Age-stable fungal effect",
      na.rm = TRUE
    )
  )
)

PI_summary_table
```



I would not hard-code 129 and 1386 long term, though. For a reproducible final notebook, replace them with calculations:

```{r}
PI_summary_table <- tibble(

  result = c(
    "Raw MS-DIAL features",
    "Features after positive-mode filtering",
    "Plant-associated features",
    "Fungus-associated features",
    "Plant + fungus shared features",
    "Strict root-exudate-fungus candidates",
    "Exudate-associated detection features",
    "Exudate-soil quantifiable features",
    "Exudate-soil Tier 1 candidates",
    "Old-vs-new fungal differential features",
    "Age-stable fungal effects"
  ),

  n = c(

    nrow(feature_table),

    nrow(feature_qc),

    sum(
      cross_compartment_presence$plant_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$fungus_present,
      na.rm = TRUE
    ),

    sum(
      cross_compartment_presence$plant_and_fungus,
      na.rm = TRUE
    ),

    nrow(
      root_exudate_fungal_not_controls
    ),

    sum(
      prepared_comparisons$
        Exudate_vs_Soil$
        detection_table$
        detection_category ==
        "Exudate-associated detection",
      na.rm = TRUE
    ),

    nrow(
      prepared_comparisons$
        Exudate_vs_Soil$
        abundance_features
    ),

    sum(
      exudate_soil_abundance_candidates$
        evidence_tier ==
        "Tier 1: FDR + >=2-fold",
      na.rm = TRUE
    ),

    sum(
      fungal_age_master$
        fungal_age_category ==
        "Age-sensitive fungal effect",
      na.rm = TRUE
    ),

    sum(
      fungal_age_master$
        fungal_age_category ==
        "Age-stable fungal effect",
      na.rm = TRUE
    )
  )
)

PI_summary_table
```



Save:

```{r}
write.csv(
  PI_summary_table,
  "PI_positive_mode_results_summary.csv",
  row.names = FALSE
)
```


143. Save all the new PI tables

```{r}
write.csv(
  compartment_feature_counts,
  "PI_feature_counts_by_compartment.csv",
  row.names = FALSE
)

write.csv(
  plant_fungal_overlap_summary,
  "PI_plant_fungal_overlap_summary.csv",
  row.names = FALSE
)

write.csv(
  plant_fungal_partition_long,
  "PI_plant_fungal_partition.csv",
  row.names = FALSE
)

write.csv(
  pairwise_overlap_matrix,
  "PI_pairwise_feature_overlap_counts.csv"
)

write.csv(
  pairwise_percent_overlap,
  "PI_pairwise_percent_overlap.csv"
)

write.csv(
  plant_pattern_detailed,
  "PI_plant_detection_patterns.csv",
  row.names = FALSE
)

write.csv(
  fungal_pattern_detailed,
  "PI_fungal_detection_patterns.csv",
  row.names = FALSE
)

write.csv(
  strict_cross_system_candidates,
  "PI_strict_cross_system_candidates.csv",
  row.names = FALSE
)
```

