---
title: "Pilot Study"
output: html_document
date: "2026-07-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

setwd("C:/Users/norba/Downloads/GCMS_R_Project_Report/GCMS_R_Project")
getwd()

library(tidyverse)
library(readr)
library(ggplot2)

gcms <- read_csv(
  "C:/Users/norba/Downloads/GCMS_R_Project_Report/GCMS_R_Project/data/gcms_tidy_r_input.csv",
  show_col_types = FALSE
)
```

#create folders in the project
```{r}
dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
```


#summarize data and place in the folders that were created above
```{r}
# Basic summaries
write_csv(gcms %>% count(tissue, sample_id), "tables/sample_counts.csv")
write_csv(gcms %>% distinct(tissue, compound_short) %>% count(tissue), "tables/unique_compounds_by_tissue.csv")
write_csv(gcms %>% distinct(tissue, compound_short, chemical_class) %>% count(tissue, chemical_class), "tables/chemical_class_summary.csv")
```


```{r}
# Chemical class composition
class_summary <- gcms %>% distinct(tissue, compound_short, chemical_class) %>% count(tissue, chemical_class)

p <- ggplot(class_summary, aes(x = tissue, y = n, fill = chemical_class)) +
  geom_col(color = "black") +
  theme_bw(base_size = 16) +
  labs(title = "Chemical class composition by tissue",
       x = "Tissue", y = "Number of unique compounds", fill = "Chemical class")
p

#saves in figures folder
ggsave("figures/chemical_class_composition.png", p, width = 8, height = 6, dpi = 600)
```


```{r}
# Presence/absence heatmap
presence <- gcms %>% distinct(tissue, compound_short, chemical_class)

pp <- ggplot(presence, aes(x = tissue, y = reorder(compound_short, chemical_class), fill = chemical_class)) +
  geom_tile(color = "white") +
  theme_bw(base_size = 14) +
  labs(title = "Presence of putative compounds in leaf and root tissue",
       x = "Tissue", y = "Compound", fill = "Chemical class")
pp

#saves in figures folder
ggsave("figures/presence_absence_heatmap.png", p, width = 7, height = 10, dpi = 600)

```


```{r}
# ============================================================
# GC-MS Leaf vs Root Analysis Figures
# Input file: data/gcms_tidy_r_input.csv
# ============================================================

# Load data
gcms <- read_csv("C:/Users/norba/Downloads/GCMS_R_Project_Report/GCMS_R_Project/data/gcms_tidy_r_input.csv", show_col_types = FALSE)

# Check data
glimpse(gcms)

# Make tissue and chemical class factors
gcms <- gcms %>%
  mutate(
    tissue = factor(tissue, levels = c("Leaf", "Root")),
    chemical_class = factor(
      chemical_class,
      levels = c(
        "Monoterpene",
        "Sesquiterpene",
        "Putative diterpenoid",
        "Hydrocarbon",
        "Fatty alcohol",
        "Fatty acid",
        "Unknown"
      )
    )
  )
```


```{r}
# ============================================================
# SUMMARY TABLES
# ============================================================

dataset_summary <- tibble(
  metric = c(
    "Total unique GC-MS observations",
    "Unique compounds",
    "Unique samples",
    "Leaf samples",
    "Root samples"
  ),
  value = c(
    nrow(gcms),
    n_distinct(gcms$compound_short),
    n_distinct(gcms$sample_id),
    n_distinct(gcms$sample_id[gcms$tissue == "Leaf"]),
    n_distinct(gcms$sample_id[gcms$tissue == "Root"])
  )
)

write_csv(dataset_summary, "tables/dataset_summary.csv")

sample_counts <- gcms %>%
  count(tissue, sample_id, name = "unique_observations")

write_csv(sample_counts, "tables/sample_counts.csv")

chemical_class_summary <- gcms %>%
  distinct(tissue, compound_short, chemical_class) %>%
  count(tissue, chemical_class, name = "unique_compounds")

write_csv(chemical_class_summary, "tables/chemical_class_summary.csv")

compound_detection_counts <- gcms %>%
  count(compound_short, chemical_class, name = "detections") %>%
  arrange(desc(detections), compound_short)

write_csv(compound_detection_counts, "tables/compound_detection_counts.csv")
```


```{r}
# ============================================================
# FIGURE 1: UNIQUE OBSERVATIONS PER SAMPLE
# ============================================================

p1 <- ggplot(sample_counts, aes(x = sample_id, y = unique_observations, fill = tissue)) +
  geom_col(color = "black", width = 0.7) +
  labs(
    title = "Unique GC-MS Observations per Sample",
    x = "Sample",
    y = "Number of Unique Observations",
    fill = "Tissue"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 15, face = "bold"),
    axis.text.y = element_text(size = 15),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14)
  )

p1
ggsave("figures/Figure1_unique_observations_per_sample.png", p1, width = 8, height = 5, dpi = 600)
```

```{r}

# ============================================================
# FIGURE 2: UNIQUE COMPOUNDS BY TISSUE
# ============================================================

gcms <- gcms %>%
  mutate(
    tissue = case_when(
      str_detect(sample_id, "Leaf") ~ "Leaf",
      str_detect(sample_id, "Root") ~ "Root",
      TRUE ~ tissue
    )
  )

unique(gcms$tissue)

compound_counts_tissue <- gcms %>%
  distinct(tissue, compound_short) %>%
  count(tissue, name = "unique_compounds")

p2 <- ggplot(compound_counts_tissue, aes(x = tissue, y = unique_compounds, fill = tissue)) +
  geom_col(color = "black", width = 0.6) +
  labs(
    title = "Unique Putative Compounds Detected by Tissue",
    x = "Tissue",
    y = "Number of Unique Compounds"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )

p2
ggsave("figures/Figure2_unique_compounds_by_tissue.png", p2, width = 7, height = 5, dpi = 600)

```


```{r}
# ============================================================
# FIGURE 3: CHEMICAL CLASS COMPOSITION BY TISSUE
# ============================================================

p3 <- ggplot(chemical_class_summary, aes(x = tissue, y = unique_compounds, fill = chemical_class)) +
  geom_col(color = "black", width = 0.7) +
  labs(
    title = "Chemical Class Composition by Tissue",
    x = "Tissue",
    y = "Number of Unique Compounds",
    fill = "Chemical Class"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 13)
  )

p3
ggsave("figures/Figure3_chemical_class_composition.png", p3, width = 8, height = 6, dpi = 600)
```



```{r}
# ============================================================
# FIGURE 4: CHEMICAL CLASS PERCENTAGES
# ============================================================

chemical_class_percent <- chemical_class_summary %>%
  group_by(tissue) %>%
  mutate(percent = 100 * unique_compounds / sum(unique_compounds)) %>%
  ungroup()

write_csv(chemical_class_percent, "tables/chemical_class_percentages.csv")

p4 <- ggplot(chemical_class_percent, aes(x = tissue, y = percent, fill = chemical_class)) +
  geom_col(color = "black", width = 0.7) +
  labs(
    title = "Chemical Class Percentages by Tissue",
    x = "Tissue",
    y = "Percent of Unique Compounds",
    fill = "Chemical Class"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 13)
  )

p4
ggsave("figures/Figure4_chemical_class_percentages.png", p4, width = 8, height = 6, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 5: PRESENCE/ABSENCE HEATMAP BY TISSUE
# ============================================================

presence_tissue <- gcms %>%
  distinct(tissue, compound_short, chemical_class) %>%
  arrange(chemical_class, compound_short) %>%
  mutate(compound_short = factor(compound_short, levels = unique(compound_short)))

p5 <- ggplot(presence_tissue, aes(x = tissue, y = compound_short, fill = chemical_class)) +
  geom_tile(color = "white", linewidth = 0.8) +
  labs(
    title = "Presence of Putative Compounds in Leaf and Root Tissue",
    x = "Tissue",
    y = "Compound",
    fill = "Chemical Class"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 15, face = "bold"),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p5
ggsave("figures/Figure5_presence_absence_heatmap_tissue.png", p5, width = 7, height = 10, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 6: SAMPLE-BY-COMPOUND HEATMAP
# ============================================================

presence_sample <- gcms %>%
  distinct(sample_id, tissue, compound_short, chemical_class) %>%
  arrange(chemical_class, compound_short) %>%
  mutate(compound_short = factor(compound_short, levels = unique(compound_short)))

p6 <- ggplot(presence_sample, aes(x = sample_id, y = compound_short, fill = chemical_class)) +
  geom_tile(color = "white", linewidth = 0.8) +
  labs(
    title = "Compound Occurrence Across Individual Samples",
    x = "Sample",
    y = "Compound",
    fill = "Chemical Class"
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 13, face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p6
ggsave("figures/Figure6_sample_compound_heatmap.png", p6, width = 10, height = 10, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 7: DETECTION FREQUENCY OF COMPOUNDS
# ============================================================

p7 <- ggplot(compound_detection_counts, aes(x = reorder(compound_short, detections), y = detections, fill = chemical_class)) +
  geom_col(color = "black") +
  coord_flip() +
  labs(
    title = "Detection Frequency of Putative Compounds",
    x = "Compound",
    y = "Number of Detections",
    fill = "Chemical Class"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 13),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p7
ggsave("figures/Figure7_compound_detection_frequency.png", p7, width = 8, height = 8, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 8: R.MATCH SCORES BY COMPOUND
# ============================================================

rmatch_by_compound <- gcms %>%
  group_by(compound_short, chemical_class) %>%
  summarise(
    mean_r_match = mean(r_match, na.rm = TRUE),
    max_r_match = max(r_match, na.rm = TRUE),
    detections = n(),
    .groups = "drop"
  )

write_csv(rmatch_by_compound, "tables/rmatch_by_compound.csv")

p8 <- ggplot(rmatch_by_compound, aes(x = max_r_match, y = reorder(compound_short, max_r_match), color = chemical_class)) +
  geom_point(size = 5) +
  geom_vline(xintercept = 950, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 930, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 900, linetype = "dotted", linewidth = 1) +
  annotate("text", x = 952, y = 2, label = "Excellent ≥950", hjust = 0, size = 4) +
  annotate("text", x = 932, y = 4, label = "High ≥930", hjust = 0, size = 4) +
  annotate("text", x = 902, y = 6, label = "Good ≥900", hjust = 0, size = 4) +
  labs(
    title = "Reverse Match Scores for Putative GC-MS Identifications",
    x = "Maximum R.Match Score",
    y = "Compound",
    color = "Chemical Class"
  ) +
  xlim(780, 1000) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 13),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p8
ggsave("figures/Figure8_rmatch_scores_by_compound.png", p8, width = 9, height = 9, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 9: R.MATCH CONFIDENCE BINS
# ============================================================

rmatch_bins <- gcms %>%
  mutate(
    rmatch_bin = case_when(
      r_match >= 950 ~ "Excellent (≥950)",
      r_match >= 930 ~ "High (930–949)",
      r_match >= 900 ~ "Good (900–929)",
      r_match >= 850 ~ "Moderate (850–899)",
      TRUE ~ "Tentative (<850)"
    ),
    rmatch_bin = factor(
      rmatch_bin,
      levels = c(
        "Excellent (≥950)",
        "High (930–949)",
        "Good (900–929)",
        "Moderate (850–899)",
        "Tentative (<850)"
      )
    )
  ) %>%
  count(tissue, rmatch_bin, name = "observations")

write_csv(rmatch_bins, "tables/rmatch_confidence_bins.csv")

p9 <- ggplot(rmatch_bins, aes(x = rmatch_bin, y = observations, fill = tissue)) +
  geom_col(position = "dodge", color = "black") +
  labs(
    title = "Identification Confidence Based on R.Match Score",
    x = "R.Match Confidence Category",
    y = "Number of Observations",
    fill = "Tissue"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold", angle = 35, hjust = 1),
    axis.text.y = element_text(size = 13),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p9
ggsave("figures/Figure9_rmatch_confidence_bins.png", p9, width = 9, height = 5, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 10: RETENTION TIME DISTRIBUTION
# ============================================================

p10 <- ggplot(gcms, aes(x = retention_min, fill = tissue)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Retention Time Distribution of Detected Compounds",
    x = "Retention Time (min)",
    y = "Density",
    fill = "Tissue"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 15),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14)
  )

p10
ggsave("figures/Figure10_retention_time_distribution.png", p10, width = 8, height = 5, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 11: RETENTION TIME BY CHEMICAL CLASS
# ============================================================

p11 <- ggplot(gcms, aes(x = chemical_class, y = retention_min, fill = tissue)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75),
    size = 2.5,
    alpha = 0.8
  ) +
  labs(
    title = "Retention Time Patterns by Chemical Class",
    x = "Chemical Class",
    y = "Retention Time (min)",
    fill = "Tissue"
  ) +
  theme_bw(base_size = 15) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold", angle = 35, hjust = 1),
    axis.text.y = element_text(size = 13),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p11
ggsave("figures/Figure11_retention_time_by_class.png", p11, width = 9, height = 6, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 12: BUBBLE PLOT OF CHEMICAL CLASS RICHNESS
# ============================================================

p12 <- ggplot(chemical_class_summary, aes(x = tissue, y = chemical_class, size = unique_compounds, color = chemical_class)) +
  geom_point(alpha = 0.85) +
  scale_size(range = c(4, 14)) +
  labs(
    title = "Chemical Class Richness by Tissue",
    x = "Tissue",
    y = "Chemical Class",
    size = "Unique Compounds",
    color = "Chemical Class"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p12
ggsave("figures/Figure12_bubble_chemical_class_richness.png", p12, width = 8, height = 5, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 13: SHARED VS TISSUE-SPECIFIC COMPOUNDS
# ============================================================

shared_status <- gcms %>%
  distinct(tissue, compound_short) %>%
  group_by(compound_short) %>%
  summarise(
    tissues_detected = paste(sort(unique(as.character(tissue))), collapse = " + "),
    n_tissues = n_distinct(tissue),
    .groups = "drop"
  ) %>%
  mutate(
    status = case_when(
      tissues_detected == "Leaf" ~ "Leaf only",
      tissues_detected == "Root" ~ "Root only",
      n_tissues == 2 ~ "Shared",
      TRUE ~ "Other"
    )
  ) %>%
  count(status, name = "unique_compounds")

write_csv(shared_status, "tables/shared_vs_tissue_specific_compounds.csv")

p13 <- ggplot(shared_status, aes(x = status, y = unique_compounds, fill = status)) +
  geom_col(color = "black", width = 0.65) +
  labs(
    title = "Shared and Tissue-Specific Putative Compounds",
    x = "",
    y = "Number of Unique Compounds",
    fill = "Category"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 15, face = "bold"),
    axis.text.y = element_text(size = 14),
    legend.position = "none"
  )

p13
ggsave("figures/Figure13_shared_vs_tissue_specific.png", p13, width = 7, height = 5, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 14: MONOTERPENE / SESQUITERPENE / DITERPENOID ONLY
# ============================================================

terpene_data <- gcms %>%
  filter(chemical_class %in% c("Monoterpene", "Sesquiterpene", "Putative diterpenoid")) %>%
  distinct(tissue, compound_short, chemical_class)

terpene_summary <- terpene_data %>%
  count(tissue, chemical_class, name = "unique_compounds")

write_csv(terpene_summary, "tables/terpene_class_summary.csv")

p14 <- ggplot(terpene_summary, aes(x = chemical_class, y = unique_compounds, fill = tissue)) +
  geom_col(position = "dodge", color = "black") +
  labs(
    title = "Terpene Class Comparison Between Leaf and Root Tissue",
    x = "Terpene Class",
    y = "Number of Unique Compounds",
    fill = "Tissue"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(size = 19, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 14, face = "bold", angle = 25, hjust = 1),
    axis.text.y = element_text(size = 14),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 13)
  )

p14
ggsave("figures/Figure14_terpene_class_comparison.png", p14, width = 8, height = 5, dpi = 600)
```


```{r}
# ============================================================
# FIGURE 15: COMPOUND DETECTION ACROSS SAMPLES
# ============================================================

compound_by_sample <- gcms %>%
  distinct(sample_id, tissue, compound_short, chemical_class) %>%
  count(compound_short, chemical_class, name = "samples_detected") %>%
  arrange(desc(samples_detected))

write_csv(compound_by_sample, "tables/compound_detection_across_samples.csv")

p15 <- ggplot(compound_by_sample, aes(x = reorder(compound_short, samples_detected), y = samples_detected, fill = chemical_class)) +
  geom_col(color = "black") +
  coord_flip() +
  labs(
    title = "Number of Samples in Which Each Compound Was Detected",
    x = "Compound",
    y = "Number of Samples",
    fill = "Chemical Class"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 13),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

p15
ggsave("figures/Figure15_compound_detection_across_samples.png", p15, width = 8, height = 8, dpi = 600)

# ============================================================
# PRINT QUICK SUMMARY TO CONSOLE
# ============================================================

dataset_summary
sample_counts
compound_counts_tissue
chemical_class_summary
shared_status
```
```{r}

network_data <- gcms %>%
  distinct(tissue, compound_short, chemical_class)

ggplot(network_data, aes(x = tissue, y = reorder(compound_short, chemical_class))) +
  geom_line(aes(group = compound_short), color = "grey70", linewidth = 0.8) +
  geom_point(aes(color = chemical_class), size = 5) +
  labs(
    title = "Shared and Tissue-Specific Putative Metabolites",
    x = "Tissue",
    y = "Putative Compound",
    color = "Chemical Class"
  ) +
  theme_bw(base_size = 18) +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text.x = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 14),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 15)
  )

ggsave("Figure4_Network_Shared_Tissue_Specific_Metabolites.png", width = 9, height = 10, dpi = 600)

```

##retention time pattern
```{r}
compound_rt <- gcms %>%
  group_by(tissue, compound_short, chemical_class) %>%
  summarise(
    mean_rt = mean(retention_min, na.rm = TRUE),
    mean_rmatch = mean(r_match, na.rm = TRUE),
    detections = n(),
    .groups = "drop"
  )

p4 <- ggplot(compound_rt,
             aes(x = mean_rt, y = reorder(compound_short, mean_rt))) +
  geom_point(aes(color = tissue, size = mean_rmatch), alpha = 0.85) +
  facet_wrap(~ tissue, nrow = 1) +
  theme_bw(base_size = 15) +
  labs(
    title = "Retention Time Pattern of Putative Compounds by Tissue",
    x = "Mean Retention Time (min)",
    y = "Putative Compound",
    color = "Tissue",
    size = "Mean R.Match"
  ) +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 11),
    strip.text = element_text(size = 16, face = "bold")
  )

p4

ggsave("figures/Figure4_compound_retention_patterns.png", p4,
       width = 11, height = 9, dpi = 600)
```


```{r}
pathway_table <- gcms %>%
  distinct(tissue, compound_short, chemical_class) %>%
  mutate(
    pathway = case_when(
      chemical_class == "Monoterpene" ~ "MEP → GPP → Monoterpenes",
      chemical_class == "Sesquiterpene" ~ "MVA → FPP → Sesquiterpenes",
      str_detect(chemical_class, "Diterpen") ~ "GGPP → Diterpenoids",
      TRUE ~ "Other / Unknown"
    )
  )

p5 <- ggplot(pathway_table,
             aes(x = pathway, y = compound_short, color = tissue)) +
  geom_point(size = 4) +
  theme_bw(base_size = 15) +
  labs(
    title = "Detected Putative Metabolites Mapped to Biosynthetic Pathways",
    x = "Known Biosynthetic Pathway",
    y = "Putative Compound",
    color = "Tissue"
  ) +
  theme(
    plot.title = element_text(size = 21, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 15, face = "bold")
  )

p5

ggsave("figures/Figure5_biosynthetic_pathway_mapping.png", p5,
       width = 11, height = 9, dpi = 600)
```



```{r}

```


```{r}
# Venn diagram using only ggplot2/tidyverse

library(tidyverse)

gcms <- read_csv(
  "C:/Users/norba/Downloads/GCMS_R_Project_Report/GCMS_R_Project/data/gcms_tidy_r_input.csv",
  show_col_types = FALSE
)

gcms <- gcms %>%
  mutate(
    tissue = str_to_title(tissue),
    compound_short = str_replace_all(compound_short, "_", "-")
  )

leaf_compounds <- gcms %>%
  filter(tissue == "Leaf") %>%
  distinct(compound_short)

root_compounds <- gcms %>%
  filter(tissue == "Root") %>%
  distinct(compound_short)

shared <- intersect(leaf_compounds$compound_short, root_compounds$compound_short)
leaf_only <- setdiff(leaf_compounds$compound_short, root_compounds$compound_short)
root_only <- setdiff(root_compounds$compound_short, leaf_compounds$compound_short)

venn_counts <- tibble(
  category = c("Leaf only", "Shared", "Root only"),
  count = c(length(leaf_only), length(shared), length(root_only))
)

print(venn_counts)

p_venn <- ggplot() +
  annotate("point", x = -0.7, y = 0, size = 95, shape = 21, alpha = 0.25, fill = "pink") +
  annotate("point", x = 0.7, y = 0, size = 95, shape = 21, alpha = 0.25, fill = "lightblue") +
  annotate("text", x = -1.25, y = 0, label = length(leaf_only), size = 10, fontface = "bold") +
  annotate("text", x = 0, y = 0, label = length(shared), size = 10, fontface = "bold") +
  annotate("text", x = 1.25, y = 0, label = length(root_only), size = 10, fontface = "bold") +
  annotate("text", x = -0.7, y = 1.1, label = "Leaf", size = 8, fontface = "bold") +
  annotate("text", x = 0.7, y = 1.1, label = "Root", size = 8, fontface = "bold") +
  annotate("text", x = -1.25, y = -0.9, label = "Leaf only", size = 5) +
  annotate("text", x = 0, y = -0.9, label = "Shared", size = 5) +
  annotate("text", x = 1.25, y = -0.9, label = "Root only", size = 5) +
  coord_fixed() +
  theme_void() +
  labs(title = "Shared and Tissue-Specific Putative Metabolites") +
  theme(plot.title = element_text(size = 19, face = "bold", hjust = 0.5))

p_venn

ggsave("figures/Figure_Venn_leaf_root_compounds.png", p_venn,
       width = 8, height = 6, dpi = 600)
```



##LC-MS data


```{r}
install.packages(c(
  "readxl",
  "dplyr",
  "tidyr",
  "stringr",
  "ggplot2",
  "ggrepel",
  "tibble"
))
```


```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(tibble)
```
```{r}
file_path <- "C:/Users/norba/Downloads/postive mode height.xlsx"

msdial <- read_excel(
  path = file_path,
  sheet = "postive mode height",
  skip = 4
)

dim(msdial)
names(msdial)
```

#identify raw sample-intensity coloumns 
```{r}
sample_columns <- c(
  "blank_1", "blank_2", "blank_3", "blank_4",
  "exudate_1", "exudate_2", "exudate_3",
  "FC_1", "FC_2", "FC_3",
  "fungi_old_1", "fungi_old_2", "fungi_old_3",
  "leaf_1", "leaf_2", "leaf_3",
  "new_in_P08", "new_in_P09", "new_in_P10",
  "new_out_P08", "new_out_P09", "new_out_P10",
  "root_1", "root_2", "root_3",
  "Soil_control_1", "Soil_control_2", "Soil_control_3"
)

stopifnot(all(sample_columns %in% names(msdial)))
```


#convert intensities to numeric values 
```{r}
msdial <- msdial %>%
  mutate(
    across(
      all_of(sample_columns),
      ~ as.numeric(.x)
    )
  )
```

#create sample metadata, a table to tell us which biological group each column belongs to 
```{r}
sample_metadata <- tibble(
  sample = sample_columns
) %>%
  mutate(
    group = case_when(
      str_detect(sample, "^blank")        ~ "Blank",
      str_detect(sample, "^exudate")      ~ "Plant-soil exudate",
      str_detect(sample, "^Soil_control") ~ "Soil control",
      str_detect(sample, "^FC")           ~ "Fungal control",
      str_detect(sample, "^fungi_old")    ~ "Fungi old",
      str_detect(sample, "^leaf")         ~ "Leaf",
      str_detect(sample, "^root")         ~ "Root",
      str_detect(sample, "^new_in")       ~ "Fungal agar in",
      str_detect(sample, "^new_out")      ~ "Fungal agar out",
      TRUE                                ~ "Unknown"
    ),
    replicate = case_when(
      str_detect(sample, "P08") ~ "1",
      str_detect(sample, "P09") ~ "2",
      str_detect(sample, "P10") ~ "3",
      TRUE ~ str_extract(sample, "\\d+$")
    ),
    broad_matrix = case_when(
      group %in% c("Plant-soil exudate", "Soil control") ~ "Soil extract",
      group %in% c(
        "Fungal control",
        "Fungi old",
        "Fungal agar in",
        "Fungal agar out"
      ) ~ "Fungal agar",
      group %in% c("Leaf", "Root") ~ "Plant tissue",
      group == "Blank" ~ "Blank",
      TRUE ~ "Unknown"
    )
  )

sample_metadata
```

#ensure that each group has the expected number of samples
```{r}
sample_metadata %>%
  count(group)
```

#check sample quality before filtering. We look at the total intensity because it is the sum of the signal intensities of all ions detected by the mass spectrometer at every given point in time. So, it combines every single mass-to-charge value recorded during a scan, including both your target analytes and background chemical noise. 

#quality control: a sudden drop of flatline in total intensities could signal that something is wrong, a clogged emmitter, a bad electrical connection to the ion source

#batch consistency, comparing the total ion profiles of repeat injections or blanks helps ensure that instrument sensntivitiy remains stable throughout a large sample que. 

#sample load estimation. a massice spike or an unexpectedly weak overall signal gives you an immediate, macro-level look at whether your sample was over-concentrated or too dilute before you dive into deeper data analysis. 
```{r}
sample_qc <- tibble(
  sample = sample_columns,
  total_intensity = colSums(
    as.matrix(msdial[, sample_columns]),
    na.rm = TRUE
  ),
  detected_features = colSums(
    as.matrix(msdial[, sample_columns]) > 0,
    na.rm = TRUE
  )
) %>%
  left_join(sample_metadata, by = "sample")

sample_qc
```


#now we plot the total signal
```{r}
ggplot(
  sample_qc,
  aes(
    x = reorder(sample, total_intensity),
    y = total_intensity,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Total LC-MS signal per sample",
    x = NULL,
    y = "Sum of feature intensities"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```


#plot the number of detected features
```{r}
ggplot(
  sample_qc,
  aes(
    x = reorder(sample, detected_features),
    y = detected_features,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of detected LC-MS features per sample",
    x = NULL,
    y = "Features with intensity > 0"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

#check to see what samples have no detected features
```{r}
sample_qc %>%
  filter(detected_features == 0)
```


#prepare the quantitative feature matrix
```{r}
feature_table <- msdial %>%
  select(
    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `Fill %`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    `S/N average`,
    all_of(sample_columns)
  )
```


#make unique feature names

```{r}
feature_table <- feature_table %>%
  mutate(
    feature_name = paste0(
      "F", `Alignment ID`,
      "_mz", round(`Average Mz`, 4),
      "_rt", round(`Average Rt(min)`, 2)
    )
  )
```


#remove the empty blank injection (as it seems to have an issue)
```{r}
valid_blank_columns <- c(
  "blank_1",
  "blank_2",
  "blank_3"
)

biological_columns <- setdiff(
  sample_columns,
  c(valid_blank_columns, "blank_4")
)
```

#check to see how many biological samples are there
```{r}
length(biological_columns)
```

#calculate the mean from the blank signal, and biological signal for each feature.
```{r}
feature_qc <- feature_table %>%
  mutate(
    blank_mean = rowMeans(
      across(all_of(valid_blank_columns)),
      na.rm = TRUE
    ),

    biological_mean = rowMeans(
      across(all_of(biological_columns)),
      na.rm = TRUE
    ),

    maximum_biological_signal = apply(
      as.matrix(pick(all_of(biological_columns))),
      1,
      max,
      na.rm = TRUE
    ),

    sample_to_blank_ratio =
      (maximum_biological_signal + 1) /
      (blank_mean + 1)
  )
```


#apply a reasonable exploratory filter to keep features whose signal in at least one biological sample is at least 5 times the average blank signal. 


```{r}
feature_qc <- feature_qc %>%
  filter(sample_to_blank_ratio >= 5)
```

#check how many remain
```{r}
nrow(feature_table)
nrow(feature_qc)
```

#apply a prevalence filter. PCAs should detect features in enough samples to provide stable variation. so we want to keep features detected in at least two biological samples, not only within one of the three replicate groups. 
```{r}
feature_qc <- feature_qc %>%
  mutate(
    number_detected = rowSums(
      across(all_of(biological_columns), ~ .x > 0),
      na.rm = TRUE
    )
  ) %>%
  filter(number_detected >= 2)
```


#group columns together appropriately
```{r}
group_columns <- list(
  exudate = c("exudate_1", "exudate_2", "exudate_3"),
  fungal_control = c("FC_1", "FC_2", "FC_3"),
  fungi_old = c("fungi_old_1", "fungi_old_2", "fungi_old_3"),
  leaf = c("leaf_1", "leaf_2", "leaf_3"),
  fungal_agar_in = c(
    "new_in_P08",
    "new_in_P09",
    "new_in_P10"
  ),
  fungal_agar_out = c(
    "new_out_P08",
    "new_out_P09",
    "new_out_P10"
  ),
  root = c("root_1", "root_2", "root_3"),
  soil_control = c(
    "Soil_control_1",
    "Soil_control_2",
    "Soil_control_3"
  )
)

presence_by_group <- sapply(
  group_columns,
  function(cols) {
    rowSums(
      as.matrix(feature_qc[, cols]) > 0,
      na.rm = TRUE
    ) >= 2
  }
)

feature_qc$present_in_any_group <- apply(
  presence_by_group,
  1,
  any
)

feature_qc <- feature_qc %>%
  filter(present_in_any_group)
```


#build PCA Matrix
```{r}
intensity_matrix <- feature_qc %>%
  select(feature_name, all_of(biological_columns)) %>%
  column_to_rownames("feature_name") %>%
  as.matrix()

pca_matrix <- t(intensity_matrix)

dim(pca_matrix)
```

#replace zeros before log transformation, zero means the detection or peak picking thredgold was low. so we use a small feature-specific replacement equal to half the smallest positive value for that feature.
```{r}
replace_zero_half_min <- function(x) {
  positive_values <- x[x > 0 & is.finite(x)]

  if (length(positive_values) == 0) {
    return(rep(NA_real_, length(x)))
  }

  replacement <- min(positive_values) / 2
  x[x == 0 | is.na(x)] <- replacement
  x
}

pca_matrix_imputed <- apply(
  pca_matrix,
  2,
  replace_zero_half_min
)
```


#now we remove any feature that could not be imputed
```{r}
pca_matrix_imputed <- pca_matrix_imputed[
  ,
  colSums(is.na(pca_matrix_imputed)) == 0,
  drop = FALSE
]
```



#normalize sample intensities, median normalization
```{r}
sample_medians <- apply(
  pca_matrix_imputed,
  1,
  median,
  na.rm = TRUE
)

global_median <- median(sample_medians)

pca_matrix_normalized <- sweep(
  pca_matrix_imputed,
  1,
  sample_medians,
  "/"
) * global_median
```



#total-ion-signal-normalization
```{r}
sample_sums <- rowSums(pca_matrix_imputed)

pca_matrix_tic <- sweep(
  pca_matrix_imputed,
  1,
  sample_sums,
  "/"
) * median(sample_sums)
```


#log transform the data
```{r}
pca_matrix_log <- log2(
  pca_matrix_normalized + 1
)
```


#remove zero-variance features
```{r}
feature_variance <- apply(
  pca_matrix_log,
  2,
  var,
  na.rm = TRUE
)

pca_matrix_log <- pca_matrix_log[
  ,
  feature_variance > 0 &
    is.finite(feature_variance),
  drop = FALSE
]
```


#run the PCA
```{r}
pca_result <- prcomp(
  pca_matrix_log,
  center = TRUE,
  scale. = TRUE
)

summary(pca_result)
```

#calculate the proportion of variance explained
```{r}
variance_explained <- (
  pca_result$sdev^2 /
  sum(pca_result$sdev^2)
) * 100

variance_explained[1:10]
```


#prepare PCA scores for plotting
```{r}
pca_scores <- as.data.frame(
  pca_result$x
) %>%
  rownames_to_column("sample") %>%
  left_join(
    sample_metadata,
    by = "sample"
  )
```


#plot PC1 VS pc2
```{r}
ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC2,
    shape = broad_matrix,
    fill = group
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    title = "PCA of positive-mode LC-MS features",
    subtitle = "Blank-filtered, median-normalized and log2-transformed",
    x = paste0(
      "PC1 (",
      round(variance_explained[1], 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(variance_explained[2], 1),
      "%)"
    ),
    fill = "Sample group",
    shape = "Broad matrix"
  ) +
  theme_classic() +
  theme(
    legend.position = "right"
  )
```

#plot pc1 vs pc3
```{r}
ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC3,
    shape = broad_matrix,
    fill = group
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    title = "PCA: PC1 versus PC3",
    x = paste0(
      "PC1 (",
      round(variance_explained[1], 1),
      "%)"
    ),
    y = paste0(
      "PC3 (",
      round(variance_explained[3], 1),
      "%)"
    )
  ) +
  theme_classic()
```

#make scree plot to show how much variation each principal component explains

```{r}
scree_data <- tibble(
  principal_component = seq_along(variance_explained),
  variance_percent = variance_explained
) %>%
  slice_head(n = 15)

ggplot(
  scree_data,
  aes(
    x = principal_component,
    y = variance_percent
  )
) +
  geom_col() +
  geom_point() +
  geom_line() +
  scale_x_continuous(
    breaks = scree_data$principal_component
  ) +
  labs(
    title = "PCA scree plot",
    x = "Principal component",
    y = "Variance explained (%)"
  ) +
  theme_classic()
```


#plant-soil exudate versus soil control
```{r}
soil_samples <- sample_metadata %>%
  filter(
    group %in% c(
      "Plant-soil exudate",
      "Soil control"
    )
  ) %>%
  pull(sample)

soil_pca_matrix <- pca_matrix_log[
  soil_samples,
  ,
  drop = FALSE
]

soil_variances <- apply(
  soil_pca_matrix,
  2,
  var
)

soil_pca_matrix <- soil_pca_matrix[
  ,
  soil_variances > 0,
  drop = FALSE
]

soil_pca <- prcomp(
  soil_pca_matrix,
  center = TRUE,
  scale. = TRUE
)

soil_variance <- (
  soil_pca$sdev^2 /
  sum(soil_pca$sdev^2)
) * 100

soil_scores <- as.data.frame(
  soil_pca$x
) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

ggplot(
  soil_scores,
  aes(
    x = PC1,
    y = PC2,
    fill = group
  )
) +
  geom_point(
    shape = 21,
    size = 5,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample)
  ) +
  labs(
    title = "PCA: plant-soil exudate versus soil control",
    x = paste0(
      "PC1 (",
      round(soil_variance[1], 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(soil_variance[2], 1),
      "%)"
    )
  ) +
  theme_classic()
```

#fungal agar PCA
```{r}
fungal_groups <- c(
  "Fungal control",
  "Fungal agar in",
  "Fungal agar out"
)

fungal_samples <- sample_metadata %>%
  filter(group %in% fungal_groups) %>%
  pull(sample)

fungal_pca_matrix <- pca_matrix_log[
  fungal_samples,
  ,
  drop = FALSE
]

fungal_variances <- apply(
  fungal_pca_matrix,
  2,
  var
)

fungal_pca_matrix <- fungal_pca_matrix[
  ,
  fungal_variances > 0,
  drop = FALSE
]

fungal_pca <- prcomp(
  fungal_pca_matrix,
  center = TRUE,
  scale. = TRUE
)

fungal_variance <- (
  fungal_pca$sdev^2 /
  sum(fungal_pca$sdev^2)
) * 100

fungal_scores <- as.data.frame(
  fungal_pca$x
) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

ggplot(
  fungal_scores,
  aes(
    x = PC1,
    y = PC2,
    fill = group
  )
) +
  geom_point(
    shape = 21,
    size = 5,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample)
  ) +
  labs(
    title = "PCA of fungal agar samples",
    subtitle = "Agar beneath colony, outside colony and fungal control",
    x = paste0(
      "PC1 (",
      round(fungal_variance[1], 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(fungal_variance[2], 1),
      "%)"
    )
  ) +
  theme_classic()
```

#leaf versus root PCA
```{r}
tissue_samples <- sample_metadata %>%
  filter(group %in% c("Leaf", "Root")) %>%
  pull(sample)

tissue_pca_matrix <- pca_matrix_log[
  tissue_samples,
  ,
  drop = FALSE
]

tissue_variances <- apply(
  tissue_pca_matrix,
  2,
  var
)

tissue_pca_matrix <- tissue_pca_matrix[
  ,
  tissue_variances > 0,
  drop = FALSE
]

tissue_pca <- prcomp(
  tissue_pca_matrix,
  center = TRUE,
  scale. = TRUE
)

tissue_variance <- (
  tissue_pca$sdev^2 /
  sum(tissue_pca$sdev^2)
) * 100

tissue_scores <- as.data.frame(
  tissue_pca$x
) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

ggplot(
  tissue_scores,
  aes(
    x = PC1,
    y = PC2,
    fill = group
  )
) +
  geom_point(
    shape = 21,
    size = 5,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample)
  ) +
  labs(
    title = "PCA of leaf and root extracts",
    x = paste0(
      "PC1 (",
      round(tissue_variance[1], 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(tissue_variance[2], 1),
      "%)"
    )
  ) +
  theme_classic()
```



#step 2, identify the features driving the PCA. PCA scores tell you where the samples fall. PCA loadings tell you which LC-MS features are pulling samples apart

#extract the PCA loadings
```{r}
pca_loadings <- as.data.frame(
  pca_result$rotation
) %>%
  rownames_to_column("feature_name")
```


#add MS-DIAL annotation information
```{r}
feature_annotations <- feature_qc %>%
  select(
    feature_name,
    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    `S/N average`,
    sample_to_blank_ratio,
    number_detected
  )

pca_loadings <- pca_loadings %>%
  left_join(
    feature_annotations,
    by = "feature_name"
  )
```


#find the strongest PC1 loadings
```{r}
top_pc1_loadings <- pca_loadings %>%
  mutate(
    absolute_PC1 = abs(PC1)
  ) %>%
  arrange(desc(absolute_PC1)) %>%
  slice_head(n = 30)

top_pc1_loadings %>%
  select(
    feature_name,
    PC1,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    Ontology,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```


#plot top PC1 features
```{r}
top_pc1_plot <- top_pc1_loadings %>%
  mutate(
    display_name = case_when(
      !is.na(`Metabolite name`) &
        `Metabolite name` != "" &
        `Metabolite name` != "Unknown" ~
        paste0(
          `Metabolite name`,
          " | ",
          feature_name
        ),

      TRUE ~ feature_name
    )
  )

ggplot(
  top_pc1_plot,
  aes(
    x = reorder(display_name, PC1),
    y = PC1
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Features contributing most strongly to PC1",
    x = NULL,
    y = "PC1 loading"
  ) +
  theme_classic()
```

#create the PC2 loadings data

```{r}
top_pc2_loadings <- pca_loadings %>%
  mutate(
    absolute_PC2 = abs(PC2)
  ) %>%
  arrange(desc(absolute_PC2)) %>%
  slice_head(n = 30)
```


#check to make sure it works
```{r}
head(top_pc2_loadings)
```


#create the plotting table
```{r}
top_pc2_plot <- top_pc2_loadings %>%
  mutate(
    display_name = case_when(
      !is.na(`Metabolite name`) &
        `Metabolite name` != "" &
        `Metabolite name` != "Unknown" ~
        paste0(
          `Metabolite name`,
          " | ",
          feature_name
        ),

      TRUE ~ feature_name
    )
  )
```

#plot PC2 loadings
```{r}
ggplot(
  top_pc2_plot,
  aes(
    x = reorder(display_name, PC2),
    y = PC2
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Features contributing most strongly to PC2",
    x = NULL,
    y = "PC2 loading"
  ) +
  theme_classic()
```

#export the results
```{r}
write.csv(
  pca_scores,
  "positive_mode_PCA_scores.csv",
  row.names = FALSE
)

write.csv(
  pca_loadings,
  "positive_mode_PCA_loadings.csv",
  row.names = FALSE
)

write.csv(
  top_pc1_loadings,
  "positive_mode_top_PC1_loadings.csv",
  row.names = FALSE
)

write.csv(
  top_pc2_loadings,
  "positive_mode_top_PC2_loadings.csv",
  row.names = FALSE
)
```


#save PCA images
```{r}
broad_pca_plot <- ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC2,
    shape = broad_matrix,
    fill = group
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    title = "PCA of positive-mode LC-MS features",
    subtitle = "Blank-filtered, normalized and log2-transformed",
    x = paste0(
      "PC1 (",
      round(variance_explained[1], 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(variance_explained[2], 1),
      "%)"
    )
  ) +
  theme_classic()

broad_pca_plot
```

#save it
```{r}
ggsave(
  filename = "positive_mode_broad_PCA.png",
  plot = broad_pca_plot,
  width = 10,
  height = 7,
  dpi = 300
)
```



#print the variance explained
```{r}
variance_explained[1:10]
```
#create a clean table
```{r}
variance_table <- tibble(
  PC = paste0("PC", seq_along(variance_explained)),
  variance_percent = variance_explained,
  cumulative_percent = cumsum(variance_explained)
)

head(variance_table, 10)
```


#make a scree plot
```{r}
scree_data <- variance_table %>%
  slice_head(n = 15)

scree_plot <- ggplot(
  scree_data,
  aes(
    x = factor(PC, levels = PC),
    y = variance_percent,
    group = 1
  )
) +
  geom_col() +
  geom_point(size = 2) +
  geom_line() +
  labs(
    title = "Positive-mode LC-MS PCA scree plot",
    x = "Principal component",
    y = "Variance explained (%)"
  ) +
  theme_classic()

scree_plot
```

#plot pc1 vs pc3
```{r}
pc1_pc3_plot <- ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC3,
    fill = group,
    shape = broad_matrix
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  ggrepel::geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    title = "PCA: PC1 versus PC3",
    x = paste0(
      "PC1 (",
      round(variance_explained[1], 1),
      "%)"
    ),
    y = paste0(
      "PC3 (",
      round(variance_explained[3], 1),
      "%)"
    ),
    fill = "Sample group",
    shape = "Matrix"
  ) +
  theme_classic()

pc1_pc3_plot
```


#plot pc2 vs pc3
```{r}
pc2_pc3_plot <- ggplot(
  pca_scores,
  aes(
    x = PC2,
    y = PC3,
    fill = group,
    shape = broad_matrix
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  ggrepel::geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(
    title = "PCA: PC2 versus PC3",
    x = paste0(
      "PC2 (",
      round(variance_explained[2], 1),
      "%)"
    ),
    y = paste0(
      "PC3 (",
      round(variance_explained[3], 1),
      "%)"
    )
  ) +
  theme_classic()

pc2_pc3_plot
```

#inspect pca sample scores
```{r}
pca_scores %>%
  select(
    sample,
    group,
    broad_matrix,
    PC1,
    PC2,
    PC3
  ) %>%
  arrange(group)
```

#export it
```{r}
write.csv(
  pca_scores,
  "positive_mode_PCA_sample_scores.csv",
  row.names = FALSE
)
```


#extract pc1, pc2, pc3 loadings
```{r}
pca_loadings <- as.data.frame(
  pca_result$rotation
) %>%
  rownames_to_column("feature_name")
```

#add the MS-DIAL information

```{r}
feature_annotations <- feature_qc %>%
  select(
    feature_name,
    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    `S/N average`,
    sample_to_blank_ratio,
    number_detected
  )

pca_loadings <- pca_loadings %>%
  left_join(
    feature_annotations,
    by = "feature_name"
  )
```


#export all loadings
```{r}
write.csv(
  pca_loadings,
  "positive_mode_all_PCA_loadings.csv",
  row.names = FALSE
)
```


#create a reusable function for top loadings
```{r}
get_top_loadings <- function(
  loading_table,
  pc_column,
  number_to_keep = 30
) {

  loading_table %>%
    mutate(
      selected_loading = .data[[pc_column]],
      absolute_loading = abs(.data[[pc_column]])
    ) %>%
    arrange(desc(absolute_loading)) %>%
    slice_head(n = number_to_keep)
}
```


#generate top features
```{r}
top_pc1_loadings <- get_top_loadings(
  pca_loadings,
  "PC1",
  30
)

top_pc2_loadings <- get_top_loadings(
  pca_loadings,
  "PC2",
  30
)

top_pc3_loadings <- get_top_loadings(
  pca_loadings,
  "PC3",
  30
)
```


#view the most useful coloumns
```{r}
top_pc1_loadings %>%
  select(
    feature_name,
    PC1,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```

#repeat for PC2
```{r}
top_pc2_loadings %>%
  select(
    feature_name,
    PC2,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```

#create a combined list of important PCA features
```{r}
important_pca_features <- pca_loadings %>%
  mutate(
    maximum_absolute_loading = pmax(
      abs(PC1),
      abs(PC2),
      na.rm = TRUE
    )
  ) %>%
  arrange(desc(maximum_absolute_loading)) %>%
  slice_head(n = 50)

important_pca_features %>%
  select(
    feature_name,
    PC1,
    PC2,
    maximum_absolute_loading,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`
  )
```

#export the data completely 
```{r}
write.csv(
  important_pca_features,
  "positive_mode_top_PCA_features.csv",
  row.names = FALSE
)
```


#make a loading scatterplot
```{r}
loading_plot_data <- pca_loadings %>%
  mutate(
    maximum_absolute_loading = pmax(
      abs(PC1),
      abs(PC2),
      na.rm = TRUE
    ),
    label = if_else(
      rank(
        -maximum_absolute_loading,
        ties.method = "first"
      ) <= 20,
      feature_name,
      NA_character_
    )
  )
```


#plot it
```{r}
loading_scatter_plot <- ggplot(
  loading_plot_data,
  aes(
    x = PC1,
    y = PC2
  )
) +
  geom_point(alpha = 0.5) +
  ggrepel::geom_text_repel(
    aes(label = label),
    size = 3,
    na.rm = TRUE,
    max.overlaps = Inf
  ) +
  labs(
    title = "PCA loading plot",
    subtitle = "Features driving PC1 and PC2",
    x = "PC1 loading",
    y = "PC2 loading"
  ) +
  theme_classic()

loading_scatter_plot
```
install pheatmap
```{r}
install.packages("pheatmap")
```
#load the map
```{r}
library(pheatmap)
```


#select the top 30 features
```{r}
top_feature_names <- important_pca_features %>%
  slice_head(n = 30) %>%
  pull(feature_name)
```

#extract their normalized, log-transformed intensities
```{r}
heatmap_matrix <- pca_matrix_log[
  ,
  colnames(pca_matrix_log) %in% top_feature_names,
  drop = FALSE
]
```

#transpose so features are rows
```{r}
heatmap_matrix <- t(heatmap_matrix)
```

#create sample annotations
```{r}
heatmap_annotation <- sample_metadata %>%
  filter(sample %in% colnames(heatmap_matrix)) %>%
  select(
    sample,
    group,
    broad_matrix
  ) %>%
  column_to_rownames("sample")
```

#order the annotation to match the heatmap columns
```{r}
heatmap_annotation <- heatmap_annotation[
  colnames(heatmap_matrix),
  ,
  drop = FALSE
]
```


#draw the heatmap
```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  annotation_col = heatmap_annotation,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "Top features driving positive-mode PCA"
)
```

#run these checks/trouble shooting
```{r}
dim(heatmap_matrix)
dim(heatmap_annotation)

colnames(heatmap_matrix)
rownames(heatmap_annotation)

setdiff(
  colnames(heatmap_matrix),
  rownames(heatmap_annotation)
)
```


#rebuild annotation safely
```{r}
heatmap_annotation <- sample_metadata %>%
  filter(sample %in% colnames(heatmap_matrix)) %>%
  select(
    sample,
    group,
    broad_matrix
  ) %>%
  distinct(sample, .keep_all = TRUE) %>%
  as.data.frame()

rownames(heatmap_annotation) <- heatmap_annotation$sample
heatmap_annotation$sample <- NULL
```

#order it to match the heatmap
```{r}
heatmap_annotation <- heatmap_annotation[
  colnames(heatmap_matrix),
  ,
  drop = FALSE
]
```


#check for missing values
```{r}
heatmap_annotation

colSums(is.na(heatmap_annotation))
```

#convert annotations to factors
```{r}
heatmap_annotation[] <- lapply(
  heatmap_annotation,
  function(x) factor(as.character(x))
)
```

#rerun heatmap
```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  annotation_col = heatmap_annotation,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "Top features driving positive-mode PCA"
)
```

#the issue persists, does the heatmap work without the annotations
```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "Top features driving positive-mode PCA"
)
```

#create a general pairwise comparison function to use the normalized, log-transformed matrix 
```{r}
pca_matrix_log
```


```{r}
run_pairwise_comparison <- function(
    data_matrix,
    metadata,
    annotation_table,
    group_1,
    group_2
) {

  group_1_samples <- metadata %>%
    filter(group == group_1) %>%
    pull(sample)

  group_2_samples <- metadata %>%
    filter(group == group_2) %>%
    pull(sample)

  group_1_samples <- intersect(
    group_1_samples,
    rownames(data_matrix)
  )

  group_2_samples <- intersect(
    group_2_samples,
    rownames(data_matrix)
  )

  message(
    group_1, ": ",
    paste(group_1_samples, collapse = ", ")
  )

  message(
    group_2, ": ",
    paste(group_2_samples, collapse = ", ")
  )

  results <- lapply(
    colnames(data_matrix),
    function(feature) {

      x <- as.numeric(
        data_matrix[group_1_samples, feature]
      )

      y <- as.numeric(
        data_matrix[group_2_samples, feature]
      )

      p_value <- tryCatch(
        t.test(
          x,
          y,
          var.equal = FALSE
        )$p.value,
        error = function(e) NA_real_
      )

      tibble(
        feature_name = feature,

        group_1 = group_1,
        group_2 = group_2,

        group_1_mean = mean(
          x,
          na.rm = TRUE
        ),

        group_2_mean = mean(
          y,
          na.rm = TRUE
        ),

        log2_fold_change =
          mean(x, na.rm = TRUE) -
          mean(y, na.rm = TRUE),

        group_1_sd = sd(
          x,
          na.rm = TRUE
        ),

        group_2_sd = sd(
          y,
          na.rm = TRUE
        ),

        p_value = p_value
      )
    }
  ) %>%
    bind_rows() %>%
    mutate(
      adjusted_p_value = p.adjust(
        p_value,
        method = "BH"
      ),

      fold_change = 2^log2_fold_change,

      direction = case_when(
        log2_fold_change > 0 ~
          paste0("Higher in ", group_1),

        log2_fold_change < 0 ~
          paste0("Higher in ", group_2),

        TRUE ~ "No difference"
      )
    ) %>%
    left_join(
      annotation_table,
      by = "feature_name"
    ) %>%
    arrange(
      adjusted_p_value,
      p_value
    )

  return(results)
}
```

#check input matrix first
```{r}
dim(pca_matrix_log)
rownames(pca_matrix_log)
colnames(pca_matrix_log)[1:10]
```
#exudate vs soil control
```{r}
exudate_vs_soil <- run_pairwise_comparison(
  data_matrix = pca_matrix_log,
  metadata = sample_metadata,
  annotation_table = feature_annotations,
  group_1 = "Plant-soil exudate",
  group_2 = "Soil control"
)
```

#view the first results
```{r}
exudate_vs_soil %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    group_1_mean,
    group_2_mean,
    log2_fold_change,
    fold_change,
    p_value,
    adjusted_p_value,
    direction
  ) %>%
  head(30)
```



#count nominally significant features
```{r}
exudate_vs_soil %>%
  summarise(
    total_features = n(),

    raw_p_below_0.05 = sum(
      p_value < 0.05,
      na.rm = TRUE
    ),

    FDR_below_0.05 = sum(
      adjusted_p_value < 0.05,
      na.rm = TRUE
    ),

    raw_p_and_twofold = sum(
      p_value < 0.05 &
        abs(log2_fold_change) >= 1,
      na.rm = TRUE
    ),

    FDR_and_twofold = sum(
      adjusted_p_value < 0.05 &
        abs(log2_fold_change) >= 1,
      na.rm = TRUE
    )
  )
```



#export
```{r}
write.csv(
  exudate_vs_soil,
  "positive_mode_exudate_vs_soil_statistics.csv",
  row.names = FALSE
)
```


#fungal agar in vs fungal agar out
```{r}
fungal_in_vs_out <- run_pairwise_comparison(
  data_matrix = pca_matrix_log,
  metadata = sample_metadata,
  annotation_table = feature_annotations,
  group_1 = "Fungal agar in",
  group_2 = "Fungal agar out"
)
```

#inspect the data
```{r}
fungal_in_vs_out %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    log2_fold_change,
    fold_change,
    p_value,
    adjusted_p_value,
    direction
  ) %>%
  head(30)
```


#export the data
```{r}
write.csv(
  fungal_in_vs_out,
  "positive_mode_fungal_in_vs_out_statistics.csv",
  row.names = FALSE
)
```



#fungal agar vs agar control
```{r}
fungal_in_vs_control <- run_pairwise_comparison(
  data_matrix = pca_matrix_log,
  metadata = sample_metadata,
  annotation_table = feature_annotations,
  group_1 = "Fungal agar in",
  group_2 = "Fungal control"
)
```

#export the data
```{r}
write.csv(
  fungal_in_vs_control,
  "positive_mode_fungal_in_vs_control_statistics.csv",
  row.names = FALSE
)
```


#leaf vs root
```{r}
leaf_vs_root <- run_pairwise_comparison(
  data_matrix = pca_matrix_log,
  metadata = sample_metadata,
  annotation_table = feature_annotations,
  group_1 = "Leaf",
  group_2 = "Root"
)
```
#export
```{r}
write.csv(
  leaf_vs_root,
  "positive_mode_leaf_vs_root_statistics.csv",
  row.names = FALSE
)
```

#create a volcano plot
```{r}
exudate_vs_soil_plot <- exudate_vs_soil %>%
  mutate(
    minus_log10_p = -log10(
      pmax(p_value, 1e-300)
    ),

    significance = case_when(
      adjusted_p_value < 0.05 &
        log2_fold_change >= 1 ~
        "Higher in exudate: FDR < 0.05",

      adjusted_p_value < 0.05 &
        log2_fold_change <= -1 ~
        "Higher in soil: FDR < 0.05",

      p_value < 0.05 &
        log2_fold_change >= 1 ~
        "Higher in exudate: raw p < 0.05",

      p_value < 0.05 &
        log2_fold_change <= -1 ~
        "Higher in soil: raw p < 0.05",

      TRUE ~ "Not selected"
    )
  )
```


#plot the data
```{r}
volcano_exudate_soil <- ggplot(
  exudate_vs_soil_plot,
  aes(
    x = log2_fold_change,
    y = minus_log10_p,
    fill = significance
  )
) +
  geom_point(
    shape = 21,
    size = 2,
    alpha = 0.75
  ) +
  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  labs(
    title = "Plant-soil exudate versus soil control",
    subtitle = "Positive-mode LC-MS features",
    x = "Log2 fold change: exudate / soil control",
    y = "-log10 raw p-value",
    fill = "Result"
  ) +
  theme_classic()

volcano_exudate_soil
```
#save the data
```{r}
ggsave(
  "positive_mode_exudate_vs_soil_volcano.png",
  volcano_exudate_soil,
  width = 9,
  height = 7,
  dpi = 300
)
```


#label the top features, selecting for important features using both effect size and p-value
```{r}
top_exudate_features <- exudate_vs_soil_plot %>%
  filter(
    p_value < 0.05,
    abs(log2_fold_change) >= 1
  ) %>%
  arrange(
    p_value,
    desc(abs(log2_fold_change))
  ) %>%
  slice_head(n = 15) %>%
  mutate(
    plot_label = case_when(
      !is.na(`Metabolite name`) &
        `Metabolite name` != "" &
        `Metabolite name` != "Unknown" ~
        paste0(
          `Metabolite name`,
          "\n",
          feature_name
        ),

      TRUE ~ feature_name
    )
  )
```


#plot individual significant features-- volcano plots can hide replicate behavior.

```{r}
strongest_exudate_feature <- exudate_vs_soil %>%
  filter(
    p_value < 0.05,
    log2_fold_change > 0
  ) %>%
  arrange(
    p_value,
    desc(log2_fold_change)
  ) %>%
  slice(1) %>%
  pull(feature_name)

strongest_exudate_feature
```

#create long-format data
```{r}
normalized_long <- as.data.frame(
  pca_matrix_log
) %>%
  rownames_to_column("sample") %>%
  pivot_longer(
    cols = -sample,
    names_to = "feature_name",
    values_to = "log2_normalized_intensity"
  ) %>%
  left_join(
    sample_metadata,
    by = "sample"
  ) %>%
  left_join(
    feature_annotations,
    by = "feature_name"
  )
```


#plot the data
```{r}
normalized_long %>%
  filter(
    feature_name == strongest_exudate_feature,
    group %in% c(
      "Plant-soil exudate",
      "Soil control"
    )
  ) %>%
  ggplot(
    aes(
      x = group,
      y = log2_normalized_intensity,
      fill = group
    )
  ) +
  geom_boxplot(
    width = 0.5,
    outlier.shape = NA,
    alpha = 0.6
  ) +
  geom_jitter(
    width = 0.08,
    size = 3
  ) +
  labs(
    title = strongest_exudate_feature,
    x = NULL,
    y = "Log2 normalized intensity"
  ) +
  theme_classic() +
  theme(
    legend.position = "none"
  )
```

#check the distribution of p-values
```{r}
ggplot(
  exudate_vs_soil,
  aes(x = p_value)
) +
  geom_histogram(
    bins = 30,
    boundary = 0
  ) +
  labs(
    title = "P-value distribution: exudate versus soil",
    x = "Raw p-value",
    y = "Number of features"
  ) +
  theme_classic()
```

#significant feature heatmap without annotations

#select features using exploratory criteria
```{r}
significant_feature_names <- exudate_vs_soil %>%
  filter(
    p_value < 0.05,
    abs(log2_fold_change) >= 1
  ) %>%
  arrange(p_value) %>%
  slice_head(n = 40) %>%
  pull(feature_name)
```

#create the matrix
```{r}
exudate_soil_samples <- sample_metadata %>%
  filter(
    group %in% c(
      "Plant-soil exudate",
      "Soil control"
    )
  ) %>%
  pull(sample)

significant_heatmap_matrix <- t(
  pca_matrix_log[
    exudate_soil_samples,
    significant_feature_names,
    drop = FALSE
  ]
)
```


#run it
```{r}
pheatmap(
  significant_heatmap_matrix,
  scale = "row",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 9,
  main = "Differential features: exudate versus soil"
)
```



#for now, run the exudate-versus-soil code through
```{r}
exudate_vs_soil %>%
  summarise(...)
```






