---
title: "MetaboAnalyst_Pilot Study 1 LCMS"
output: html_document
date: "2026-08-17"
---

```{r}
library(MetaboAnalystR)

packageVersion("MetaboAnalystR")
```


```{r}
library(readxl)
```
#1: Read and convert the MS-DIAL workbooks
```{r}
positive_file <- file.choose()
# Select: postive mode height(2).xlsx

negative_file <- file.choose()
# Select: negative mode height(1).xlsx
```

#extract features and metadata
```{r}
prepare_msdial <- function(file, mode = c("positive", "negative")) {

  mode <- match.arg(mode)

  raw <- readxl::read_excel(
    file,
    col_names = FALSE,
    .name_repair = "minimal"
  )

  # MS-DIAL information rows
  classes   <- as.character(unlist(raw[1, ]))
  filetypes <- as.character(unlist(raw[2, ]))
  orders    <- as.character(unlist(raw[3, ]))
  batches   <- as.character(unlist(raw[4, ]))
  headers   <- as.character(unlist(raw[5, ]))

  # Columns designated by MS-DIAL as samples or blanks
  sample_cols <- which(filetypes %in% c("Sample", "Blank"))

  sample_names <- headers[sample_cols]
  sample_class <- classes[sample_cols]

  # Feature metadata begin after the five MS-DIAL header rows
  feature_metadata <- data.frame(
    Alignment_ID = raw[[1]][-(1:5)],
    RT_min       = as.numeric(raw[[2]][-(1:5)]),
    MZ           = as.numeric(raw[[3]][-(1:5)]),
    Name         = as.character(raw[[4]][-(1:5)]),
    Adduct       = as.character(raw[[5]][-(1:5)]),
    stringsAsFactors = FALSE
  )

  # Make unique and informative feature names
  feature_names <- paste0(
    mode,
    "_ID", feature_metadata$Alignment_ID,
    "_mz", sprintf("%.5f", feature_metadata$MZ),
    "_rt", sprintf("%.3f", feature_metadata$RT_min)
  )

  # Extract feature-intensity matrix
  intensities <- raw[-(1:5), sample_cols, drop = FALSE]
  intensities <- as.data.frame(
    lapply(intensities, as.numeric),
    check.names = FALSE
  )

  colnames(intensities) <- sample_names
  rownames(intensities) <- feature_names

  # Samples in rows, features in columns
  metabo_table <- data.frame(
    Sample = sample_names,
    Class  = sample_class,
    t(as.matrix(intensities)),
    check.names = FALSE
  )

  metadata_table <- data.frame(
    Sample          = sample_names,
    Class           = sample_class,
    File_type       = filetypes[sample_cols],
    Injection_order = as.numeric(orders[sample_cols]),
    Batch           = batches[sample_cols],
    stringsAsFactors = FALSE
  )

  list(
    data = metabo_table,
    features = feature_metadata,
    metadata = metadata_table,
    intensities = intensities
  )
}
```


#2 create positive and negative objects, separating data
```{r}
positive <- prepare_msdial(positive_file, "positive")
negative <- prepare_msdial(negative_file, "negative")
```


#remove confirmed empty columns
```{r}
positive$data <- subset(
  positive$data,
  Sample != "blank_4"
)

positive$metadata <- subset(
  positive$metadata,
  Sample != "blank_4"
)

negative$data <- subset(
  negative$data,
  Sample != "blank_7"
)

negative$metadata <- subset(
  negative$metadata,
  Sample != "blank_7"
)
```

#verify sample tables
```{r}
positive$metadata
negative$metadata

table(positive$data$Class)
table(negative$data$Class)
```

#feature identifiers to the metadata

#Your feature names contain alignment ID, m/z, and retention time, but positive$features and negative$features currently do not contain the constructed feature identifier
```{r}
positive$features$Feature <- colnames(positive$data)[-(1:2)]
negative$features$Feature <- colnames(negative$data)[-(1:2)]
```

#verify. This will allow us to join the eventual p-values and fold changes back to:

*Alignment ID*
*m/z*
*Retention time*
*Adduct*
*Tentative name*

#Without this addition, interpreting the statistical results becomes more difficult.
```{r}
head(positive$features)
head(negative$features)
```


#3: Save untouched MetaboAnalyst input tables; preserve all 5,169 positive and 3,186 negative features while excluding only the two completely empty columns.
```{r}
write.csv(
  positive$data,
  "positive_mode_MetaboAnalyst_unfiltered.csv",
  row.names = FALSE
)


write.csv(
  negative$data,
  "negative_mode_MetaboAnalyst_unfiltered.csv",
  row.names = FALSE
)


```


#4 Before filtering or normalization, let’s assess detection and sample signal. Notes: median intensity among features with an intensity greater than zero.
```{r}
sample_qc <- function(dat) {


  x <- as.data.frame(dat[, -(1:2), drop = FALSE])
  x[] <- lapply(x, as.numeric)


  data.frame(
    Sample = dat$Sample,
    Class = dat$Class,
    Total_signal = rowSums(x, na.rm = TRUE),
    Median_positive = apply(
      x,
      1,
      function(z) median(z[z > 0], na.rm = TRUE)
    ),
    Detected_features = rowSums(x > 0, na.rm = TRUE),
    Zero_percent = 100 * rowMeans(x == 0, na.rm = TRUE)
  )
}


positive_qc <- sample_qc(positive$data)
negative_qc <- sample_qc(negative$data)


positive_qc
negative_qc
```

#Create sample-level QC plots:

```{r}
par(mfrow = c(2, 2), mar = c(9, 4, 3, 1))


barplot(
  positive_qc$Detected_features,
  names.arg = positive_qc$Sample,
  las = 2,
  main = "Positive mode: detected features",
  ylab = "Number of features"
)


barplot(
  log10(positive_qc$Total_signal + 1),
  names.arg = positive_qc$Sample,
  las = 2,
  main = "Positive mode: total signal",
  ylab = "log10 total intensity"
)


barplot(
  negative_qc$Detected_features,
  names.arg = negative_qc$Sample,
  las = 2,
  main = "Negative mode: detected features",
  ylab = "Number of features"
)


barplot(
  log10(negative_qc$Total_signal + 1),
  names.arg = negative_qc$Sample,
  las = 2,
  main = "Negative mode: total signal",
  ylab = "log10 total intensity"
)


par(mfrow = c(1, 1))
```

#Save the QC summaries:
```{r}
write.csv(
  positive_qc,
  "positive_mode_sample_QC.csv",
  row.names = FALSE
)


write.csv(
  negative_qc,
  "negative_mode_sample_QC.csv",
  row.names = FALSE
)
```


```{r}
qc_plot <- function(qc, title_prefix) {

  group_cols <- rainbow(length(unique(qc$Class)))
  names(group_cols) <- unique(qc$Class)

  par(mfrow = c(2, 1), mar = c(10, 5, 3, 1))

  barplot(
    qc$Detected_features,
    names.arg = qc$Sample,
    col = group_cols[qc$Class],
    las = 2,
    cex.names = 0.7,
    main = paste0(title_prefix, ": detected features"),
    ylab = "Features with intensity > 0"
  )

  barplot(
    log10(qc$Total_signal + 1),
    names.arg = qc$Sample,
    col = group_cols[qc$Class],
    las = 2,
    cex.names = 0.7,
    main = paste0(title_prefix, ": total signal"),
    ylab = "log10 total intensity",
    ylim = c(
      min(log10(qc$Total_signal + 1)) - 0.5,
      max(log10(qc$Total_signal + 1)) + 0.5
    )
  )

  par(mfrow = c(1, 1))
}

qc_plot(positive_qc, "Positive mode")
qc_plot(negative_qc, "Negative mode")
```
```{r}
# Close any active plotting device
while (dev.cur() > 1) {
  dev.off()
}
```

#positive mode
```{r}
png(
  "positive_mode_sample_QC.png",
  width = 2400,
  height = 1800,
  res = 200
)

qc_plot(positive_qc, "Positive mode")

dev.off()
```


#negative mode
```{r}
png(
  "negative_mode_sample_QC.png",
  width = 2400,
  height = 1800,
  res = 200
)

qc_plot(negative_qc, "Negative mode")

dev.off()
```
The images will be saved in your working directory. Find that location with:
```{r}
getwd()
```

#To display one inside RStudio:

```{r}
img <- png::readPNG("positive_mode_sample_QC.png")
grid::grid.raster(img)
```

#If png is not installed:

```{r}
install.packages("png")
```


#Alternatively, enlarge the RStudio Plots pane and reduce the margins:

```{r}
par(mar = c(6, 4, 3, 1))
```


#5. Create the comparison-preparation function

```{r}
prepare_comparison <- function(
    dat,
    group1,
    group2,
    min_detected = 2
) {


  # Select only the requested groups
  selected <- dat$Class %in% c(group1, group2)
  comparison <- dat[selected, , drop = FALSE]


  # Keep sample information separate
  sample_info <- comparison[, c("Sample", "Class")]


  # Extract the intensity matrix
  intensities <- comparison[, -(1:2), drop = FALSE]
  intensities[] <- lapply(intensities, as.numeric)


  # Treat MS-DIAL zeros as nondetections
  intensities[intensities == 0] <- NA


  group1_rows <- sample_info$Class == group1
  group2_rows <- sample_info$Class == group2


  # Count detections within each group
  detected_group1 <- colSums(
    !is.na(intensities[group1_rows, , drop = FALSE])
  )


  detected_group2 <- colSums(
    !is.na(intensities[group2_rows, , drop = FALSE])
  )


  # Retain features detected in at least 2/3 replicates
  # in either comparison group
  keep <- detected_group1 >= min_detected |
          detected_group2 >= min_detected


  filtered_intensities <- intensities[, keep, drop = FALSE]


  output <- data.frame(
    Sample = sample_info$Sample,
    Class = sample_info$Class,
    filtered_intensities,
    check.names = FALSE
  )


  detection_summary <- data.frame(
    Feature = names(intensities),
    Group1 = group1,
    Group2 = group2,
    Detected_group1 = detected_group1,
    Detected_group2 = detected_group2,
    Retained = keep,
    stringsAsFactors = FALSE
  )


  list(
    data = output,
    detection = detection_summary
  )
}
```

#6. Prepare positive and negative comparisons

```{r}
fungal_positive <- prepare_comparison(
  positive$data,
  group1 = "FC",
  group2 = "fungi_old",
  min_detected = 2
)


fungal_negative <- prepare_comparison(
  negative$data,
  group1 = "FC",
  group2 = "fungi_old",
  min_detected = 2
)
```

#7. Check what was retained

```{r}
dim(fungal_positive$data)
dim(fungal_negative$data)


table(fungal_positive$data$Class)
table(fungal_negative$data$Class)


table(fungal_positive$detection$Retained)
table(fungal_negative$detection$Retained)
```

#8. Save the comparison tables
```{r}
write.csv(
  fungal_positive$data,
  "positive_fungi_old_vs_FC_prevalence_filtered.csv",
  row.names = FALSE,
  na = ""
)


write.csv(
  fungal_negative$data,
  "negative_fungi_old_vs_FC_prevalence_filtered.csv",
  row.names = FALSE,
  na = ""
)


write.csv(
  fungal_positive$detection,
  "positive_fungi_old_vs_FC_detection_summary.csv",
  row.names = FALSE
)


write.csv(
  fungal_negative$detection,
  "negative_fungi_old_vs_FC_detection_summary.csv",
  row.names = FALSE
)
```

#9. Load the positive comparison into MetaboAnalystR

```{r}

positive_fungal_path <- file.path(
  getwd(),
  "positive_fungi_old_vs_FC_prevalence_filtered.csv"
)

# Confirm that R can locate the CSV
file.exists(positive_fungal_path)
```
```{r}
mSet_pos_fungal <- InitDataObjects(
  "pktable",
  "stat",
  FALSE,
  default.dpi = 72
)
```
```{r}
mSet_pos_fungal <- Read.TextData(
  mSet_pos_fungal,
  positive_fungal_path,
  "rowu",
  "disc"
)
```


```{r}
mSet_pos_fungal <- SanityCheckData(
  mSet_pos_fungal
)
```

#10. Preserve the pre-imputation object. Notes: Imputation means replacing a missing or nondetected measurement with a carefully defined numerical value so statistical calculations can proceed.
```{r}
mSet_pos_fungal_preimpute <- mSet_pos_fungal
```

#create an imputation function/ older version new version mismatch with function.  ReplaceMin
```{r}
impute_one_fifth_min <- function(dat) {

  result <- dat

  feature_cols <- setdiff(
    names(result),
    c("Sample", "Class")
  )

  imputation_report <- data.frame(
    Feature = feature_cols,
    Missing_before = integer(length(feature_cols)),
    Minimum_positive = numeric(length(feature_cols)),
    Imputed_value = numeric(length(feature_cols)),
    stringsAsFactors = FALSE
  )

  for (i in seq_along(feature_cols)) {

    feature <- feature_cols[i]
    values <- as.numeric(result[[feature]])

    missing <- is.na(values) | values == 0
    positive <- values[!missing & values > 0]

    imputation_report$Missing_before[i] <- sum(missing)

    if (length(positive) > 0) {

      minimum_positive <- min(positive)
      replacement <- minimum_positive / 5

      values[missing] <- replacement

      imputation_report$Minimum_positive[i] <- minimum_positive
      imputation_report$Imputed_value[i] <- replacement

    } else {

      imputation_report$Minimum_positive[i] <- NA
      imputation_report$Imputed_value[i] <- NA
    }

    result[[feature]] <- values
  }

  list(
    data = result,
    report = imputation_report
  )
}
```

#11. Replace remaining missing values. # Custom function replaces each nondetection with
# one-fifth of that feature's minimum positive intensity.

```{r}
fungal_positive_imputed <- impute_one_fifth_min(
  fungal_positive$data
)
```

#check operations, results should be zero
```{r}
sum(fungal_positive_imputed$report$Missing_before)

sum(is.na(fungal_positive_imputed$data))

head(fungal_positive_imputed$report)
```
#save imputed table and audit report
```{r}
write.csv(
  fungal_positive_imputed$data,
  "positive_fungi_old_vs_FC_imputed.csv",
  row.names = FALSE
)

write.csv(
  fungal_positive_imputed$report,
  "positive_fungi_old_vs_FC_imputation_report.csv",
  row.names = FALSE
)
```


#12. Reinitialize MetaboAnalystR with the imputed table
```{r}
positive_fungal_imputed_path <- file.path(
  getwd(),
  "positive_fungi_old_vs_FC_imputed.csv"
)

mSet_pos_fungal <- InitDataObjects(
  "pktable",
  "stat",
  FALSE,
  default.dpi = 72
)

mSet_pos_fungal <- Read.TextData(
  mSet_pos_fungal,
  positive_fungal_imputed_path,
  "rowu",
  "disc"
)

mSet_pos_fungal <- SanityCheckData(
  mSet_pos_fungal
)
```

#13
```{r}
mSet_pos_fungal <- PreparePrenormData(
  mSet_pos_fungal
)
```

#Step 14: Connect feature identifiers to MS-DIAL metadata

#This should ideally appear shortly after creating positive and negative, but you can run it now:

```{r}

positive$features$Feature <- colnames(positive$data)[-(1:2)]
negative$features$Feature <- colnames(negative$data)[-(1:2)]


head(positive$features)
head(negative$features)
```

#verify nothing is mismatched
```{r}
nrow(positive$features) ==
  length(colnames(positive$data)[-(1:2)])

nrow(negative$features) ==
  length(colnames(negative$data)[-(1:2)])
```
#Step 15: Normalize the positive fungal comparison
```{r}

mSet_pos_fungal <- Normalization(
  mSetObj = mSet_pos_fungal,
  rowNorm = "NULL",
  transNorm = "LogNorm",
  scaleNorm = "ParetoNorm",
  ref = "NULL",
  ratio = FALSE,
  ratioNum = 20
)
```

#convert data to numeric
```{r}
norm_matrix <- do.call(
  cbind,
  lapply(
    mSet_pos_fungal$dataSet$norm,
    as.numeric
  )
)

colnames(norm_matrix) <- colnames(
  mSet_pos_fungal$dataSet$norm
)

rownames(norm_matrix) <- rownames(
  mSet_pos_fungal$dataSet$norm
)
```

#Step 16: Check the normalized data

```{r}
dim(norm_matrix)

anyNA(norm_matrix)

any(is.infinite(norm_matrix))

summary(as.vector(norm_matrix))
```


#Step 17: Generate normalization QC plots

```{r}
pre_norm_matrix <- as.matrix(
  fungal_positive_imputed$data[, -(1:2)]
)

storage.mode(pre_norm_matrix) <- "numeric"

rownames(pre_norm_matrix) <-
  fungal_positive_imputed$data$Sample
```

#Confirm the files were created:

```{r}
norm_matrix <- do.call(
  cbind,
  lapply(
    mSet_pos_fungal$dataSet$norm,
    as.numeric
  )
)

colnames(norm_matrix) <- colnames(
  mSet_pos_fungal$dataSet$norm
)

rownames(norm_matrix) <- rownames(
  mSet_pos_fungal$dataSet$norm
)
```

#verify that the samples align
```{r}
rownames(pre_norm_matrix)

rownames(norm_matrix)

identical(
  rownames(pre_norm_matrix),
  rownames(norm_matrix)
)
```

#create before NORMALIZATION
```{r positive-fungal-before-normalization, fig.width=9, fig.height=6, dpi=150, fig.align="center", out.width="90%"}

par(mar = c(8, 5, 4, 1))

boxplot(
  lapply(
    seq_len(nrow(pre_norm_matrix)),
    function(i) log2(pre_norm_matrix[i, ])
  ),
  names = rownames(pre_norm_matrix),
  las = 2,
  cex.axis = 0.8,
  col = c(
    rep("lightblue", 3),
    rep("salmon", 3)
  ),
  main = "Before normalization",
  ylab = "Log2 imputed peak height",
  outline = FALSE
)
```
#after normalization

The following plot shows the data after MetaboAnalystR applied logarithmic transformation and Pareto scaling.

```{r positive-fungal-after-normalization, fig.width=9, fig.height=6, dpi=150, fig.align="center", out.width="90%"}

par(mar = c(8, 5, 4, 1))

boxplot(
  lapply(
    seq_len(nrow(norm_matrix)),
    function(i) norm_matrix[i, ]
  ),
  names = rownames(norm_matrix),
  las = 2,
  cex.axis = 0.8,
  col = c(
    rep("lightblue", 3),
    rep("salmon", 3)
  ),
  main = "After log transformation and Pareto scaling",
  ylab = "Normalized intensity",
  outline = FALSE
)

abline(
  h = 0,
  lty = 2,
  col = "gray50"
)

legend(
  "topright",
  legend = c("Fungal control", "Fungi old"),
  fill = c("lightblue", "salmon"),
  bty = "n"
)
```

#create numerical normalization summry

```{r positive-fungal-normalization-summary}

normalization_qc <- data.frame(
  Sample = rownames(pre_norm_matrix),

  Class = fungal_positive_imputed$data$Class,

  Pre_log2_median = apply(
    log2(pre_norm_matrix),
    1,
    median
  ),

  Pre_log2_IQR = apply(
    log2(pre_norm_matrix),
    1,
    IQR
  ),

  Post_median = apply(
    norm_matrix,
    1,
    median
  ),

  Post_IQR = apply(
    norm_matrix,
    1,
    IQR
  )
)

normalization_qc
```

#save it

```{r save-positive-fungal-normalization-summary}

write.csv(
  normalization_qc,
  "positive_fungal_normalization_QC_summary.csv",
  row.names = FALSE
)
```


#Step 18: Connect additional MS-DIAL metadata

```{r}
positive_raw <- readxl::read_excel(
  positive_file,
  col_names = FALSE,
  .name_repair = "minimal"
)


positive$features$Feature <- colnames(
  positive$data
)[-(1:2)]


positive$features$MSMS_assigned <- as.logical(
  positive_raw[[8]][-(1:5)]
)


positive$features$S_N_average <- as.numeric(
  positive_raw[[29]][-(1:5)]
)


positive$features$Post_curation_result <- as.character(
  positive_raw[[6]][-(1:5)]
)


head(positive$features)
```

#Verify alignment:

```{r}
stopifnot(
  nrow(positive$features) ==
    length(colnames(positive$data)[-(1:2)])
)


stopifnot(
  identical(
    positive$features$Feature,
    colnames(positive$data)[-(1:2)]
  )
)
```

#Step 19. Define the differential-analysis function. This function calculates detection counts, group means, log2 fold change, welch's t-test, and Benjamini-Hochberg FDR for every retained feature.

#detection count: number of replicates with an original nonzero measurement
#log2 fold change: difference between the mean log2 intensity in fungi_old vs FC
#Welch's t-test compares group means without assuming equal variances.
#benjamini-Hochberg FDR: corrects the p-values for testing thousands of features simultaneously. Positive log2 fold change= high abundance in fungi_old, neg= higher abundance in FC.

```{r}
calculate_differential <- function(
    original_data,
    imputed_data,
    control = "FC",
    treatment = "fungi_old"
) {

  feature_names <- setdiff(
    names(imputed_data),
    c("Sample", "Class")
  )

  stopifnot(
    identical(
      feature_names,
      setdiff(
        names(original_data),
        c("Sample", "Class")
      )
    )
  )

  safe_mean <- function(x) {

    x <- x[
      !is.na(x) &
      is.finite(x) &
      x > 0
    ]

    if (length(x) == 0) {
      return(NA_real_)
    }

    mean(x)
  }

  safe_welch <- function(
      control_values,
      treatment_values
  ) {

    tryCatch(
      t.test(
        treatment_values,
        control_values,
        paired = FALSE,
        var.equal = FALSE
      )$p.value,
      error = function(e) NA_real_
    )
  }

  results <- lapply(
    feature_names,
    function(feature) {

      original_values <- as.numeric(
        original_data[[feature]]
      )

      imputed_values <- as.numeric(
        imputed_data[[feature]]
      )

      original_control <- original_values[
        original_data$Class == control
      ]

      original_treatment <- original_values[
        original_data$Class == treatment
      ]

      imputed_control <- imputed_values[
        imputed_data$Class == control
      ]

      imputed_treatment <- imputed_values[
        imputed_data$Class == treatment
      ]

      log2_control <- log2(
        imputed_control
      )

      log2_treatment <- log2(
        imputed_treatment
      )

      data.frame(
        Feature = feature,

        Detected_FC = sum(
          !is.na(original_control) &
          original_control > 0
        ),

        Detected_fungi_old = sum(
          !is.na(original_treatment) &
          original_treatment > 0
        ),

        Mean_observed_FC = safe_mean(
          original_control
        ),

        Mean_observed_fungi_old = safe_mean(
          original_treatment
        ),

        Mean_imputed_FC = mean(
          imputed_control
        ),

        Mean_imputed_fungi_old = mean(
          imputed_treatment
        ),

        Log2FC_fungi_vs_FC =
          mean(log2_treatment) -
          mean(log2_control),

        Welch_p_value = safe_welch(
          log2_control,
          log2_treatment
        ),

        stringsAsFactors = FALSE
      )
    }
  )

  results <- do.call(
    rbind,
    results
  )

  results$FDR_BH <- p.adjust(
    results$Welch_p_value,
    method = "BH"
  )

  results
}
```

#step 20: run positive-mode differential analysis

```{r}
positive_fungal_stats <- calculate_differential(
  original_data = fungal_positive$data,
  imputed_data = fungal_positive_imputed$data,
  control = "FC",
  treatment = "fungi_old"
)
```


#check output
```{r}
dim(positive_fungal_stats)

head(positive_fungal_stats)

summary(
  positive_fungal_stats$Log2FC_fungi_vs_FC
)

summary(
  positive_fungal_stats$Welch_p_value
)

summary(
  positive_fungal_stats$FDR_BH
)

sum(
  is.na(
    positive_fungal_stats$Welch_p_value
  )
)
```

#count the initial significant features
```{r}
sum(
  positive_fungal_stats$FDR_BH < 0.05,
  na.rm = TRUE
)

sum(
  positive_fungal_stats$FDR_BH < 0.05 &
    abs(
      positive_fungal_stats$Log2FC_fungi_vs_FC
    ) >= 1,
  na.rm = TRUE
)
```
#step 21: join the statistics to MS-DUAL metadata
```{r}
metadata_match <- match(
  positive_fungal_stats$Feature,
  positive$features$Feature
)

sum(
  is.na(metadata_match)
)
```
#join the tables
```{r}
positive_fungal_results <- cbind(
  positive_fungal_stats,

  positive$features[
    metadata_match,
    c(
      "Alignment_ID",
      "RT_min",
      "MZ",
      "Name",
      "Adduct",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result"
    )
  ]
)
```


#verify the joined tables
```{r}
dim(positive_fungal_results)

head(
  positive_fungal_results[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Adduct",
      "Log2FC_fungi_vs_FC",
      "Welch_p_value",
      "FDR_BH"
    )
  ]
)
```



#flag retention times below one minute
```{r}
positive_fungal_results$RT_below_1_min <-
  positive_fungal_results$RT_min < 1

table(
  positive_fungal_results$RT_below_1_min,
  useNA = "ifany"
)
```
#step 22: add detection-pattern categories
```{r}
positive_fungal_results$Detection_pattern <- with(
  positive_fungal_results,

  ifelse(
    Detected_fungi_old == 3 &
      Detected_FC == 0,

    "Fungi 3/3; FC 0/3",

    ifelse(
      Detected_fungi_old == 3 &
        Detected_FC <= 1,

      "Fungi consistent; FC sparse",

      ifelse(
        Detected_FC == 3 &
          Detected_fungi_old == 0,

        "FC 3/3; fungi 0/3",

        ifelse(
          Detected_FC == 3 &
            Detected_fungi_old == 3,

          "Detected 3/3 in both",

          "Variably detected"
        )
      )
    )
  )
)
```

#count the categories
```{r}
table(
  positive_fungal_results$Detection_pattern
)
```
#step 23: sort and save all differential results
```{r}
positive_fungal_results <- positive_fungal_results[
  order(
    positive_fungal_results$FDR_BH,

    -abs(
      positive_fungal_results$Log2FC_fungi_vs_FC
    ),

    na.last = TRUE
  ),
]
```

#save all tested features
```{r}
write.csv(
  positive_fungal_results,
  "positive_fungi_old_vs_FC_full_differential_results.csv",
  row.names = FALSE
)
```

#confirm file was saved
```{r}
file.exists(
  "positive_fungi_old_vs_FC_full_differential_results.csv"
)
```
#step 24: create the preliminary candidate shortlist. FDR= <0.05, absoloute log2 fold change of at least 1 (at least twofold difference)
```{r}
positive_fungal_candidates <- subset(
  positive_fungal_results,

  !is.na(FDR_BH) &
    FDR_BH < 0.05 &
    abs(Log2FC_fungi_vs_FC) >= 1
)
```

#save candidate list
```{r}
write.csv(
  positive_fungal_candidates,
  "positive_fungi_old_vs_FC_FDR05_FC2_candidates.csv",
  row.names = FALSE
)
```


#summarize results
```{r}
nrow(positive_fungal_results)

nrow(positive_fungal_candidates)

table(
  positive_fungal_candidates$Detection_pattern
)

table(
  positive_fungal_candidates$RT_below_1_min,
  useNA = "ifany"
)

table(
  positive_fungal_candidates$MSMS_assigned,
  useNA = "ifany"
)
```


#display the first 20 candidates
```{r}
head(
  positive_fungal_candidates[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Adduct",
      "Detected_FC",
      "Detected_fungi_old",
      "Mean_observed_FC",
      "Mean_observed_fungi_old",
      "Log2FC_fungi_vs_FC",
      "Welch_p_value",
      "FDR_BH",
      "MSMS_assigned",
      "S_N_average",
      "RT_below_1_min",
      "Detection_pattern"
    )
  ],
  20
)
```

#step 25: display the three FDR-significant features
```{r}
positive_fungal_FDR_significant <- subset(
  positive_fungal_results,
  !is.na(FDR_BH) &
    FDR_BH < 0.05
)

positive_fungal_FDR_significant[
  ,
  c(
    "Feature",
    "Alignment_ID",
    "MZ",
    "RT_min",
    "Adduct",
    "Detected_FC",
    "Detected_fungi_old",
    "Mean_observed_FC",
    "Mean_observed_fungi_old",
    "Log2FC_fungi_vs_FC",
    "Welch_p_value",
    "FDR_BH",
    "MSMS_assigned",
    "S_N_average",
    "RT_below_1_min",
    "Detection_pattern"
  )
]
```
#save without twofold restriction
```{r}
write.csv(
  positive_fungal_FDR_significant,
  "positive_fungi_old_vs_FC_FDR05_all_effect_sizes.csv",
  row.names = FALSE
)
```

#step 26" count nominal and large-effect features
```{r}
sum(
  positive_fungal_results$Welch_p_value < 0.05,
  na.rm = TRUE
)

sum(
  positive_fungal_results$Welch_p_value < 0.05 &
    abs(
      positive_fungal_results$Log2FC_fungi_vs_FC
    ) >= 1,
  na.rm = TRUE
)

sum(
  abs(
    positive_fungal_results$Log2FC_fungi_vs_FC
  ) >= 1,
  na.rm = TRUE
)
```

#step 27: create the reproducible detection list.
```{r}
positive_fungal_detection_candidates <- subset(
  positive_fungal_results,

  Detection_pattern %in% c(
    "Fungi 3/3; FC 0/3",
    "Fungi consistent; FC sparse"
  )
)
```

#sort the list
```{r}
positive_fungal_detection_candidates <-
  positive_fungal_detection_candidates[
    order(
      -positive_fungal_detection_candidates$Detected_fungi_old,
      positive_fungal_detection_candidates$Detected_FC,
      -positive_fungal_detection_candidates$Log2FC_fungi_vs_FC
    ),
  ]
```


#save the list
```{r}
write.csv(
  positive_fungal_detection_candidates,
  "positive_fungi_old_vs_FC_reproducible_detection_candidates.csv",
  row.names = FALSE
)

nrow(
  positive_fungal_detection_candidates
)
```


#list of FDR-supported ion list
```{r}
positive_fungal_FDR_ion_list <-
  positive_fungal_FDR_significant[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Adduct",
      "Detected_FC",
      "Detected_fungi_old",
      "Log2FC_fungi_vs_FC",
      "Welch_p_value",
      "FDR_BH",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result",
      "RT_below_1_min"
    )
  ]

positive_fungal_FDR_ion_list
```



#save it
```{r}
write.csv(
  positive_fungal_FDR_ion_list,
  "positive_fungi_old_vs_FC_FDR_significant_ion_list.csv",
  row.names = FALSE
)
```


#create the reproducible fungal-assosicated ion list
```{r}
positive_fungal_detection_ion_list <-
  positive_fungal_detection_candidates[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Adduct",
      "Detected_FC",
      "Detected_fungi_old",
      "Mean_observed_FC",
      "Mean_observed_fungi_old",
      "Log2FC_fungi_vs_FC",
      "Welch_p_value",
      "FDR_BH",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result",
      "RT_below_1_min",
      "Detection_pattern"
    )
  ]
```

#display first 20
```{r}
head(
  positive_fungal_detection_ion_list,
  20
)
```

#save it
```{r}
write.csv(
  positive_fungal_detection_ion_list,
  "positive_fungi_old_vs_FC_reproducible_ion_list.csv",
  row.names = FALSE
)
```


#count ions with MS/MS spectra
```{r}
table(
  positive_fungal_detection_ion_list$MSMS_assigned,
  useNA = "ifany"
)

sum(
  positive_fungal_detection_ion_list$MSMS_assigned %in% TRUE,
  na.rm = TRUE
)
```

##step 1 verify class names

```{r}
table(positive$data$Class)

stopifnot(
  all(c("root", "leaf") %in% positive$data$Class)
)
```

#step 2 prevalence-folter root versus leaf in 2/3 replicates
```{r}
root_leaf_positive <- prepare_comparison(
  positive$data,
  group1 = "leaf",
  group2 = "root",
  min_detected = 2
)

dim(root_leaf_positive$data)
table(root_leaf_positive$data$Class)
table(root_leaf_positive$detection$Retained)
```

#confrim 3 samples in each group
```{r}
stopifnot(
  all(
    table(root_leaf_positive$data$Class)[c("leaf", "root")] == 3
  )
)
```

#save filtered data
```{r}
write.csv(
  root_leaf_positive$data,
  "positive_root_vs_leaf_prevalence_filtered.csv",
  row.names = FALSE,
  na = ""
)

write.csv(
  root_leaf_positive$detection,
  "positive_root_vs_leaf_detection_summary.csv",
  row.names = FALSE
)
```


#step 3 impute remaining nondetections
```{r}
root_leaf_positive_imputed <- impute_one_fifth_min(
  root_leaf_positive$data
)

sum(root_leaf_positive_imputed$report$Missing_before)
sum(is.na(root_leaf_positive_imputed$data))

head(root_leaf_positive_imputed$report)
```


#save imputed data and audit report
```{r}
write.csv(
  root_leaf_positive_imputed$data,
  "positive_root_vs_leaf_imputed.csv",
  row.names = FALSE
)

write.csv(
  root_leaf_positive_imputed$report,
  "positive_root_vs_leaf_imputation_report.csv",
  row.names = FALSE
)
```


#step 4 load and normalize in MetaboAnalystR
```{r}
positive_root_leaf_path <- file.path(
  getwd(),
  "positive_root_vs_leaf_imputed.csv"
)

stopifnot(file.exists(positive_root_leaf_path))

mSet_pos_root_leaf <- InitDataObjects(
  "pktable",
  "stat",
  FALSE,
  default.dpi = 72
)

mSet_pos_root_leaf <- Read.TextData(
  mSet_pos_root_leaf,
  positive_root_leaf_path,
  "rowu",
  "disc"
)

mSet_pos_root_leaf <- SanityCheckData(
  mSet_pos_root_leaf
)

mSet_pos_root_leaf <- PreparePrenormData(
  mSet_pos_root_leaf
)

mSet_pos_root_leaf <- Normalization(
  mSetObj = mSet_pos_root_leaf,
  rowNorm = "NULL",
  transNorm = "LogNorm",
  scaleNorm = "ParetoNorm",
  ref = "NULL",
  ratio = FALSE,
  ratioNum = 20
)
```

#extract and check the normalized matrix
```{r}
root_leaf_norm_matrix <- do.call(
  cbind,
  lapply(
    mSet_pos_root_leaf$dataSet$norm,
    as.numeric
  )
)

colnames(root_leaf_norm_matrix) <- colnames(
  mSet_pos_root_leaf$dataSet$norm
)

rownames(root_leaf_norm_matrix) <- rownames(
  mSet_pos_root_leaf$dataSet$norm
)

dim(root_leaf_norm_matrix)
anyNA(root_leaf_norm_matrix)
any(is.infinite(root_leaf_norm_matrix))
```

#step 5 use a comparison-safe differential function
```{r}
calculate_differential_general <- function(
    original_data,
    imputed_data,
    reference,
    treatment
) {

  feature_names <- setdiff(
    names(imputed_data),
    c("Sample", "Class")
  )

  stopifnot(
    identical(
      feature_names,
      setdiff(
        names(original_data),
        c("Sample", "Class")
      )
    )
  )

  stopifnot(
    all(c(reference, treatment) %in% original_data$Class)
  )

  safe_mean <- function(x) {

    x <- x[
      !is.na(x) &
        is.finite(x) &
        x > 0
    ]

    if (length(x) == 0) {
      return(NA_real_)
    }

    mean(x)
  }

  safe_welch <- function(reference_values, treatment_values) {

    if (
      length(reference_values) < 2 ||
      length(treatment_values) < 2 ||
      !all(is.finite(reference_values)) ||
      !all(is.finite(treatment_values))
    ) {
      return(NA_real_)
    }

    tryCatch(
      t.test(
        treatment_values,
        reference_values,
        paired = FALSE,
        var.equal = FALSE
      )$p.value,
      error = function(e) NA_real_
    )
  }

  results <- lapply(
    feature_names,
    function(feature) {

      original_values <- as.numeric(
        original_data[[feature]]
      )

      imputed_values <- as.numeric(
        imputed_data[[feature]]
      )

      original_reference <- original_values[
        original_data$Class == reference
      ]

      original_treatment <- original_values[
        original_data$Class == treatment
      ]

      imputed_reference <- imputed_values[
        imputed_data$Class == reference
      ]

      imputed_treatment <- imputed_values[
        imputed_data$Class == treatment
      ]

      log2_reference <- log2(imputed_reference)
      log2_treatment <- log2(imputed_treatment)

      data.frame(
        Feature = feature,

        Detected_reference = sum(
          !is.na(original_reference) &
            original_reference > 0
        ),

        Detected_treatment = sum(
          !is.na(original_treatment) &
            original_treatment > 0
        ),

        Mean_observed_reference = safe_mean(
          original_reference
        ),

        Mean_observed_treatment = safe_mean(
          original_treatment
        ),

        Mean_imputed_reference = mean(
          imputed_reference
        ),

        Mean_imputed_treatment = mean(
          imputed_treatment
        ),

        Log2FC = mean(log2_treatment) -
          mean(log2_reference),

        Welch_p_value = safe_welch(
          log2_reference,
          log2_treatment
        ),

        stringsAsFactors = FALSE
      )
    }
  )

  results <- do.call(rbind, results)

  names(results)[
    names(results) == "Detected_reference"
  ] <- paste0("Detected_", reference)

  names(results)[
    names(results) == "Detected_treatment"
  ] <- paste0("Detected_", treatment)

  names(results)[
    names(results) == "Mean_observed_reference"
  ] <- paste0("Mean_observed_", reference)

  names(results)[
    names(results) == "Mean_observed_treatment"
  ] <- paste0("Mean_observed_", treatment)

  names(results)[
    names(results) == "Mean_imputed_reference"
  ] <- paste0("Mean_imputed_", reference)

  names(results)[
    names(results) == "Mean_imputed_treatment"
  ] <- paste0("Mean_imputed_", treatment)

  names(results)[
    names(results) == "Log2FC"
  ] <- paste0("Log2FC_", treatment, "_vs_", reference)

  results$FDR_BH <- p.adjust(
    results$Welch_p_value,
    method = "BH"
  )

  results
}
```


#step 6 run root versus leaf
```{r}
positive_root_leaf_stats <- calculate_differential_general(
  original_data = root_leaf_positive$data,
  imputed_data = root_leaf_positive_imputed$data,
  reference = "leaf",
  treatment = "root"
)

dim(positive_root_leaf_stats)
head(positive_root_leaf_stats)
```

#check the results
```{r}
summary(
  positive_root_leaf_stats$Log2FC_root_vs_leaf
)

summary(
  positive_root_leaf_stats$Welch_p_value
)

summary(
  positive_root_leaf_stats$FDR_BH
)

sum(
  is.na(positive_root_leaf_stats$Welch_p_value)
)
```


#count statistically supported ions
```{r}
sum(
  positive_root_leaf_stats$FDR_BH < 0.05,
  na.rm = TRUE
)

sum(
  positive_root_leaf_stats$FDR_BH < 0.05 &
    abs(
      positive_root_leaf_stats$Log2FC_root_vs_leaf
    ) >= 1,
  na.rm = TRUE
)
```

#step 7 join the MS-DIAL metadata
```{r}
root_leaf_metadata_match <- match(
  positive_root_leaf_stats$Feature,
  positive$features$Feature
)

stopifnot(
  !any(is.na(root_leaf_metadata_match))
)

positive_root_leaf_results <- cbind(
  positive_root_leaf_stats,

  positive$features[
    root_leaf_metadata_match,
    c(
      "Alignment_ID",
      "RT_min",
      "MZ",
      "Name",
      "Adduct",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result"
    )
  ]
)
```


#flag ions eluting before one minute
```{r}
positive_root_leaf_results$RT_below_1_min <-
  positive_root_leaf_results$RT_min < 1

table(
  positive_root_leaf_results$RT_below_1_min,
  useNA = "ifany"
)
```

#step 8 add detection-pattern categories
```{r}
positive_root_leaf_results$Detection_pattern <- with(
  positive_root_leaf_results,

  ifelse(
    Detected_root == 3 &
      Detected_leaf == 0,

    "Root 3/3; leaf 0/3",

    ifelse(
      Detected_root == 3 &
        Detected_leaf <= 1,

      "Root consistent; leaf sparse",

      ifelse(
        Detected_leaf == 3 &
          Detected_root == 0,

        "Leaf 3/3; root 0/3",

        ifelse(
          Detected_leaf == 3 &
            Detected_root == 3,

          "Detected 3/3 in both",

          "Variably detected"
        )
      )
    )
  )
)

table(
  positive_root_leaf_results$Detection_pattern
)
```
#step 9 sort and save all tested ions
```{r}
positive_root_leaf_results <-
  positive_root_leaf_results[
    order(
      positive_root_leaf_results$FDR_BH,
      -abs(
        positive_root_leaf_results$Log2FC_root_vs_leaf
      ),
      na.last = TRUE
    ),
  ]

write.csv(
  positive_root_leaf_results,
  "positive_root_vs_leaf_full_differential_results.csv",
  row.names = FALSE
)
```


#step 10 prioritu identification list
```{r}
positive_root_leaf_candidates <- subset(
  positive_root_leaf_results,

  !is.na(FDR_BH) &
    FDR_BH < 0.05 &
    abs(Log2FC_root_vs_leaf) >= 1
)

write.csv(
  positive_root_leaf_candidates,
  "positive_root_vs_leaf_FDR05_FC2_candidates.csv",
  row.names = FALSE
)

nrow(positive_root_leaf_candidates)
```
#create concise identification-ready list
```{r}
positive_root_leaf_ion_list <-
  positive_root_leaf_candidates[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Name",
      "Adduct",
      "Detected_leaf",
      "Detected_root",
      "Mean_observed_leaf",
      "Mean_observed_root",
      "Log2FC_root_vs_leaf",
      "Welch_p_value",
      "FDR_BH",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result",
      "RT_below_1_min",
      "Detection_pattern"
    )
  ]

head(
  positive_root_leaf_ion_list,
  20
)

write.csv(
  positive_root_leaf_ion_list,
  "positive_root_vs_leaf_priority_ion_list.csv",
  row.names = FALSE
)
```

#step 11 save all FDR-supported ions without the fold change restriction
```{r}
positive_root_leaf_FDR_significant <- subset(
  positive_root_leaf_results,

  !is.na(FDR_BH) &
    FDR_BH < 0.05
)

write.csv(
  positive_root_leaf_FDR_significant,
  "positive_root_vs_leaf_FDR05_all_effect_sizes.csv",
  row.names = FALSE
)

nrow(positive_root_leaf_FDR_significant)
```

#turface with roots, turface without roots


#step 1: verify and rename the groups
```{r}
table(positive$data$Class)

stopifnot(
  all(
    c("exudate", "Soil_control") %in%
      positive$data$Class
  )
)
```
#display samples that will be used
```{r}
positive$metadata[
  positive$metadata$Class %in%
    c("exudate", "Soil_control"),
  c("Sample", "Class")
]
```

#Create the comparison using the original class names:
```{r}
turface_positive <- prepare_comparison(
  positive$data,
  group1 = "Soil_control",
  group2 = "exudate",
  min_detected = 2
)
```

#Rename the classes:

```{r}
turface_positive$data$Class[
  turface_positive$data$Class == "Soil_control"
] <- "Turface_without_root"


turface_positive$data$Class[
  turface_positive$data$Class == "exudate"
] <- "Turface_with_root"
```

#Also update the labels in the detection summary:

```{r}
turface_positive$detection$Group1[
  turface_positive$detection$Group1 == "Soil_control"
] <- "Turface_without_root"


turface_positive$detection$Group2[
  turface_positive$detection$Group2 == "exudate"
] <- "Turface_with_root"
```


#Verify the renamed groups:
```{r}
table(turface_positive$data$Class)


head(
  turface_positive$data[
    ,
    c("Sample", "Class")
  ]
)
```

#Confirm that there are three replicates per group:
```{r}
stopifnot(
  all(
    table(turface_positive$data$Class)[
      c(
        "Turface_without_root",
        "Turface_with_root"
      )
    ] == 3
  )
)
```

#Step 2: Examine prevalence filtering

```{r}
dim(turface_positive$data)


table(
  turface_positive$detection$Retained
)


sum(
  turface_positive$detection$Retained
)
```

#Save the prevalence-filtered data:
```{r}
write.csv(
  turface_positive$data,
  "positive_Turface_with_root_vs_without_root_prevalence_filtered.csv",
  row.names = FALSE,
  na = ""
)


write.csv(
  turface_positive$detection,
  "positive_Turface_with_root_vs_without_root_detection_summary.csv",
  row.names = FALSE
)
```

#Confirm that the files were created:

```{r}
file.exists(
  "positive_Turface_with_root_vs_without_root_prevalence_filtered.csv"
)


file.exists(
  "positive_Turface_with_root_vs_without_root_detection_summary.csv"
)
```

#Step 3: Impute nondetected values. Use the same one-fifth minimum imputation applied to the earlier comparisons:

```{r}
turface_positive_imputed <- impute_one_fifth_min(
  turface_positive$data
)
```

#Check the imputation:

```{r}
sum(
  turface_positive_imputed$report$Missing_before
)


sum(
  is.na(turface_positive_imputed$data)
)


head(
  turface_positive_imputed$report
)
```

#Save the imputed data and imputation audit report:

```{r}
write.csv(
  turface_positive_imputed$data,
  "positive_Turface_with_root_vs_without_root_imputed.csv",
  row.names = FALSE
)


write.csv(
  turface_positive_imputed$report,
  "positive_Turface_with_root_vs_without_root_imputation_report.csv",
  row.names = FALSE
)
```

#Step 4: Load the comparison into MetaboAnalystR

```{r}
positive_turface_path <- file.path(
  getwd(),
  "positive_Turface_with_root_vs_without_root_imputed.csv"
)


stopifnot(
  file.exists(positive_turface_path)
)
```

#Initialize the MetaboAnalyst object:

```{r}
mSet_pos_turface <- InitDataObjects(
  "pktable",
  "stat",
  FALSE,
  default.dpi = 72
)
```

#Read the data:

```{r}
mSet_pos_turface <- Read.TextData(
  mSet_pos_turface,
  positive_turface_path,
  "rowu",
  "disc"
)
```

#Check and prepare the data:

```{r}
mSet_pos_turface <- SanityCheckData(
  mSet_pos_turface
)


mSet_pos_turface <- PreparePrenormData(
  mSet_pos_turface
)
```

#Step 5: Normalize the data. Use the same normalization settings as the fungal and root-versus-leaf comparisons:


```{r}
mSet_pos_turface <- Normalization(
  mSetObj = mSet_pos_turface,
  rowNorm = "NULL",
  transNorm = "LogNorm",
  scaleNorm = "ParetoNorm",
  ref = "NULL",
  ratio = FALSE,
  ratioNum = 20
)
```

#Extract the normalized matrix:

```{r}
turface_norm_matrix <- do.call(
  cbind,
  lapply(
    mSet_pos_turface$dataSet$norm,
    as.numeric
  )
)


colnames(turface_norm_matrix) <- colnames(
  mSet_pos_turface$dataSet$norm
)


rownames(turface_norm_matrix) <- rownames(
  mSet_pos_turface$dataSet$norm
)
```

#Check the normalized data:

```{r}
dim(turface_norm_matrix)


anyNA(turface_norm_matrix)


any(
  is.infinite(turface_norm_matrix)
)


summary(
  as.vector(turface_norm_matrix)
)
```

#Step 6: Create before- and after-normalization matrices

```{r}
turface_pre_norm_matrix <- as.matrix(
  turface_positive_imputed$data[
    ,
    -(1:2)
  ]
)


storage.mode(turface_pre_norm_matrix) <- "numeric"


rownames(turface_pre_norm_matrix) <-
  turface_positive_imputed$data$Sample
```

#Confirm sample alignment:

```{r}
rownames(turface_pre_norm_matrix)


rownames(turface_norm_matrix)


identical(
  rownames(turface_pre_norm_matrix),
  rownames(turface_norm_matrix)
)
```

#Step 7: Plot before normalization

```{r}
par(
  mar = c(8, 5, 4, 1)
)


turface_colors <- ifelse(
  turface_positive_imputed$data$Class ==
    "Turface_without_root",
  "lightblue",
  "darkseagreen3"
)


boxplot(
  lapply(
    seq_len(
      nrow(turface_pre_norm_matrix)
    ),
    function(i) {
      log2(
        turface_pre_norm_matrix[i, ]
      )
    }
  ),
  names = rownames(
    turface_pre_norm_matrix
  ),
  las = 2,
  cex.axis = 0.8,
  col = turface_colors,
  main = paste(
    "Turface with root vs Turface without root:",
    "before normalization"
  ),
  ylab = "Log2 imputed peak height",
  outline = FALSE
)


legend(
  "topright",
  legend = c(
    "Turface without root",
    "Turface with root"
  ),
  fill = c(
    "lightblue",
    "darkseagreen3"
  ),
  bty = "n"
)
```

#Step 8: Plot after normalization

```{r}
par(
  mar = c(8, 5, 4, 1)
)


boxplot(
  lapply(
    seq_len(
      nrow(turface_norm_matrix)
    ),
    function(i) {
      turface_norm_matrix[i, ]
    }
  ),
  names = rownames(
    turface_norm_matrix
  ),
  las = 2,
  cex.axis = 0.8,
  col = turface_colors,
  main = paste(
    "Turface with root vs Turface without root:",
    "after normalization"
  ),
  ylab = "Normalized intensity",
  outline = FALSE
)


abline(
  h = 0,
  lty = 2,
  col = "gray50"
)


legend(
  "topright",
  legend = c(
    "Turface without root",
    "Turface with root"
  ),
  fill = c(
    "lightblue",
    "darkseagreen3"
  ),
  bty = "n"
)
```

#Step 9: Create the normalization QC summary

```{r}
turface_normalization_qc <- data.frame(
  Sample = rownames(
    turface_pre_norm_matrix
  ),


  Class = turface_positive_imputed$data$Class,


  Pre_log2_median = apply(
    log2(turface_pre_norm_matrix),
    1,
    median
  ),


  Pre_log2_IQR = apply(
    log2(turface_pre_norm_matrix),
    1,
    IQR
  ),


  Post_median = apply(
    turface_norm_matrix,
    1,
    median
  ),


  Post_IQR = apply(
    turface_norm_matrix,
    1,
    IQR
  )
)


turface_normalization_qc
```

#Save it:

```{r}
write.csv(
  turface_normalization_qc,
  "positive_Turface_with_root_vs_without_root_normalization_QC.csv",
  row.names = FALSE
)
```

#Step 10: Run the differential-abundance analysis. Set Turface without root as the reference and Turface with root as the treatment:
```{r}
positive_turface_stats <-
  calculate_differential_general(
    original_data =
      turface_positive$data,


    imputed_data =
      turface_positive_imputed$data,


    reference =
      "Turface_without_root",


    treatment =
      "Turface_with_root"
  )
```

#Check the results:
```{r}
dim(
  positive_turface_stats
)


head(
  positive_turface_stats
)


summary(
  positive_turface_stats$
    Log2FC_Turface_with_root_vs_Turface_without_root
)


summary(
  positive_turface_stats$Welch_p_value
)


summary(
  positive_turface_stats$FDR_BH
)


sum(
  is.na(
    positive_turface_stats$Welch_p_value
  )
)
```

#Count FDR-supported ions:

```{r}
sum(
  positive_turface_stats$FDR_BH < 0.05,
  na.rm = TRUE
)
```

#Count FDR-supported ions with at least a twofold difference:

```{r}
sum(
  positive_turface_stats$FDR_BH < 0.05 &
    abs(
      positive_turface_stats$
        Log2FC_Turface_with_root_vs_Turface_without_root
    ) >= 1,
  na.rm = TRUE
)
```

#Step 11: Join the statistics to the MS-DIAL metadata

```{r}
turface_metadata_match <- match(
  positive_turface_stats$Feature,
  positive$features$Feature
)


sum(
  is.na(turface_metadata_match)
)


stopifnot(
  !any(
    is.na(turface_metadata_match)
  )
)
```

#Join the tables:

```{r}
positive_turface_results <- cbind(
  positive_turface_stats,


  positive$features[
    turface_metadata_match,
    c(
      "Alignment_ID",
      "RT_min",
      "MZ",
      "Name",
      "Adduct",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result"
    )
  ]
)
```

#Verify the joined table:

```{r}
dim(
  positive_turface_results
)


head(
  positive_turface_results[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Adduct",
      "Log2FC_Turface_with_root_vs_Turface_without_root",
      "Welch_p_value",
      "FDR_BH"
    )
  ]
)
```

#Step 12: Flag retention times below one minute

```{r}
positive_turface_results$RT_below_1_min <-
  positive_turface_results$RT_min < 1


table(
  positive_turface_results$RT_below_1_min,
  useNA = "ifany"
)
```

#Step 13: Add abundance direction

```{r}
positive_turface_results$Abundance_direction <- with(
  positive_turface_results,


  ifelse(
    Log2FC_Turface_with_root_vs_Turface_without_root > 0,


    "Higher in Turface with root",


    ifelse(
      Log2FC_Turface_with_root_vs_Turface_without_root < 0,


      "Higher in Turface without root",


      "No change"
    )
  )
)
```

#Add an ordinary root-containing/root-free abundance ratio:
```{r}
positive_turface_results$
  Fold_change_with_root_over_without_root <-
  2^positive_turface_results$
    Log2FC_Turface_with_root_vs_Turface_without_root
```

#Add the fold-change magnitude:

```{r}
positive_turface_results$Fold_change_magnitude <-
  2^abs(
    positive_turface_results$
      Log2FC_Turface_with_root_vs_Turface_without_root
  )
```

#Check the direction counts:

```{r}
table(
  positive_turface_results$Abundance_direction
)
```

# Positive Log2FC:
# higher in Turface with root


# Negative Log2FC:
# higher in Turface without root


#Step 14: Add detection-pattern categories

```{r}
positive_turface_results$Detection_pattern <- with(
  positive_turface_results,


  ifelse(
    Detected_Turface_with_root == 3 &
      Detected_Turface_without_root == 0,


    "With root 3/3; without root 0/3",


    ifelse(
      Detected_Turface_with_root == 3 &
        Detected_Turface_without_root <= 1,


      "With root consistent; without root sparse",


      ifelse(
        Detected_Turface_without_root == 3 &
          Detected_Turface_with_root == 0,


        "Without root 3/3; with root 0/3",


        ifelse(
          Detected_Turface_without_root == 3 &
            Detected_Turface_with_root == 3,


          "Detected 3/3 in both",


          "Variably detected"
        )
      )
    )
  )
)
```

#Count the categories:


```{r}
table(
  positive_turface_results$Detection_pattern
)
```

Step 15: Sort and save all differential results


```{r}
positive_turface_results <-
  positive_turface_results[
    order(
      positive_turface_results$FDR_BH,
      -abs(
        positive_turface_results$
          Log2FC_Turface_with_root_vs_Turface_without_root
      ),


      na.last = TRUE
    ),
  ]
```

#Save all tested ions:

```{r}
write.csv(
  positive_turface_results,
  "positive_Turface_with_root_vs_without_root_full_differential_results.csv",
  row.names = FALSE
)
```

#Step 16: Create the FDR-supported list

This list includes every ion with FDR < 0.05, regardless of fold-change magnitude:
```{r}
positive_turface_FDR_significant <- subset(
  positive_turface_results,


  !is.na(FDR_BH) &
    FDR_BH < 0.05
)


nrow(
  positive_turface_FDR_significant
)
```

#Save it:

```{r}
write.csv(
  positive_turface_FDR_significant,
  "positive_Turface_with_root_vs_without_root_FDR05_all_effect_sizes.csv",
  row.names = FALSE
)
```

#Step 17: Create the strict priority list. Require:FDR < 0.05 absolute log₂ fold change of at least 1
```{r}
positive_turface_candidates <- subset(
  positive_turface_results,


  !is.na(FDR_BH) &
    FDR_BH < 0.05 &
    abs(
      Log2FC_Turface_with_root_vs_Turface_without_root
    ) >= 1
)
```

#Count the candidates:

```{r}
nrow(
  positive_turface_candidates
)


table(
  positive_turface_candidates$Abundance_direction
)


table(
  positive_turface_candidates$RT_below_1_min,
  useNA = "ifany"
)


table(
  positive_turface_candidates$MSMS_assigned,
  useNA = "ifany"
)
```

#Step 18: Create the identification-ready ion list

```{r}
positive_turface_ion_list <-
  positive_turface_candidates[
    ,
    c(
      "Feature",
      "Alignment_ID",
      "MZ",
      "RT_min",
      "Name",
      "Adduct",
      "Detected_Turface_without_root",
      "Detected_Turface_with_root",
      "Mean_observed_Turface_without_root",
      "Mean_observed_Turface_with_root",
      "Log2FC_Turface_with_root_vs_Turface_without_root",
      "Abundance_direction",
      "Fold_change_with_root_over_without_root",
      "Fold_change_magnitude",
      "Welch_p_value",
      "FDR_BH",
      "MSMS_assigned",
      "S_N_average",
      "Post_curation_result",
      "RT_below_1_min",
      "Detection_pattern"
    )
  ]
```


#Display the first 20 candidates:

```{r}
head(
  positive_turface_ion_list,
  20
)
```

#save priority list
```{r}
write.csv(
  positive_turface_ion_list,
  "positive_Turface_with_root_vs_without_root_priority_ion_list.csv",
  row.names = FALSE
)
```

#Step 19: Create a supplementary reproducible-detection list. This captures ions consistently detected in Turface with root but absent or sparse in Turface without root:
```{r}
positive_turface_detection_candidates <- subset(
  positive_turface_results,


  Detection_pattern %in% c(
    "With root 3/3; without root 0/3",
    "With root consistent; without root sparse"
  )
)
```

#Sort the list:

```{r}
positive_turface_detection_candidates <-
  positive_turface_detection_candidates[
    order(
      -positive_turface_detection_candidates$
        Detected_Turface_with_root,


      positive_turface_detection_candidates$
        Detected_Turface_without_root,


      -positive_turface_detection_candidates$
        Log2FC_Turface_with_root_vs_Turface_without_root
    ),
  ]
```

#Save it:

```{r}
write.csv(
  positive_turface_detection_candidates,
  "positive_Turface_with_root_vs_without_root_reproducible_detection_list.csv",
  row.names = FALSE
)


nrow(
  positive_turface_detection_candidates
)
```

#Step 20: Final result summary

```{r}
turface_analysis_summary <- data.frame(
  Comparison =
    "Turface with root vs Turface without root",


  Prevalence_filtered_ions =
    nrow(positive_turface_results),


  Raw_p_below_0_05 =
    sum(
      positive_turface_results$Welch_p_value < 0.05,
      na.rm = TRUE
    ),


  FDR_below_0_05 =
    sum(
      positive_turface_results$FDR_BH < 0.05,
      na.rm = TRUE
    ),


  FDR_below_0_05_and_twofold =
    sum(
      positive_turface_results$FDR_BH < 0.05 &
        abs(
          positive_turface_results$
            Log2FC_Turface_with_root_vs_Turface_without_root
        ) >= 1,
      na.rm = TRUE
    ),


  Reproducible_with_root_detection =
    nrow(
      positive_turface_detection_candidates
    )
)


turface_analysis_summary
```

#Save the summary:

```{r}
write.csv(
  turface_analysis_summary,
  "positive_Turface_with_root_vs_without_root_analysis_summary.csv",
  row.names = FALSE
)
```















































































































































































