---
title: "Pilot Study 2 LC-MS Data"
output: html_document
date: "2026-08-10"
---


1. Packages

Don't repeatedly run install.packages() every time you knit the document. Install once in the Console if needed:

```{r}
install.packages(c(
  "readxl",
  "dplyr",
  "tidyr",
  "stringr",
  "ggplot2",
  "ggrepel",
  "tibble",
  "pheatmap"
))
```



Then your R Markdown only needs:

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(tibble)
library(pheatmap)
```


2. Import positive-mode MS-DIAL data

```{r}
file_path <- "C:/Users/norba/Downloads/postive mode height.xlsx"

msdial <- read_excel(
  path = file_path,
  sheet = "postive mode height",
  skip = 4
)

dim(msdial)
names(msdial)
```


3. Identify sample-intensity columns

These are the columns containing the actual MS-DIAL feature peak heights.

```{r}
sample_columns <- c(
  "blank_1", "blank_2", "blank_3", "blank_4",
  "exudate_1", "exudate_2", "exudate_3",
  "FC_1", "FC_2", "FC_3",
  "fungi_old_1", "fungi_old_2", "fungi_old_3",
  "leaf_1", "leaf_2", "leaf_3",
  "new_in_P08", "new_in_P09", "new_in_P10",
  "new_out_P08", "new_out_P09", "new_out_P10",
  "root_1", "root_2", "root_3",
  "Soil_control_1", "Soil_control_2", "Soil_control_3"
)
```



Safety check:

```{r}
stopifnot(
  all(sample_columns %in% names(msdial))
)
```



There should be 28 injections total:

```{r}
length(sample_columns)

```


4. Make sure intensities are numeric

```{r}
msdial <- msdial %>%
  mutate(
    across(
      all_of(sample_columns),
      as.numeric
    )
  )
```


Check:

```{r}
sapply(
  msdial[, sample_columns],
  class
)
```


These should be numeric.

5. Create sample metadata

This tells R what each sample actually represents biologically.

```{r}
sample_metadata <- tibble(
  sample = sample_columns
) %>%
  mutate(

    group = case_when(
      str_detect(sample, "^blank") ~
        "Blank",

      str_detect(sample, "^exudate") ~
        "Plant-soil exudate",

      str_detect(sample, "^Soil_control") ~
        "Soil control",

      str_detect(sample, "^FC") ~
        "Fungal control",

      str_detect(sample, "^fungi_old") ~
        "Fungi old",

      str_detect(sample, "^leaf") ~
        "Leaf",

      str_detect(sample, "^root") ~
        "Root",

      str_detect(sample, "^new_in") ~
        "Fungal agar in",

      str_detect(sample, "^new_out") ~
        "Fungal agar out",

      TRUE ~ "Unknown"
    ),

    replicate = case_when(
      str_detect(sample, "P08") ~ "1",
      str_detect(sample, "P09") ~ "2",
      str_detect(sample, "P10") ~ "3",

      TRUE ~
        str_extract(sample, "\\d+$")
    ),

    broad_matrix = case_when(

      group %in% c(
        "Plant-soil exudate",
        "Soil control"
      ) ~ "Soil extract",

      group %in% c(
        "Fungal control",
        "Fungi old",
        "Fungal agar in",
        "Fungal agar out"
      ) ~ "Fungal agar",

      group %in% c(
        "Leaf",
        "Root"
      ) ~ "Plant tissue",

      group == "Blank" ~
        "Blank",

      TRUE ~ "Unknown"
    )
  )

```


Check it:

```{r}
sample_metadata
```



And:

```{r}
sample_metadata %>%
  count(group)
```


You expect eight biological groups with three replicates each, plus four blanks. Your original metadata assignment already establishes these groups.

6. Initial sample-level QC

Here we haven't filtered anything yet.

Calculate:

number of features with nonzero peak height;
sum of detected MS-DIAL feature peak heights.
sum of MS-DIAL feature peak heights, not the raw LC-MS total ion chromatogram.

```{r}
sample_qc <- tibble(

  sample = sample_columns,

  summed_feature_intensity = colSums(
    as.matrix(
      msdial[, sample_columns]
    ),
    na.rm = TRUE
  ),

  detected_features = colSums(
    as.matrix(
      msdial[, sample_columns]
    ) > 0,
    na.rm = TRUE
  )

) %>%
  left_join(
    sample_metadata,
    by = "sample"
  )

sample_qc
```




7. Find failed/empty injections
```{r}
sample_qc %>%
  filter(
    detected_features == 0 |
      summed_feature_intensity == 0
  )
```




8. Plot summed feature signal
```{r}
signal_plot <- ggplot(
  sample_qc,
  aes(
    x = reorder(
      sample,
      summed_feature_intensity
    ),
    y = summed_feature_intensity,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title =
      "Sum of detected positive-mode LC-MS feature intensities",
    x = NULL,
    y =
      "Sum of MS-DIAL feature peak heights",
    fill = "Sample group"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )

signal_plot
```


9. Plot number of detected features

```{r}
feature_count_plot <- ggplot(
  sample_qc,
  aes(
    x = reorder(
      sample,
      detected_features
    ),
    y = detected_features,
    fill = group
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title =
      "Detected positive-mode LC-MS features per sample",
    x = NULL,
    y =
      "Features with peak height > 0",
    fill =
      "Sample group"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )

feature_count_plot
```


10. Calculate replicate QC and CV
```{r}
group_qc_summary <- sample_qc %>%
  filter(
    sample != "blank_4"
  ) %>%
  group_by(group) %>%
  summarise(

    n = n(),

    mean_features =
      mean(detected_features),

    sd_features =
      sd(detected_features),

    feature_CV_percent =
      100 *
      sd(detected_features) /
      mean(detected_features),

    mean_summed_signal =
      mean(summed_feature_intensity),

    sd_summed_signal =
      sd(summed_feature_intensity),

    signal_CV_percent =
      100 *
      sd(summed_feature_intensity) /
      mean(summed_feature_intensity),

    .groups = "drop"
  )

group_qc_summary
```



11. Build the quantitative feature table

Now we shift from:

What does each sample look like overall?

to:

What happened to each individual LC-MS feature?

I would keep more annotation information now because we'll need it later.

```{r}
feature_table <- msdial %>%
  select(

    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `Fill %`,
    `MS/MS assigned`,

    any_of(c(
      "Reference RT",
      "Reference m/z",
      "Formula",
      "Ontology",
      "INCHIKEY",
      "SMILES",
      "Annotation tag (VS1.0)",
      "RT matched",
      "m/z matched",
      "MS/MS matched",
      "Total score",
      "RT similarity",
      "Dot product",
      "Reverse dot product",
      "Fragment presence %",
      "S/N average"
    )),

    all_of(sample_columns)
  )
```


12. Give every feature a readable unique name
```{r}
feature_table <- feature_table %>%
  mutate(

    feature_name = paste0(

      "F",
      `Alignment ID`,

      "_mz",
      round(
        `Average Mz`,
        4
      ),

      "_rt",
      round(
        `Average Rt(min)`,
        2
      )
    )
  )
```



Example:

F1540_mz187.1078_rt0.74

means:

MS-DIAL Alignment ID = 1540
m/z ≈ 187.1078
RT ≈ 0.74 min


13. Define usable blanks and biological samples
```{r}
valid_blank_columns <- c(
  "blank_1",
  "blank_2",
  "blank_3"
)

biological_columns <- setdiff(
  sample_columns,
  c(
    valid_blank_columns,
    "blank_4"
  )
)
```



Check:

```{r}
length(biological_columns)
```



14. Calculate blank/background information for every feature

```{r}
feature_qc <- feature_table %>%
  mutate(

    blank_mean = rowMeans(
      across(
        all_of(valid_blank_columns)
      ),
      na.rm = TRUE
    ),

    maximum_biological_signal = apply(
      as.matrix(
        pick(
          all_of(biological_columns)
        )
      ),
      1,
      max,
      na.rm = TRUE
    ),

    sample_to_blank_ratio =
      (
        maximum_biological_signal + 1
      ) /
      (
        blank_mean + 1
      )
  )
```


15. Examine different blank thresholds

Before filtering, save a copy:

```{r}
feature_qc_unfiltered <- feature_qc
```




Then compare 3×, 5× and 10×:

```{r}
blank_filter_summary <- tibble(

  threshold = c(
    "3x",
    "5x",
    "10x"
  ),

  retained_features = c(

    sum(
      feature_qc_unfiltered$
        sample_to_blank_ratio >= 3,
      na.rm = TRUE
    ),

    sum(
      feature_qc_unfiltered$
        sample_to_blank_ratio >= 5,
      na.rm = TRUE
    ),

    sum(
      feature_qc_unfiltered$
        sample_to_blank_ratio >= 10,
      na.rm = TRUE
    )
  )
)

blank_filter_summary

```


For our main exploratory analysis:

```{r}
feature_qc <- feature_qc_unfiltered %>%
  filter(
    sample_to_blank_ratio >= 5
  )
```



This preserves the 5× choice you've already used.

16. Group-aware prevalence filter

```{r}
group_columns <- list(

  exudate = c(
    "exudate_1",
    "exudate_2",
    "exudate_3"
  ),

  fungal_control = c(
    "FC_1",
    "FC_2",
    "FC_3"
  ),

  fungi_old = c(
    "fungi_old_1",
    "fungi_old_2",
    "fungi_old_3"
  ),

  leaf = c(
    "leaf_1",
    "leaf_2",
    "leaf_3"
  ),

  fungal_agar_in = c(
    "new_in_P08",
    "new_in_P09",
    "new_in_P10"
  ),

  fungal_agar_out = c(
    "new_out_P08",
    "new_out_P09",
    "new_out_P10"
  ),

  root = c(
    "root_1",
    "root_2",
    "root_3"
  ),

  soil_control = c(
    "Soil_control_1",
    "Soil_control_2",
    "Soil_control_3"
  )
)

```

The group-specific rule is better.


Calculate detection:

```{r}
presence_by_group <- sapply(

  group_columns,

  function(cols) {

    rowSums(
      as.matrix(
        feature_qc[, cols]
      ) > 0,
      na.rm = TRUE
    ) >= 2
  }
)
```



Keep features detected in at least 2 of 3 replicates in at least one biological group:

```{r}
feature_qc$present_in_any_group <-
  apply(
    presence_by_group,
    1,
    any
  )

feature_qc <- feature_qc %>%
  filter(
    present_in_any_group
  )
```



Now calculate overall detection count just for reporting:

```{r}
feature_qc <- feature_qc %>%
  mutate(

    number_detected = rowSums(

      across(
        all_of(biological_columns),
        ~ .x > 0
      ),

      na.rm = TRUE
    )
  )
```



Check filtering:

```{r}
nrow(feature_table)
nrow(feature_qc)
```



17. Build the PCA intensity matrix


```{r}
intensity_matrix <- feature_qc %>%
  select(
    feature_name,
    all_of(biological_columns)
  ) %>%
  column_to_rownames(
    "feature_name"
  ) %>%
  as.matrix()

```


Transpose:

```{r}
pca_matrix_raw <- t(
  intensity_matrix
)

dim(pca_matrix_raw)
```



Now:

rows = samples
columns = LC-MS features.

18. Replace zeros

For PCA only, replace zeros/missing values with half of that feature's smallest positive measurement.
```{r}
replace_zero_half_min <- function(x) {

  positive_values <-
    x[
      x > 0 &
        is.finite(x)
    ]

  if (
    length(positive_values) == 0
  ) {

    return(
      rep(
        NA_real_,
        length(x)
      )
    )
  }

  replacement <-
    min(positive_values) / 2

  x[
    x == 0 |
      is.na(x)
  ] <- replacement

  x
}

```


Apply:

```{r}
pca_matrix_imputed <- apply(
  pca_matrix_raw,
  2,
  replace_zero_half_min
)
```



Remove anything that still cannot be analyzed:

```{r}
pca_matrix_imputed <-
  pca_matrix_imputed[
    ,
    colSums(
      is.na(
        pca_matrix_imputed
      )
    ) == 0,
    drop = FALSE
  ]
```


19. Median normalize

We're choosing one normalization method for this pipeline.

Your previous code calculated median normalization and TIC normalization but then only used median normalization.

So remove the unused TIC block.

```{r}
sample_medians <- apply(
  pca_matrix_imputed,
  1,
  median,
  na.rm = TRUE
)

global_median <-
  median(
    sample_medians
  )

pca_matrix_normalized <- sweep(
  pca_matrix_imputed,
  1,
  sample_medians,
  "/"
) * global_median
```


20. Log2 transformation

```{r}
pca_matrix_log <- log2(
  pca_matrix_normalized + 1
)
```


21. Remove zero-variance features

```{r}
feature_variance <- apply(
  pca_matrix_log,
  2,
  var,
  na.rm = TRUE
)

keep_variable_features <-
  is.finite(
    feature_variance
  ) &
  feature_variance > 0

pca_matrix_log <-
  pca_matrix_log[
    ,
    keep_variable_features,
    drop = FALSE
  ]
```


22. Run broad PCA

```{r}
pca_result <- prcomp(
  pca_matrix_log,
  center = TRUE,
  scale. = TRUE
)

summary(pca_result)
```


23. Variance explained
```{r}
variance_explained <-
  (
    pca_result$sdev^2 /
      sum(
        pca_result$sdev^2
      )
  ) * 100

variance_explained[1:10]
```



Create table:

```{r}
variance_table <- tibble(

  PC = paste0(
    "PC",
    seq_along(
      variance_explained
    )
  ),

  variance_percent =
    variance_explained,

  cumulative_percent =
    cumsum(
      variance_explained
    )
)

head(
  variance_table,
  10
)
```


24. PCA sample scores

```{r}
pca_scores <-
  as.data.frame(
    pca_result$x
  ) %>%
  rownames_to_column(
    "sample"
  ) %>%
  left_join(
    sample_metadata,
    by = "sample"
  )
```


  
25. Broad PC1 versus PC2 plot

```{r}
broad_pca_plot <- ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC2,
    fill = group,
    shape = broad_matrix
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(

    title =
      "PCA of positive-mode LC-MS features",

    subtitle =
      "5× blank-filtered, median-normalized and log2-transformed",

    x = paste0(
      "PC1 (",
      round(
        variance_explained[1],
        1
      ),
      "%)"
    ),

    y = paste0(
      "PC2 (",
      round(
        variance_explained[2],
        1
      ),
      "%)"
    ),

    fill =
      "Sample group",

    shape =
      "Sample matrix"
  ) +
  theme_classic()

broad_pca_plot
```


26. PC1 vs PC3
```{r}
pc1_pc3_plot <- ggplot(
  pca_scores,
  aes(
    x = PC1,
    y = PC3,
    fill = group,
    shape = broad_matrix
  )
) +
  geom_point(
    size = 4,
    color = "black"
  ) +
  geom_text_repel(
    aes(label = sample),
    size = 3,
    max.overlaps = Inf
  ) +
  labs(

    title =
      "PCA: PC1 versus PC3",

    x = paste0(
      "PC1 (",
      round(
        variance_explained[1],
        1
      ),
      "%)"
    ),

    y = paste0(
      "PC3 (",
      round(
        variance_explained[3],
        1
      ),
      "%)"
    )
  ) +
  theme_classic()

pc1_pc3_plot
```


27. Scree plot

```{r}
scree_plot <- variance_table %>%
  slice_head(
    n = 15
  ) %>%
  ggplot(
    aes(
      x = factor(
        PC,
        levels = PC
      ),
      y = variance_percent,
      group = 1
    )
  ) +
  geom_col() +
  geom_point() +
  geom_line() +
  labs(
    title =
      "Positive-mode PCA scree plot",
    x =
      "Principal component",
    y =
      "Variance explained (%)"
  ) +
  theme_classic()

scree_plot
```


28. Create reusable focused-PCA function

Instead of repeating the same code three times:
```{r}
run_focused_pca <- function(
    groups_to_use,
    title_text
) {

  samples_to_use <-
    sample_metadata %>%
    filter(
      group %in% groups_to_use
    ) %>%
    pull(sample)

  matrix_subset <-
    pca_matrix_log[
      samples_to_use,
      ,
      drop = FALSE
    ]

  variances <- apply(
    matrix_subset,
    2,
    var,
    na.rm = TRUE
  )

  matrix_subset <-
    matrix_subset[
      ,
      is.finite(variances) &
        variances > 0,
      drop = FALSE
    ]

  result <- prcomp(
    matrix_subset,
    center = TRUE,
    scale. = TRUE
  )

  variance <-
    (
      result$sdev^2 /
        sum(
          result$sdev^2
        )
    ) * 100

  scores <-
    as.data.frame(
      result$x
    ) %>%
    rownames_to_column(
      "sample"
    ) %>%
    left_join(
      sample_metadata,
      by = "sample"
    )

  plot <- ggplot(
    scores,
    aes(
      x = PC1,
      y = PC2,
      fill = group
    )
  ) +
    geom_point(
      shape = 21,
      size = 5,
      color = "black"
    ) +
    geom_text_repel(
      aes(label = sample)
    ) +
    labs(
      title = title_text,

      x = paste0(
        "PC1 (",
        round(
          variance[1],
          1
        ),
        "%)"
      ),

      y = paste0(
        "PC2 (",
        round(
          variance[2],
          1
        ),
        "%)"
      )
    ) +
    theme_classic()

  list(
    pca = result,
    scores = scores,
    variance = variance,
    plot = plot
  )
}
```


29. Exudate versus soil PCA
```{r}
exudate_soil_pca <-
  run_focused_pca(
    groups_to_use = c(
      "Plant-soil exudate",
      "Soil control"
    ),
    title_text =
      "PCA: plant-soil exudate versus soil control"
  )

exudate_soil_pca$plot
```


30. Fungal agar PCA


```{r}
fungal_pca <-
  run_focused_pca(
    groups_to_use = c(
      "Fungal control",
      "Fungal agar in",
      "Fungal agar out"
    ),
    title_text =
      "PCA of fungal agar samples"
  )

fungal_pca$plot
```


31. Leaf versus root PCA

```{r}
tissue_pca <-
  run_focused_pca(
    groups_to_use = c(
      "Leaf",
      "Root"
    ),
    title_text =
      "PCA of leaf and root extracts"
  )

tissue_pca$plot
```


32. Extract PCA loadings

```{r}
pca_loadings <-
  as.data.frame(
    pca_result$rotation
  ) %>%
  rownames_to_column(
    "feature_name"
  )
```



Create annotation table:

```{r}
feature_annotations <- feature_qc %>%
  select(
    feature_name,
    `Alignment ID`,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `Post curation result`,
    `Fill %`,
    `MS/MS assigned`,
    any_of(c(
      "Formula",
      "Ontology",
      "INCHIKEY",
      "SMILES",
      "Annotation tag (VS1.0)",
      "RT matched",
      "m/z matched",
      "MS/MS matched",
      "Total score",
      "RT similarity",
      "Dot product",
      "Reverse dot product",
      "Fragment presence %",
      "S/N average"
    )),
    blank_mean,
    sample_to_blank_ratio,
    number_detected
  )
```


Join:


```{r}
pca_loadings <- pca_loadings %>%
  left_join(
    feature_annotations,
    by = "feature_name"
  )
```


33. Identify strongest PC1, PC2 and PC3 features

Reusable function:

```{r}
get_top_loadings <- function(
    loading_table,
    pc_column,
    number_to_keep = 30
) {

  loading_table %>%
    mutate(
      selected_loading =
        .data[[pc_column]],

      absolute_loading =
        abs(
          .data[[pc_column]]
        )
    ) %>%
    arrange(
      desc(
        absolute_loading
      )
    ) %>%
    slice_head(
      n = number_to_keep
    )
}
```



Run:

```{r}

top_pc1_loadings <-
  get_top_loadings(
    pca_loadings,
    "PC1"
  )

top_pc2_loadings <-
  get_top_loadings(
    pca_loadings,
    "PC2"
  )

top_pc3_loadings <-
  get_top_loadings(
    pca_loadings,
    "PC3"
  )

```

Inspect:

```{r}
top_pc1_loadings %>%
  select(
    feature_name,
    PC1,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```



And:

```{r}
top_pc2_loadings %>%
  select(
    feature_name,
    PC2,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    `Total score`,
    sample_to_blank_ratio
  )
```


34. Check suspicious very-early RT loadings

This is important because we previously noticed many high-loading features around RT 0.4–0.8 min.

Don't automatically delete them yet.

Quantify them first:

```{r}
pca_loadings %>%
  mutate(
    important_PC12 =
      pmax(
        abs(PC1),
        abs(PC2)
      )
  ) %>%
  arrange(
    desc(
      important_PC12
    )
  ) %>%
  slice_head(
    n = 100
  ) %>%
  count(
    early_RT =
      `Average Rt(min)` < 1
  )

```


And inspect:

```{r}
pca_loadings %>%
  filter(
    `Average Rt(min)` < 1
  ) %>%
  arrange(
    desc(
      abs(PC1)
    )
  ) %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    PC1,
    PC2,
    sample_to_blank_ratio
  ) %>%
  head(30)
```



This should happen before interpreting PCA-driving metabolites.

35. Create combined PCA candidate table

```{r}
important_pca_features <-
  pca_loadings %>%
  mutate(

    maximum_absolute_loading =
      pmax(
        abs(PC1),
        abs(PC2),
        abs(PC3),
        na.rm = TRUE
      )

  ) %>%
  arrange(
    desc(
      maximum_absolute_loading
    )
  )

```


Inspect top 50:

```{r}
important_pca_features %>%
  slice_head(
    n = 50
  )
```


36. PCA feature heatmap

Since the annotated pheatmap continues to fail and the unannotated version works, leave the annotation problem alone for now. It is not blocking the analysis.

```{r}
top_feature_names <-
  important_pca_features %>%
  slice_head(
    n = 30
  ) %>%
  pull(
    feature_name
  )

heatmap_matrix <-
  t(
    pca_matrix_log[
      ,
      top_feature_names,
      drop = FALSE
    ]
  )

```


Plot:
```{r}
pheatmap(
  heatmap_matrix,
  scale = "row",
  clustering_distance_rows =
    "euclidean",
  clustering_distance_cols =
    "euclidean",
  clustering_method =
    "complete",
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 8,
  main =
    "Top features driving positive-mode PCA"
)

```


Your previous unannotated heatmap is the one that successfully rendered.


37. Presence/absence analysis across biological groups

Use feature_qc, not the imputed PCA matrix. Presence/absence must be based on actual observed peak heights.

First define all sample groups:

# ------------------------------------------------------------
# PRESENCE / ABSENCE ANALYSIS
# ------------------------------------------------------------

# Define sample columns for each biological group


```{r}
presence_groups <- list(

  Exudate = c(
    "exudate_1",
    "exudate_2",
    "exudate_3"
  ),

  Soil = c(
    "Soil_control_1",
    "Soil_control_2",
    "Soil_control_3"
  ),

  Leaf = c(
    "leaf_1",
    "leaf_2",
    "leaf_3"
  ),

  Root = c(
    "root_1",
    "root_2",
    "root_3"
  ),

  Fungal_in = c(
    "new_in_P08",
    "new_in_P09",
    "new_in_P10"
  ),

  Fungal_out = c(
    "new_out_P08",
    "new_out_P09",
    "new_out_P10"
  ),

  Fungal_control = c(
    "FC_1",
    "FC_2",
    "FC_3"
  ),

  Fungi_old = c(
    "fungi_old_1",
    "fungi_old_2",
    "fungi_old_3"
  )
)
```


38. Function for any pairwise presence/absence comparison

We'll define a feature as present when it occurs in at least 2 of 3 biological replicates.

```{r}
run_presence_comparison <- function(
    data,
    group1_cols,
    group2_cols,
    group1_name,
    group2_name,
    min_replicates = 2
) {

  result <- data %>%
    mutate(

      group1_detected_n = rowSums(
        across(
          all_of(group1_cols),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

      group2_detected_n = rowSums(
        across(
          all_of(group2_cols),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

      group1_present =
        group1_detected_n >= min_replicates,

      group2_present =
        group2_detected_n >= min_replicates,

      presence_category = case_when(

        group1_present &
          group2_present ~
          "Shared",

        group1_present &
          !group2_present ~
          paste0(
            group1_name,
            " only"
          ),

        !group1_present &
          group2_present ~
          paste0(
            group2_name,
            " only"
          ),

        TRUE ~
          "Neither consistently detected"
      )
    )

  summary <- result %>%
    count(
      presence_category
    ) %>%
    mutate(

      percent =
        100 * n / sum(n),

      comparison =
        paste0(
          group1_name,
          " vs ",
          group2_name
        )
    ) %>%
    select(
      comparison,
      presence_category,
      n,
      percent
    )

  return(
    list(
      feature_table = result,
      summary = summary
    )
  )
}
```


39. Exudate vs soil

This asks:

Which features appear with the plant-containing soil extraction compared with soil alone?

```{r}
presence_exudate_soil <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Exudate,

    group2_cols =
      presence_groups$Soil,

    group1_name =
      "Exudate",

    group2_name =
      "Soil"
  )

presence_exudate_soil$summary
```


40. Leaf vs root

This asks:

Which features are shared between aboveground and belowground plant tissue, and which are tissue-associated?

```{r}
presence_leaf_root <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Leaf,

    group2_cols =
      presence_groups$Root,

    group1_name =
      "Leaf",

    group2_name =
      "Root"
  )

presence_leaf_root$summary
```


41. Exudate vs root

Yes, I definitely want this comparison.

It asks:

Which features detected in roots are also detectable in the plant-soil extract?

```{r}
presence_exudate_root <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Exudate,

    group2_cols =
      presence_groups$Root,

    group1_name =
      "Exudate",

    group2_name =
      "Root"
  )

presence_exudate_root$summary
```



A feature classified as:

Shared

means it occurred in ≥2/3 exudates and ≥2/3 roots.

That is potentially interesting.

But:

shared ≠ secreted.

It only establishes chemical correspondence at the aligned-feature level.

42. Exudate vs leaf

This comparison is also worthwhile:

```{r}
presence_exudate_leaf <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Exudate,

    group2_cols =
      presence_groups$Leaf,

    group1_name =
      "Exudate",

    group2_name =
      "Leaf"
  )

presence_exudate_leaf$summary
```



This can help distinguish whether exudate-associated signals resemble general plant chemistry or are more strongly associated with roots.

43. Fungal agar in vs out

```{r}
presence_fungal_in_out <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Fungal_in,

    group2_cols =
      presence_groups$Fungal_out,

    group1_name =
      "Fungal in",

    group2_name =
      "Fungal out"
  )

presence_fungal_in_out$summary
```


44. Fungal agar in vs agar control

```{r}
presence_fungal_in_control <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Fungal_in,

    group2_cols =
      presence_groups$Fungal_control,

    group1_name =
      "Fungal in",

    group2_name =
      "Fungal control"
  )

presence_fungal_in_control$summary
```


45. Fungal agar out vs agar control

I would add this too.

It tells us whether the region outside visible fungal growth has nevertheless been chemically altered relative to the control.

```{r}
presence_fungal_out_control <-
  run_presence_comparison(

    data = feature_qc,

    group1_cols =
      presence_groups$Fungal_out,

    group2_cols =
      presence_groups$Fungal_control,

    group1_name =
      "Fungal out",

    group2_name =
      "Fungal control"
  )

presence_fungal_out_control$summary
```



That could actually be biologically interesting if diffusible metabolites extend beyond the visible colony.

46. Combine all pairwise presence results

```{r}
presence_summary_all <- bind_rows(

  presence_exudate_soil$summary,

  presence_leaf_root$summary,

  presence_exudate_root$summary,

  presence_exudate_leaf$summary,

  presence_fungal_in_out$summary,

  presence_fungal_in_control$summary,

  presence_fungal_out_control$summary
)

presence_summary_all

```


Sort it:

```{r}
presence_summary_all <- presence_summary_all %>%
  arrange(
    comparison,
    desc(n)
  )

presence_summary_all
```



Save:

```{r}

write.csv(
  presence_summary_all,
  "positive_mode_presence_absence_summary.csv",
  row.names = FALSE
)
```

47. Now do the more interesting three-way plant analysis

Pairwise comparisons are useful, but what you're proposing is actually better expressed as:

Which features occur across leaf → root → exudate?

Let's calculate presence in all three.

```{r}
plant_compartment_presence <- feature_qc %>%
  mutate(

    leaf_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Leaf
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    root_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Root
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    exudate_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Exudate
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    soil_detected_n =
      rowSums(
        across(
          all_of(
            presence_groups$Soil
          ),
          ~ .x > 0
        ),
        na.rm = TRUE
      ),

    leaf_present =
      leaf_detected_n >= 2,

    root_present =
      root_detected_n >= 2,

    exudate_present =
      exudate_detected_n >= 2,

    soil_present =
      soil_detected_n >= 2
  )
```


48. Classify plant-compartment patterns
```{r}
plant_compartment_presence <-
  plant_compartment_presence %>%
  mutate(

    plant_pattern =
      case_when(

        leaf_present &
          root_present &
          exudate_present ~
          "Leaf + Root + Exudate",

        !leaf_present &
          root_present &
          exudate_present ~
          "Root + Exudate",

        leaf_present &
          !root_present &
          exudate_present ~
          "Leaf + Exudate",

        leaf_present &
          root_present &
          !exudate_present ~
          "Leaf + Root",

        leaf_present &
          !root_present &
          !exudate_present ~
          "Leaf only",

        !leaf_present &
          root_present &
          !exudate_present ~
          "Root only",

        !leaf_present &
          !root_present &
          exudate_present ~
          "Exudate only",

        TRUE ~
          "None consistently detected"
      )
  )
```



Summarize:

```{r}
plant_pattern_summary <-
  plant_compartment_presence %>%
  count(
    plant_pattern,
    sort = TRUE
  ) %>%
  mutate(
    percent =
      100 * n / sum(n)
  )

plant_pattern_summary
```



This is probably one of the most useful exploratory tables for your plant samples.

49. Identify possible plant-associated exudate features

Now let's get closer to your biological idea.

We want features that:

are present in exudate;
are present in root and/or leaf;
are not consistently detected in soil control.

That helps eliminate features that are simply normal soil-associated signals.

Root + exudate, absent from soil

```{r}
root_exudate_not_soil <-
  plant_compartment_presence %>%
  filter(
    root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  root_exudate_not_soil
)
```



Inspect:

```{r}
root_exudate_not_soil %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio,
    leaf_present,
    root_present,
    exudate_present,
    soil_present
  ) %>%
  head(50)
```



This is a very interesting candidate list.

50. Leaf + root + exudate, absent from soil

These may represent broadly plant-associated compounds detectable from the soil/exudate extraction.

```{r}
leaf_root_exudate_not_soil <-
  plant_compartment_presence %>%
  filter(
    leaf_present,
    root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  leaf_root_exudate_not_soil
)

```


View:

```{r}
leaf_root_exudate_not_soil %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio
  ) %>%
  head(50)
```


51. Root + exudate but NOT leaf and NOT soil

This may be even more interesting for root-associated chemistry.

```{r}
root_exudate_specific_candidates <-
  plant_compartment_presence %>%
  filter(
    !leaf_present,
    root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  root_exudate_specific_candidates
)
```



Inspect:

```{r}
root_exudate_specific_candidates %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio
  ) %>%
  head(50)
```



This category would get my attention.

Conceptually:

Leaf       absent
Root       present
Exudate    present
Soil       absent

That pattern is consistent with a root-associated compound appearing in the rhizosphere/exudate extraction.

Still not proof of secretion—but much stronger biological prioritization than just “exudate peak high.”

52. Exudate-only candidates absent from soil

You should also keep these.

A compound doesn't have to accumulate strongly inside root tissue to be released outside the plant.

```{r}
exudate_only_not_soil <-
  plant_compartment_presence %>%
  filter(
    !leaf_present,
    !root_present,
    exudate_present,
    !soil_present
  ) %>%
  arrange(
    desc(
      sample_to_blank_ratio
    )
  )

nrow(
  exudate_only_not_soil
)
```



Those could potentially include:

low-abundance secreted compounds;
compounds rapidly exported from roots;
root-derived transformation products;
microbially transformed plant compounds;
compounds produced specifically in the rhizosphere.

So do not discard them simply because they aren't detected in root tissue.

53. Save the plant-compartment candidate tables

```{r}
write.csv(
  plant_compartment_presence,
  "positive_mode_plant_compartment_presence.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  plant_pattern_summary,
  "positive_mode_plant_pattern_summary.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  root_exudate_not_soil,
  "positive_mode_root_exudate_not_soil.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  leaf_root_exudate_not_soil,
  "positive_mode_leaf_root_exudate_not_soil.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  root_exudate_specific_candidates,
  "positive_mode_root_exudate_specific_candidates.csv",
  row.names = FALSE
)
```


```{r}
write.csv(
  exudate_only_not_soil,
  "positive_mode_exudate_only_not_soil.csv",
  row.names = FALSE
)
```



54. Differential abundance: exudate vs soil

First run the comparison using the pairwise function we created earlier:

```{r}
exudate_vs_soil <-
  run_pairwise_comparison(

    data_matrix =
      pca_matrix_log,

    metadata =
      sample_metadata,

    annotation_table =
      feature_annotations,

    group_1 =
      "Plant-soil exudate",

    group_2 =
      "Soil control"
  )
```



Check that it worked:

```{r}
dim(exudate_vs_soil)

head(
  exudate_vs_soil
)
```


55. Summarize the statistical results

Let's first see how many features show evidence of a difference.

```{r}
exudate_statistics_summary <-
  exudate_vs_soil %>%
  summarise(

    total_features =
      n(),

    raw_p_below_0.05 =
      sum(
        p_value < 0.05,
        na.rm = TRUE
      ),

    FDR_below_0.05 =
      sum(
        adjusted_p_value < 0.05,
        na.rm = TRUE
      ),

    absolute_log2FC_above_1 =
      sum(
        abs(log2_fold_change) >= 1,
        na.rm = TRUE
      ),

    raw_p_and_log2FC =
      sum(
        p_value < 0.05 &
          abs(log2_fold_change) >= 1,
        na.rm = TRUE
      ),

    FDR_and_log2FC =
      sum(
        adjusted_p_value < 0.05 &
          abs(log2_fold_change) >= 1,
        na.rm = TRUE
      )
  )

exudate_statistics_summary

```


Remember:

log2FC = +1    = 2× higher in exudate
log2FC = +2    = 4× higher
log2FC = +3    = 8× higher

log2FC = -1    = 2× higher in soil
log2FC = -2    = 4× higher in soil

Because you only have n = 3 per group, don't be surprised if relatively few features survive FDR correction. We shouldn't equate FDR > 0.05 with “biologically uninteresting” in this pilot.

56. Classify exudate-vs-soil features

Now make the results easier to interpret.

```{r}
exudate_vs_soil <-
  exudate_vs_soil %>%
  mutate(

    statistical_category =
      case_when(

        adjusted_p_value < 0.05 &
          log2_fold_change >= 1 ~
          "Exudate enriched - FDR significant",

        adjusted_p_value < 0.05 &
          log2_fold_change <= -1 ~
          "Soil enriched - FDR significant",

        p_value < 0.05 &
          log2_fold_change >= 1 ~
          "Exudate enriched - exploratory",

        p_value < 0.05 &
          log2_fold_change <= -1 ~
          "Soil enriched - exploratory",

        TRUE ~
          "Not selected"
      )
  )

```


Count:

```{r}
exudate_vs_soil %>%
  count(
    statistical_category,
    sort = TRUE
  )
```


57. Make the volcano plot

```{r}
exudate_volcano_data <-
  exudate_vs_soil %>%
  mutate(

    minus_log10_p =
      -log10(
        pmax(
          p_value,
          1e-300
        )
      )
  )
```


Plot:

```{r}
ggplot(
  exudate_volcano_data,
  aes(
    x = log2_fold_change,
    y = minus_log10_p,
    fill = statistical_category
  )
) +

  geom_point(
    shape = 21,
    size = 2.5,
    alpha = 0.75
  ) +

  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed"
  ) +

  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +

  labs(

    title =
      "Plant-soil exudate vs soil control",

    subtitle =
      "Positive-mode LC-MS",

    x =
      "Log2 fold change (exudate / soil)",

    y =
      "-log10 raw p-value",

    fill =
      "Classification"
  ) +

  theme_classic() +

  theme(
    legend.position = "bottom"
  )

```


This gives you the broad quantitative picture.

58. Now connect the statistics to the biology

This is the part I think is particularly important for your experiment.

You already created:

root_exudate_specific_candidates

Those are:

Leaf       absent
Root       present
Exudate    present
Soil       absent

Now join those candidates to the exudate-vs-soil statistics.

root_exudate_candidate_stats <-
  root_exudate_specific_candidates %>%

  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    Formula,
    Ontology,
    `Total score`,
    sample_to_blank_ratio
  ) %>%

  left_join(

    exudate_vs_soil %>%
      select(
        feature_name,
        group_1_mean,
        group_2_mean,
        log2_fold_change,
        fold_change,
        p_value,
        adjusted_p_value,
        statistical_category
      ),

    by = "feature_name"
  )

Then rank them:

root_exudate_candidate_stats <-
  root_exudate_candidate_stats %>%
  arrange(
    p_value,
    desc(log2_fold_change)
  )

root_exudate_candidate_stats

Now your table is doing something much more powerful.

Instead of merely:

F1207
Root = yes
Exudate = yes
Soil = no

you'll have:

Feature
m/z
RT
MS/MS?
blank ratio
exudate mean
soil mean
log2 fold change
fold change
p-value
FDR
59. Add a candidate-priority category

For a pilot study, I don't want you throwing away interesting candidates merely because FDR doesn't reach 0.05 with three replicates.

Instead, classify the evidence.

root_exudate_candidate_stats <-
  root_exudate_candidate_stats %>%
  mutate(

    candidate_priority =
      case_when(

        adjusted_p_value < 0.05 &
          log2_fold_change >= 1 ~
          "Priority 1 - FDR supported",

        p_value < 0.05 &
          log2_fold_change >= 1 ~
          "Priority 2 - exploratory statistical support",

        log2_fold_change >= 1 ~
          "Priority 3 - strong fold change",

        TRUE ~
          "Priority 4 - presence-pattern candidate"
      )
  )

Count them:

root_exudate_candidate_stats %>%
  count(
    candidate_priority
  )
60. Flag MS/MS and suspicious early retention times

Now add analytical confidence.

root_exudate_candidate_stats <-
  root_exudate_candidate_stats %>%
  mutate(

    RT_flag =
      case_when(

        `Average Rt(min)` < 0.5 ~
          "Very early RT - inspect carefully",

        `Average Rt(min)` < 1.0 ~
          "Early RT - inspect",

        TRUE ~
          "RT >= 1 min"
      ),

    MSMS_flag =
      case_when(

        `MS/MS assigned` == TRUE ~
          "MS/MS available",

        TRUE ~
          "No MS/MS assignment"
      )
  )

Now inspect the most promising ones:

root_exudate_candidate_stats %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `Metabolite name`,
    `Adduct type`,
    `MS/MS assigned`,
    RT_flag,
    sample_to_blank_ratio,
    group_1_mean,
    group_2_mean,
    log2_fold_change,
    fold_change,
    p_value,
    adjusted_p_value,
    candidate_priority
  ) %>%
  arrange(
    candidate_priority,
    p_value
  )
61. Save this important candidate table
write.csv(
  root_exudate_candidate_stats,
  "positive_mode_root_exudate_candidate_statistics.csv",
  row.names = FALSE
)

This is one of the first output files from this analysis that I would consider a biologically interpretable candidate table rather than simply a QC/statistical intermediate.

Then we should do something slightly different

After this, I wouldn't immediately repeat all of this for every comparison.

First, show me the output of:

exudate_statistics_summary

and:

root_exudate_candidate_stats %>%
  select(
    feature_name,
    `Average Rt(min)`,
    `Average Mz`,
    `MS/MS assigned`,
    sample_to_blank_ratio,
    log2_fold_change,
    fold_change,
    p_value,
    adjusted_p_value,
    candidate_priority
  )
